### 2018-07-20 PPforestVizupdate.Rnw
complete viz version

\documentclass[12pt]{article}
\usepackage{multirow}
\usepackage{mathtools}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{rotating}
\usepackage{booktabs}
\usepackage{verbatim}
\usepackage{mathrsfs}
\usepackage{hyperref}
\usepackage{pdflscape}
\usepackage{caption}
\usepackage{tablefootnote}
\usepackage{url}
\usepackage[authoryear]{natbib}
\topmargin     -1.5cm  % read Lamport p.163
\oddsidemargin -0.04cm % read Lamport p.163
\evensidemargin -0.04cm % same as oddsidemargin but for left-hand pages
\textwidth     16.59cm
\textheight     22.94cm
\parskip       7.2pt  % sets spacing between paragraphs
\parindent       3mm  % sets leading space for paragraphs
\usepackage{verbatim}
\usepackage{setspace}
\usepackage[utf8]{inputenc}
\usepackage{authblk}
\usepackage{pdfpages}
\usepackage{fancyhdr}
\fancyhf{}
\captionsetup{font={stretch=1}}
\renewcommand{\headrulewidth}{0pt}
\newcommand{\pkg}[1]{{\fontseries{b}\selectfont #1}}
\pagestyle{fancy}
\rfoot{\thepage}
\linespread{1.5}
\title{Interactive Graphics for Visually Diagnosing Forest Classifiers in R}
\author{Natalia da Silva, Dianne Cook \& Eun-Kyung Lee}

\begin{document}
\maketitle

\section*{Abstract}
This paper describes structuring data and constructing plots to explore forest classification models interactively. A forest classifier is an example of an ensemble since it is produced by bagging multiple trees. The process of bagging and combining results from multiple trees produces numerous diagnostics which, with interactive graphics, can provide a lot of insight into class structure in high dimensions. Various aspects are explored in this paper, to assess model complexity, individual model contributions, variable importance and dimension reduction, and uncertainty in prediction associated with individual observations. The ideas are applied to the random forest algorithm and projection pursuit forest, but could be more broadly applied to other bagged ensembles. Interactive graphics are built in \pkg{R} using the \pkg{ggplot2}, \pkg{plotly}, and \pkg{shiny} packages.

\section{Introduction}
The random forest (RF) algorithm~\citep{breiman1996bagging} was one of the first ensemble classifiers developed. It combines the predictions from individual classification and regression trees (CART)~\citep{breiman1984cl} built by bagging observations~\citep{breiman1996bagging}. It also samples variables at each tree node. These produce diagnostics in the form of uncertainty in predictions for each observation, importance of variables for the prediction, predictive error for future samples based on out-of-bag (OOB) case predictions, and similarity of observations based on how often they group together in the trees.

Ensemble classifiers have grown in popularity~\citep{dietterish00}~\citep{talbot09}, and the basic ideas behind the random forest can be applied to virtually any type of model. The benefits for classification are reduced variability in predictive error, and the suite of diagnostics provides the potential for better understanding the class structure in the high-dimensional data space. The use of visualization on these diagnostics, in association with multivariate data plots, completes the process to support a better understanding of the underlying problem.

A conceptual framework for model visualization can be summarized in three strategies: (1) visualize the model in the data space, (2) look all members of a collection of a model and (3) explore the complete process of model fitting~\citep{wickham2015visualizing}. The first strategy is to explore how well the model captures the data characteristics (model in the data space), which contrasts with determining if the model assumptions hold (data in the model space). The second strategy is to look at a group of models instead of only the best model. This strategy can offer a broad understanding of the problem by comparing and contrasting possible models. The last strategy focuses on exploring the process of the model fitting in addition to the end result.

There has been some, but not a lot of, research on visualizing classification models.
\cite{urbanek2002exploring} presents interactive tree visualization implemented in the java software called \pkg{KLIMT} that include zooming, selection, multiple views, interactive pruning, and tree construction as well as the interactive analysis of forests of trees using treemaps. \cite{cutler15raft} developed a java package called \pkg{RAFT} to visualize a forest classifier, that included variable selection, parallel coordinate plots, heat maps, and scatter plots of some diagnostics. Linking between plots is limited. \cite{quach2012interactive} presents interactive forest visualization using the R package \pkg{iPlots eXtreme} \citep{urbanek2011iplots} where several displays are shown in one window with some linking between them available. \cite{Silva2016} describes visualizing components of an ensemble classifier.

This paper describes structuring interactive graphics to facilitate visual exploration of ensemble classifiers using RFs and projection pursuit forests (PPF)~\citep{dasilvappforest} as examples. The PPF algorithm builds on the projection pursuit tree (PPtree)~\citep{lee2013pptree} algorithm which uses projection pursuit at each tree node to find the best linear combination of variables to separate the classes. The visualization approach is consistent with the framework in \cite{wickham2015visualizing}, and the implementation is built on the newest interactive graphics available in \pkg{R}\citep{RCore}. The purpose is to provide readily available tools for users to explore and improve ensemble fits and obtain an intuition for the underlying class structure in data. Interactive plots are a key component for model visualization that help the user see multivariate relationships and be more efficient in the model diagnosis. Multiple levels of data are constructed for exploration: observation, model and ensemble summaries.

The paper is structured as follows.  Section \ref{key} describes the ensemble components to be accessed.  Section \ref{visen} maps the ensemble components to the visual elements. The web app is described in \ref{app} and further work is discussed in section \ref{fur}.

\section{Diagnostics in forest classifiers}\label{key}

The diagnostics typically available are:

\begin{itemize} \itemsep 0in
  \item {\em Out-of-bag error:} For each model in the ensemble, some cases of the original data are not used. Predicting the response for these cases gives a better estimate for the error of the model with future data. The OOB error rate is a measure for each model that is combined in the ensemble and is used to provide the overall error of the ensemble.
  \item {\em Uncertainty measure for each observation:} Across individual (classification) models we can compute the proportion of times that a case is predicted to be each class. If a case is always predicted to be the true class, there is no uncertainty about an observation. Cases that are proportionately predicted to be multiple classes indicate difficult to classify observations. They may be important by indicating neighborhoods of the data space that would benefit from a more complex model, or more simply, they may be errors in measurements in the data.
  \item {\em Variable importance:} Each model uses samples of variables. So that, the accuracy of the models can be compared when the variable is included or omitted. There are several versions of statistics that use this to provide a measure of the variable importance for prediction.
  \item {\em Similarity measure for pairs of observations:} In each model, each pair of observations will be either in the same terminal node or not. This is used to compute a proximity matrix.  Cluster analysis on this matrix  can be used to follow up the classification to assess the original labeling. It may suggest improvements or mistakes in original labels.
\end{itemize}

In addition to these overall ensemble statistics, each component model has its own diagnostics, measuring error, variables utilized, and class predictions. Visualization will enable the individual models to be examined, relate these to the data and their contribution to the ensemble.

\section{Mapping ensemble diagnostics to visual components}~\label{visen}

This section describes the mapping of diagnostics to visualizations. These are illustrated using the Australian crabs data~\citep{CM74}. The data has 200 cases, 5 predictors and 4 classes (combinations of species and sex, blue male, blue female, orange male and orange female). The predictors are: FL (the size of the frontal lobe length, in mm), RW (rear width, in mm), CL (length of mid-line of the carapace, in mm), CW (maximum width of carapace, in mm), BD (depth of the body; for females, measured after displacement of the abdomen, in mm). This is old data, but it provides a good illustration of the visual methods.

\subsection{Individual models: PPtree}

The PPF is composed of individual projection pursuit trees. Figure \ref{pptree} shows a visual ensemble of plots of a tree model on the crab data. There are three nodes for the four class problem. The nodes of this tree are based on projections of the data; the coefficients of which form the building block to calculate the variable importance. The density plot displays the data projection at each node, and the mosaic plot shows the confusion matrix for the nodes.
The package \pkg{PPtreeViz} provides visual tools to diagnose a PPtree model. The PPF builds on these and modifies a little. The PPtree model is simpler than a regular classification tree because the classes are mostly separated by combinations of variables -- just three projections are needed to see the differences between the four classes.

<<pkg, echo=FALSE, message=FALSE, warning=FALSE>>=
# require(devtools)
# install_version("ggplot2", version = "0.9.1", repos = "http://cran.us.r-project.org")
library(ggplot2)
library(PPforest)
library(shiny)
#library(plotly)
library(GGally)
library(productplots)
library(PPtreeViz)
library(tidyr)
library(dplyr)
library(pracma)
library(RColorBrewer)
library(randomForest)
library(pROC)
library(gridExtra)
library(grid)
library(ggmosaic)
#library(PairViz)
 load("ppf.RData")
 load("rf_fish.RData")
@

\begin{figure}[hbpt]
\centering
<<trcomb, echo=FALSE, warning=FALSE, message=FALSE, fig.height=4, fig.width=6.5, results='hide', error=FALSE>>=
tree <- 18

#Data trans density and mosaic
dat_dens <- function(PPclassOBJ, node.id, Rule, legend = TRUE, std = TRUE,
                     image = FALSE, diff.prop = 0.2,c1=FALSE) {

  searchGroup <- function(node.id, TS, gName) {
    flag <- TRUE
    sel.id <- TS[node.id, 2:3]
    LR.id <- c(TRUE, FALSE)
    sel.group <- NULL
    i <- 1
    while ((sel.id[i] != 0) && (i < length(sel.id))) {
      if (TS[sel.id[i], 2] != 0) {
        sel.id <- c(sel.id, TS[sel.id[i], 2:3])
        if (LR.id[i])
          LR.id <- c(LR.id, c(TRUE, TRUE))
        else LR.id <- c(LR.id, c(FALSE, FALSE))
      }
      if (TS[sel.id[i + 1], 2] != 0) {
        sel.id <- c(sel.id, TS[sel.id[i + 1], 2:3])
        if (LR.id[i + 1])
          LR.id <- c(LR.id, c(TRUE, TRUE))
        else LR.id <- c(LR.id, c(FALSE, FALSE))
      }
      i <- i + 2
    }
    sel.Name <- TS[sel.id[which(TS[sel.id, 2] == 0)], 3]
    selName <- sort(gName[sel.Name])
    L.list <- sort(gName[sel.Name[LR.id[which(TS[sel.id,
                                                 2] == 0)]]])
    R.list <- sort(gName[sel.Name[!LR.id[which(TS[sel.id,
                                                  2] == 0)]]])
    return(list(selName = selName, Llist = L.list, Rlist = R.list))
  }

  TS <- PPclassOBJ$Tree.Struct
  Alpha <- PPclassOBJ$projbest.node
  cut.off <- PPclassOBJ$splitCutoff.node
  origdata <- PPclassOBJ$origdata
  origclass <- PPclassOBJ$origclass
  p <- ncol(origdata)
  gName <- names(table(origclass))

  if (TS[node.id, 2] != 0) {
    SG.result <- searchGroup(node.id, TS, gName)
    selG <- SG.result$selName
    selL <- SG.result$Llist
    selR <- SG.result$Rlist
    sel.id <- NULL
    LR.class <- NULL
    for (i in 1:length(selG)) {
      sel.id <- c(sel.id, which(origclass == selG[i]))
      LR.class <- c(LR.class, rep(ifelse(sum(selL == selG[i]) !=
                                           0, "L", "R"), length(which(origclass == selG[i]))))
    }
    proj.data <- c(as.matrix(origdata) %*% as.matrix(Alpha[TS[node.id,
                                                              4], ]))[sel.id]

    proj.class <- origclass[sel.id]
    plot.data <- data.frame(proj.data = proj.data, origclass = proj.class,
                            cut = cut.off[TS[node.id, 4], Rule], node.id =node.id, LR.class,Dir=as.factor(proj.data>cut.off[TS[node.id, 4], Rule]))
    colnames(plot.data)[2]<-"Class"

    plot.data

  }
}

PPtree_dens <- function(ppf, tr, nodes = NULL, pos ="none") {

  if(length(ppf[[8]][[tr]]$Tree.Struct[ppf[[8]][[tr]]$Tree.Struct[,4]!=0,1])<3){
    ns =length(ppf[[8]][[tr]]$Tree.Struct[ppf[[8]][[tr]]$Tree.Struct[,4]!=0,1])
  }else{
    ns=3
  }
  if(is.null(nodes)){
    nodes <- ppf[[8]][[tr]]$Tree.Struct[ppf[[8]][[tr]]$Tree.Struct[,4]!=0,1][1:ns]
  }
  nn <- data.frame(nn = nodes)
  densf <- function(x) {
    dat_dens(PPclassOBJ = ppf[["output.trees"]][[tr]],
             node.id = x,
             Rule = 1)

  }

dat_pl <- lapply(as.list(nodes), densf) %>% bind_rows(.id='node')

  myColors <- brewer.pal(dim(unique(ppf$train[ppf$class.var]))[1], "Dark2")
  names(myColors) <- levels(ppf$train[ppf$class.var][, 1])
  dat_pl$Class <- as.factor(dat_pl$Class)

  if(is.factor(ppf$train[ppf$class.var][, 1])){
    levels(dat_pl$Class) <-  levels(ppf$train[ppf$class.var][, 1])

  }else{
    levels(dat_pl$Class) <-  levels(as.factor(ppf$train[ppf$class.var][, 1]))
  }

  p1 <- dat_pl %>% filter(node.id %in%nodes) %>%
    ggplot( aes(  x = proj.data, group = Class, fill = Class ) ) +
    geom_density(alpha = .5) + facet_grid(~ node.id, scales = 'free') +
    scale_fill_manual(values = myColors, labels=c("BF", "BM", "OF", "OM"), name="") +geom_vline(aes(xintercept = cut),
                                                      linetype = "dashed",
                                                      color = 2) + labs(x="")

  p1 + theme(legend.position = pos,aspect.ratio = 1) +
    guides(col=guide_legend(nrow = 2))
}

p1b <- PPtree_dens(ppf, tree, nodes =c(1,2,3), pos="none")

#Mosaic plot Tab 2
PPtree_mosaic <- function(ppf,tr, nodes = NULL, pos="none"){
  if(length(ppf[[8]][[tr]]$Tree.Struct[ppf[[8]][[tr]]$Tree.Struct[,4]!=0,1])<3){
    ns =length(ppf[[8]][[tr]]$Tree.Struct[ppf[[8]][[tr]]$Tree.Struct[,4]!=0,1])
  }else{
    ns=3
  }

  if(is.null(nodes)){
    nodes <- ppf[[8]][[tr]]$Tree.Struct[ppf[[8]][[tr]]$Tree.Struct[,4]!=0,1][1:ns]
  }
  nn <- data.frame(nn = nodes)

  densf <- function(x){
    dat_dens(PPclassOBJ=ppf[[8]][[tr]],node.id=x,Rule=1)
  }

dat_pl <- lapply(as.list(nodes), densf) %>% bind_rows(.id='node')
  levels(dat_pl$Dir)<-c("Left", "Right")
  myColors <- brewer.pal(length( unique( ppf[[8]][[tr]]$origclass ) ), "Dark2")
  names(myColors) <- levels(dat_pl$Class)

  dat_mosaic <- data.frame( with(dat_pl, table(Class, Dir,node.id) ) )

  p1 <- dat_mosaic %>% filter(node.id %in%nodes) %>% ggplot() +
    geom_mosaic( aes(weight = Freq, x = product(Class,Dir) ,fill = Class),  offset=.005)+facet_grid(~node.id)+
    scale_fill_manual(values = myColors) + theme(legend.position=pos,axis.text.x  = element_text(angle=90, vjust=0.5),aspect.ratio = 1) +xlab("")

 p1

}

p1c <- PPtree_mosaic(ppf, tree, nodes = c(1,2,3))

nodes <- nrow(ppf[["output.trees"]][[tree]]$projbest.node)
projbest.node <- ppf[["output.trees"]][[tree]]$projbest.node
coefs <- data.frame(node = as.factor(1:nodes) , projbest.node)
colnames(coefs)  <- c("node", colnames(ppf$train[, -1]))
coefs <- coefs %>% gather(variable, value,-node)

p1d <-
  coefs %>%
  ggplot(aes(x = variable, y = value)) + geom_segment(aes(yend = 0, xend = variable), size = 3) +
  facet_grid(node ~ .) + geom_hline(yintercept = 0) + geom_hline(
    yintercept = c(-1, 1) * 1 / sqrt(ncol(ppf$train[, -1])), col = 2,
    linetype = "dashed") + labs(y = "Coefficients")

g_legend <- function(a.gplot){
  tmp <- ggplot_gtable(ggplot_build(a.gplot))
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
  legend <- tmp$grobs[[leg]]
  return(legend)
}

layoutmat <- cbind(c(1, 1, 1), c(2, 3, 4))

labpl <-PPtree_dens(ppf, tree, nodes =c(1,2,3), pos="bottom")

plot(ppf[["output.trees"]][[tree]], main = "", font.size = 12)
         grid.arrange(p1d , p1b, p1c,  g_legend(labpl),layout_matrix=layoutmat  ,
       widths =c(2,2) ,heights=c(2,2,.5), as.table =TRUE)

  @
\caption{Visualizing the PPtree model of the crab data. The tree has three nodes (top). The density plots show the data projections at each node, colored by group (middle). The dashed vertical red line indicates the split value of each node. At node 1 the blue species is separated from orange species. Nodes 2 and 3 separate the sexes, which are more confused for the blue species. Mosaic plots of the confusion table for each split (bottom). Node 1 shows the clear split of the species, with a small number of misclassifications. Node 2 where orange females are separated from orange males indicates small number of misclassifications. Node 3 where blue females are separated from blue males, shows a larger misclassification than for orange specie.
    \label{pptreeviz1}}
\end{figure}

\subsection{Variable importance}

The PPF algorithm calculates variable importance in two ways: (1) permuted importance using accuracy,  and (2) importance based on projection coefficients of standardized variables.

The permuted variable importance is comparable with the measure defined in the classical random forest algorithm. It is computed using the OOB sample for the tree $k\;\;(B^{(k)})$ for each $X_j$ predictor variable.  Then the
permuted importance of the variable $X_j$ in the tree $k$ can be defined as:

\[
IMP^{(k)}(X_j) = \frac{\sum_{i \in B^{(k)} } I(y_i=\hat y_i^{(k)})-I(y_i=\hat y_{i,P_j}^{(k)})}{|B^{(k)}|}
\]

\noindent where $\hat y_i^{(k)}$
is the predicted class for observation $i$ in tree $k$ and $y_{i,P_j}^{(k)}$ is the predicted class for observation $i$ in tree $k$ after permuting the values for variable $X_j$. The global permuted importance measure is the average importance over all the trees in the forest.
This measure is based on comparing the accuracy of classifying OOB observations using the true class with permuted (nonsense) class.

For the second importance measure, the coefficients of each projection are examined. The magnitude of these values indicates importance, if the variables have been standardized. The variable importance for a single tree is computed by a weighted sum of the absolute values of the coefficients across nodes. The weights takes the number of classes in each node into account ($cl_{nd}$)~\citep{lee2013pptree}.
Then the importance of the variable $X_j$ in the PPtree $k$ can be defined as:

\[
  IMP_{pptree}^{(k)}(X_j)=\sum_{nd = 1}^{nn}\frac{|\alpha_{nd}^{(k)}|}{cl_{nd} }
\]

Where $\alpha_{nd}^{(k)}$ is the projected coefficient for node $nd$, variable $k$, and $nn$ the total number of node partitions in the tree $k$.

The global variable importance in a PPforest then can be defined in different ways. The most intuitive is the average variable importance from each PPtree across all the trees in the forest.

\[
IMP_{ppforest1}(X_j)=\frac{\sum_{k=1}^K IMP_{pptree}^{(k)}(X_j)}{K}
\]
Alternatively we have defined a global importance measure for the forest as a weighted mean of the absolute value of the projection coefficients across all nodes in every tree. The weights are based on the projection pursuit indexes in each node ($Ix_{nd}$), and 1-(OOB-error of each tree)($acc_k$).

\[IMP_{ppforest2}(X_j)=\frac{\sum_{k=1}^K acc_k \sum_{nd = 1}^{nn}\frac{Ix_{nd}|\alpha_{nd}^{(k)}|}{nn }}{K}
\]

Figure \ref{impotree} shows the absolute projection coefficient of the top three nodes for all the trees in a forest model. This information is displayed by a side-by-side jittered dot plot. The red dots correspond to the absolute coefficient values for the tree model of Figure \ref{pptree}. The forest was built using random samples of two variables for each node, hence there are two coefficients for each node. At node 1, BD has a high value and CW contributes much less. The scatterplot at right shows these two variables and the resulting boundary between groups that this would produce. Node 2 uses CL and RW, and RW contributes the most to the separation. The plot at right shows the boundary that is induced. Node 3 uses FL and RW, and this is a much more even contribution by the two variables.
For each tree in the forest, different decision rules are defined; the resulting boundaries on the previous plots are based on Rule 1 $= \frac{m_1}{2} + \frac{m_2}{2} $, where $m_1$ and $m_2$ are the mean of the left and right groups at each node.

\begin{figure}[hbpt]
\centering
<<treenode, echo=FALSE, warning=FALSE, message=FALSE,fig.height=7,fig.width=7.5>>=

# Jittered side by side plot with the absolute importance variable for a each tree in each node partition. The red dot in each
#facet identified a specific tree in the forest. In crab data we have two points for each node partition because in each node
# we have selected to variables at random.

impotree <- function(data,ppf, nodes = c(1, 2, 3) ){

bnf <- function(x) {
  bn <- abs(x$projbest.node)
  bn[bn == 0] <- NA
  data.frame(node = 1:dim(x$projbest.node)[1], bn)
}

bestnode <- ppf[["output.trees"]] %>%  lapply(bnf) %>% bind_rows()

colnames(bestnode)[-1] <- colnames(data[, -1])
bestnode$node <- as.factor(bestnode$node)

aux <- bestnode %>% filter(node %in% nodes) %>% mutate(ids = rep(1:ppf$n.tree, each = length(nodes))) %>% gather(var, value, -ids, -node)

aux$Variables <- as.numeric(as.factor(aux$var) )
aux$Abs.importance <- round( aux$value, 2)

p <- ggplot(filter(aux, !ids %in% tree),
            aes(x = Variables , y = Abs.importance , group = ids)) + geom_jitter(height = 0, size = I(3),
                                                                                 alpha = 0.3) +
  facet_grid(node ~ .) + scale_x_discrete(limits = levels(as.factor(aux$var))) +
  theme(legend.position = "none")

   p + geom_jitter(
  data = filter(aux, ids %in% tree),
   aes(
    x = Variables ,
    y = Abs.importance ,
    group = ids,
    colour = "red"
  ),
  height = 0,
  size = I(3)
) + facet_grid(node ~ .)  +
  theme(legend.position = "none") + labs(y = "Absolute coefficient")
}

   pimp <- impotree(crab, ppf )

# This plot show a scatter plot for each of the two variables used in the node partition. The line represents the boundary defined
# in each node partition.

# function to select the 2 variables used in each node partition in each tree
dat_2d <- function(node) {
  indtree <- ppf[["output.trees"]][[tree]]
  sel <- (1:(ncol(ppf$train) - 1))[indtree$projbest.node[node, ] !=0] #variables used in the specific node partition
  aux <- ppf$train[ppf$oob.obs[node, ] == 0, c("Type", colnames(ppf$train[, -1])[sel[1]], colnames(ppf$train[, -1])[sel[2]])]
  aux
}

myColors <- brewer.pal(length(unique(ppf$train[, ppf$class.var])), "Dark2")

#This function  defines the intercept and the slope of the decision boundary for the scatterplot with the two selected variables, for each node partition.
# Since in each node there is a two step projection one to transform the problem in a two clkass problm
# and the second to separate the two new classes.
#The decision are made if A'X >c asign to right node else to left.
#  a1x1 + a2x2<c (c based on rule 1) we are using Rule 1 then we define the splope as - cutoff is is based on rule 1 -
# a1x1 < c -a2x2
#x1=(c -a2x2)/a1
# c/a1 is the intercept
# and a2/a1 is the slope

interslope <- function(node){
  indtree <- ppf[["output.trees"]][[tree]]
  ppf$train[,-1]
  sel <- (1:(ncol(ppf$train) - 1))[indtree$projbest.node[node, ] !=0] # select the oob observations for selected tree
  intercept <- indtree$splitCutoff.node[node, 1] / indtree$projbest.node[node, sel[1]]
  slope <- -indtree$projbest.node[node, sel[2]] / indtree$projbest.node[node, sel[1]]

  return(list(intercept,slope))
  }

d_boundary <- function(node){
  indtree <- ppf[["output.trees"]][[tree]]
  sel <- (1:(ncol(ppf$train) - 1))[indtree$projbest.node[node, ] !=0]
  if(node%in%1) dat = dat_2d(node)
  if(node%in%2) dat = dat_2d(node) %>%  filter(Type %in% c("OrangeFemale", "OrangeMale") )
  if(node%in%3) dat = dat_2d(node) %>%  filter(Type %in% c("BlueFemale", "BlueMale") )

  p1 <- ggplot(data = dat) + geom_point(ggplot2::aes_string(y = (colnames(dat_2d(node ))[[2]]),
                                                            x = colnames(dat_2d(node ))[[3]],
                                                            color =  colnames(dat_2d(node ))[[1]]),
                                        alpha = .5, size = I(3)) + xlim(c(min(dat[,2],dat[,3]),max(dat[,2],dat[,3]))) + ylim(c(min(dat[,2],dat[,3]),max(dat[,2],dat[,3]) ))+
    geom_abline(intercept = interslope(node)[[1]], slope = interslope(node)[[2]] ,size = 1 ) +
    labs(color = "Class") + theme( aspect.ratio = 1) + labs(color = "Class")

  if(node%in%1){
    p1 <- p1 +  scale_colour_brewer(name = "Class", type = "qual", palette = "Dark2")
  }
  if(node%in%2){
    p1 <-   p1 +  scale_color_manual(values = myColors[3:4])
  }
  if(node%in%3){
    p1<-  p1 +  scale_color_manual(values = myColors[1:2])
  }
  p1
}

p1 <-  d_boundary(1)
p2 <-  d_boundary(2)
p3 <-  d_boundary(3)

# combined plot with importance variable for each tree and node partition and SP plot with boundary in each node

  grid.arrange(pimp, arrangeGrob(p1, p2, p3, ncol=1), ncol=2)

@
\caption{Visualising the variable importance of all the trees in the forest, for three nodes. Each node of each tree has an importance value for each variable. The values for the whole forest are displayed using a side-by-side jittered dot plot. The importance values are the absolute values of the projection coefficients. The red points correspond to these values for the tree shown in Figure \ref{pptree}. Two variables are randomly selected at each node for creating the best projection, and split. The plots are right show the variables used and the split made at each of the nodes of this tree.  \label{impotree}}
\end{figure}

\subsection{Similarity of cases}

For each tree, every pair of observations can be in the same terminal node or not. Tallying this up across all trees in a forest gives the proximity matrix, an $n\times n$ matrix of the proportion of trees that the pair share a terminal node. It can be considered a similarity matrix.

Multidimensional scaling (MDS) is used to reduce the dimension of this matrix, to view the similarity between observations. MDS transforms the data set into a low-dimensional space where the distances are approximately the same as in the full $n$ dimensions. With $G$ groups, the low-dimensional space should be no more than $G-1$ dimensions. Figure \ref{prox1} shows the MDS plots for the 3D space induced by the four groups of the crab data. Color indicates the true species and sex. For this data two dimensions are enough to see the four groups separated quite well. Some crabs are clearly more similar to a different group, especially in examining the sex differences.

\begin{figure}[hbpt]
\centering
<<mds, echo=FALSE, warning=FALSE, message=FALSE, fig.height=4, fig.width=7.5>>=
  #Multidimensional scaling plot using proximity matrix information from PPforest for CRAB
mdspl2d <- function(ppf, lege = "bottom", siz = 3, k = 2) {
  #n.class <- length(levels(ppf[[1]])) #ppf$train %>% select_(ppf$class.var) %>% unique() %>% nrow()
  #lev <- levels(ppf[[1]]) #ppf$train %>% select_(ppf$class.var) %>% sapply(levels) %>% as.factor()

  d <- diag(nrow(ppf$train))
  d <- as.dist(d + 1 - ppf$proximity)
  rf.mds <- stats::cmdscale(d, eig = TRUE,  k = k)
  colnames(rf.mds$points) <- paste("MDS", 1:k, sep = "")
  #nlevs <- nlevels(ppf$train[, 1])

  df <- data.frame(Class = ppf$train[, 1], rf.mds$points)
  #if (lege) {
  mds <- ggplot2::ggplot(data = df) +
   geom_point(ggplot2::aes(x = MDS1, y = MDS2, color = Class),
              size = I(siz), alpha = .5) +
   scale_colour_brewer(type = "qual", palette = "Dark2", name = "Class") +
   theme(legend.position = lege, aspect.ratio = 1)
  #}
  #else  {
  #  mds <- ggplot2::ggplot(data = df) +
  #    geom_point(ggplot2::aes(x = MDS1, y = MDS2, color = Class),
  #               size = I(siz), alpha = .5) +
  #    scale_colour_brewer(type = "qual", palette = "Dark2", name = "Class") +
  #    theme(legend.position = "none", aspect.ratio = 1)
  #}
  mds
}

grid_arrange_shared_legend <- function(..., ncol = length(list(...)), nrow = 1, position = c("bottom", "right")) {

  plots <- list(...)
  position <- match.arg(position)
  g <- ggplotGrob(plots[[1]] + theme(legend.position = position))$grobs
  legend <- g[[which(sapply(g, function(x) x$name) == "guide-box")]]
  lheight <- sum(legend$height)
  lwidth <- sum(legend$width)
  gl <- lapply(plots, function(x) x + theme(legend.position="none"))
  gl <- c(gl, ncol = ncol, nrow = nrow)

  combined <- switch(position,
       "bottom" = arrangeGrob(do.call(arrangeGrob, gl),
                   legend,
                   ncol = 1,
                   heights = unit.c(unit(1, "npc") - lheight, lheight)),
       "right" = arrangeGrob(do.call(arrangeGrob, gl),
                   legend,
                   ncol = 2,
                   widths = unit.c(unit(1, "npc") - lwidth, lwidth)))
  #grid.newpage()
  grid.draw(combined)
}

k=3
lege = "none"
siz = 3
d <- diag(nrow(ppf$train))
d <- as.dist(d + 1 - ppf$proximity)
rf.mds <- stats::cmdscale(d, eig = TRUE,  k = k)
colnames(rf.mds$points) <- paste("MDS", 1:k, sep = "")
df <- data.frame(Class = ppf$train[, 1], rf.mds$points)
p12 <-  ggplot2::ggplot(data = df) +
  geom_point(ggplot2::aes(x = MDS1, y = MDS2, color = Class),
             size = I(siz), alpha = .5) +
  scale_colour_brewer(type = "qual", palette = "Dark2", name = "Class") +
  theme(legend.position = lege, aspect.ratio = 1)
p13 <-  ggplot2::ggplot(data = df) +
  geom_point(ggplot2::aes(x = MDS1, y = MDS3, color = Class),
             size = I(siz), alpha = .5) +
  scale_colour_brewer(type = "qual", palette = "Dark2", name = "Class") +
  theme(legend.position = lege, aspect.ratio = 1)
p23 <-  ggplot2::ggplot(data = df) +
  geom_point(ggplot2::aes(x = MDS2, y = MDS3, color = Class),
             size = I(siz), alpha = .5) +
  scale_colour_brewer(type = "qual", palette = "Dark2", name = "Class") +
  theme(legend.position = lege, aspect.ratio = 1)

grid_arrange_shared_legend(p12, p13, p23, ncol=3, nrow=1)

  @
\caption{Examining similarity between cases, using pairwise plots of multidimensional scaling into 3D. It can be seen that most cases are grouped closely with their class, and particularly that the two species are distinct. There is more confusion of cases between the sexes. \label{prox1}}
\end{figure}

\subsection{Uncertainty of cases}

The vote matrix ($n \times G$) contains the proportion of times each observation was classified to each class while oob. Two approaches to visualize the vote matrix information are used.

A ternary plot is a triangular diagram used to display compositional data with three components. More generally, compositional data can have any number of components, say $p$, and hence is constrained to a $(p-1)$-D simplex in $p$-space. The vote matrix is an example of compositional data with $G$ components.

With G classes, the ternary plot idea is generalized to a $(G-1)-D$ simplex~\citep{sutherland2000orca, schloerke}. This is one of the approaches used to visualize the vote matrix.

For the crab data, $G=4$ and the generalized ternary plot will be a tetrahedron. The  \pkg{tourr} package~\citep{wickham2011tourr} can be used to view it (e.g. \url{https://vimeo.com/170522736}).

Figure \ref{tetra} shows the tetrahedron structure for the crab vote matrix shown in three pairwise views. With well-separated classes, the colored points will each concentrate into one of the vertices. This is close but not perfect, indicating some crabs are commonly incorrectly predicted.

\begin{figure}[hbpt]
\centering
<<ternary, echo=FALSE, warning=FALSE, message=FALSE, fig.height=4, fig.width=7.5, out.extra='trim=0cm 2cm 0cm 2cm, clip'>>=
f.helmert <- function(d)
{
  helmert <- rep(1 / sqrt(d), d)
  for (i in 1:(d - 1))
  {
    x <- rep(1 / sqrt(i * (i + 1)), i)
    x <- c(x, -i / sqrt(i * (i + 1)))
    x <- c(x, rep(0, d - i - 1))
    helmert <- rbind(helmert, x)
  }

  return(helmert)
}

makePairs <- function(dat, id = c(a, b, c)) {
  aux <- dat[,-c(1, 2)]

  d <- aux[, id]

  grid <- expand.grid(x = id, y = id)
  grid <- subset(grid, x != y)
  all <- do.call("rbind", lapply(1:nrow(grid), function(i) {
    xcol <- grid[i, "x"]
    ycol <- grid[i, "y"]
    data.frame(
      Class = dat[, 1],
      ids = dat[, 2],
      x = dat[, xcol+2],
      y = dat[, ycol+2],
      pair = paste(grid[i, ], collapse = '-')
    )
  }))

  all
}

#ppf PPforest object
#V1,v2,v3 select the 3 proj directions
ternarydata <- function(ppf, v1, v2, v3){
  n.class <- ppf$train %>% select_(ppf$class.var) %>% unique() %>% nrow()
  projct <- t(f.helmert(length(unique(ppf$train[, ppf$class.var])))[-1,])

  dat3 <-
    data.frame(
      Class = ppf$train[, ppf$class.var],
      ids = 1:nrow(ppf$train),
      proj.vote = as.matrix(ppf$votes) %*% projct
    )

  ##with 3 or less classes
  empt <- rep(1:nrow(dat3), 3)
  #dat3.empt <- dat3[empt, ] %>% mutate(rep = rep(1:3, each = nrow(dat3)))
  if (n.class > 3) {
    gg1 <-  makePairs(dat3, c(v1,  v2, v3))
  }

  gg1 <-  makePairs(dat3, id = c(v1, v2, v3))

  return(gg1)
}

#need to run f_hermite
f_composition <- function(data) {
  d <- dim(data)[2]
  hm <- f.helmert(d)
  x <- data - matrix(1 / d, dim(data)[1], d)
  return((x %*% t(hm))[,-1])
}

simplex <- function(p = 3) {
  vert <- f_composition(diag(p + 1))
  colnames(vert) <- paste0("d", 1:ncol(vert))

  wires <-
    do.call(expand.grid, list(c(1:nrow(vert)), c(1:nrow(vert))))

  structure(list(points = vert,
                 edges = wires[!(wires[, 1] == wires[, 2]),]))
}

##ternary plot
ternaryshell <- function(ppf, sp = 3, dx = 1, dy = 2, v1=1, v2=2, v3=3){
  s <- simplex(sp)
  pts <- data.frame(s$points)
  #pts <- data.frame(s$points[,c(dx,dy)])
  gg1 <- ternarydata(ppf, v1, v2, v3)

  edg <- data.frame(x1=pts[,dx][s$edges[,1]], x2=pts[,dx][s$edg[,2]],
                    y1=pts[,dy][s$edg[,1]], y2=pts[,dy][s$edg[,2]])

  p1  <- gg1 %>% filter(pair %in% paste(dx,dy, sep="-") ) %>%
    select(Class, x, y) %>%
    ggplot(aes(x, y, color = Class)) +
      geom_segment(data = edg, aes(x = x1, xend = x2,
                                 y = y1, yend = y2), color = "black" ) +
      geom_point(size = I(3), alpha = .5) +
      labs(y = "",  x = "") +
      theme(legend.position = "none", aspect.ratio = 1) +
      scale_colour_brewer(type = "qual", palette = "Dark2") +
      labs(x = paste0("T",dx,""), y = paste0("T",dy,"")) +
    theme(aspect.ratio=1)

  p1
}
t1 <- ternaryshell(ppf, 3, 1, 2)
t2 <- ternaryshell(ppf, 3, 1, 3)
t3 <- ternaryshell(ppf, 3, 2, 3)

grid.arrange(t1, t2, t3, ncol = 3)

@
\caption{Generalized ternary plot ((G-1)-D simplex, here it is a tetrahedron) representation of the vote matrix for four classes. The tetrahedron is shown pairwise. Each point corresponds to one observation and color is the true class. This is close but not a perfect classification, since the colors are concentrated in the corners and there are some mixed colors in each corner.}
\label{tetra}
\end{figure}

Because visualizing the vote matrix with a $(G-1)$-D tetrahedron requires dynamic graphics, a low-dimensional option is also provided. For each class, each case has a value between 0-1. A side-by-side jittered dotplot is used for the display, where class is displayed on one axis and proportion is displayed on the other. For each dotplot, the ideal arrangement is that points are concentrated at 0 or 1 and only at 1 for their true class. This data is close to the ideal but not perfect, e.g. there are a few blue male crabs (orange) that are frequently predicted to be blue females (green), and a few blue female crabs predicted to be another class.

\begin{figure}[hbpt]
\centering
<<side, echo=FALSE, warning=FALSE, message=FALSE, fig.height=4,fig.width=4>>=

#side-by siede boxplot
side <-  function(ppf, ang = 0, lege = "bottom", siz = 3,
                  ttl = "Side by side dotplot") {
  voteinf <- data.frame(ids = 1:length(ppf$train[, 1]), Type = ppf$train[, 1],
                      ppf$votes, pred = ppf$prediction.oob ) %>%
  gather(Class, Probability, -pred, -ids, -Type)

  ggplot(data = voteinf, aes(Class, Probability, color = Type)) +
    geom_jitter(height = 0, size = I(siz), alpha = .5) +
    ggtitle(ttl) +
    scale_colour_brewer(type = "qual", palette = "Dark2") +
    theme(legend.position = lege, legend.text=element_text(angle = ang)) +
    labs(colour = "Class", y = "Proportion")
}

side(ppf, ttl="")

@
\caption{Another representation of the vote matrix, as a jittered side-by-side dotplot. It is not as elegant as the ternary plot, but it is useful because it places focus on each group. Each dotplot shows the proportion of times the case was predicted into the group, with 1 indicating that the case was always predicted to the group and 0 being never. On each dotplot, a single color dominates the top, indicating fairly clear distinctions between classes. Crabs from the blue species, both male and female, have more uncertainty in predictions, as seen by more crabs from other classes having higher vote proportions, than is seen in the orange species.}
\label{sideby}
\end{figure}

\section{Interactive web app }\label{app}

Interaction is added to the plots described in Section \ref{visen} and other plots, and they are organized into an interactive web app using \pkg{shiny}~\citep{chang11shiny} for exploring the ensemble model. The app is organized into three tabs: individual cases; models; and performance comparison, to provide a model diagnostic tool. Interaction is provided as mouse-over labeling, mouse-click selection, and brushing, with results linked across multiple plots. The app takes advantage of new tools provided in the \pkg{plotly}~\citep{plotly} package, developed as a part of Sievert's PhD thesis research~\citep{sievertthesis}.

The \pkg{plotly} functions directly translate a static \pkg{ggplot2}  object by  extracting  information  and  storing it in JavaScript Object Notation (JSON). This information is passed as input to a javascript function to produce a web graphic.  Interactions in a single display, and links between different graphics are two key tasks an interactive visualization should accomplish~\citep{xie2014reactive}.

As \cite{sievertthesis} describes, one of the biggest difficulties for the app in order manage linking between plots is the data structure management for each widget. Each widget has it own data structure and interaction. Putting them into the structure of a shiny app facilitates access to the widget data and coordinates selections across multiple plots.

The fishcatch data~\citep{fishcatch} is used to illustrate the shiny app characteristics. It contains 159 observations, with 6 physical measurement variables, and  7 types of fish, all caught from the same lake (Laengelmavesi) near Tampere in Finland. There are 35 bream, 11 parkki, 56 perch, 17 pike, 20 roach, 14 smelt and 6 whitewish. The shiny app showing fishcatch data can be accessed at \url{https://natydasilva.shinyapps.io/shinyppforest}.

\subsection{Individual cases}

This tab is designed to examine uncertainty in the classification of observations and to explore the similarity between pairs of observations. The data feeding the display is an $n\times p$ data frame, containing the original data, the model statistics generated from the full $n\times G$ vote matrix along with its generalized ternary coordinates, and the first two MDS projections of the proximity matrix.
Figure \ref{tab1} shows the arrangement of plots.
The plots in the tab are (1) a parallel coordinate plot (PCP) of the data, (2) the MDS display of the proximity matrix, (3) side-by-side jittered dotplot, and (4) generalized ternary plot of the vote matrix. Each of these plots are interactive in the sense that each one presents individual interactions (mouse-over) and they are linked so that selections in one display are propagated to other plots (clicking and selecting).

This selection of plots enables aspects of the model, relating to performance for individual cases, to be examined in the data space. The data plot is an essential element following the {\em model-in-the-data-space} philosophy of \citet{wickham2015visualizing}. The choice was made to use a parallel coordinate plot because it provides a space-efficient display of the data. Alternatives include the tour, a dynamic plot, or a scatterplot matrix. Theoretically, either of these could be substituted or added.

\begin{figure}[hbpt]
\centering
\includegraphics[width=1\linewidth]{figure/fish11.png}
\caption{Entry page for the web app, focusing on examining cases, in terms of similarity and uncertainty. The top plot shows the data, the remaining plots show similarity of cases and uncertainty in predictions. All of the plots are linked, so that selecting elements of one will generate changes in the others. \label{tab1}}
\end{figure}

The diagram in Figure \ref{tab1diag} illustrates the data pipeline \citep{BAHM88,wickham2009plumbing} for the interactive graphics in the case level tab. Solid lines indicate notifications from the source data to the plots, and dashed lines indicate notification of user action on the plot, that notifies the data source of actions to take. The data table is a reactive object that has a listener associated with it. Each of the plots is reactive and has numerous listeners. When users make selections on a plot, either by clicking or group selection, a change to the data is made in terms of an update on the selected cases. This invokes a note to other plots to re-draw themselves. The linking between plots is effectively one-to-one, based on the row id of the data. The side-by-side jittered dotplot has $n\times G$ points, but selection can only be done within a dotplot. Selecting in one of the dotplots notifies the data table of the selection which triggers a re-draw of the other dotplots. Mouseovers on the plot pull additional information about the point or line under the cursor but doesn't link between plots.

Two alternatives can be selected in shiny to draw the parallel coordinate plot: parallel or enhanced. Parallel draws the classic PCP and enhanced draws a modified version where variables are repeated \citep{hurley2011eulerian}.  Because reading a PCP is really only possible for neighboring variables, the variables are repeated so that all variables are neighboring.

\begin{figure}[hbpt]
\centering
\includegraphics[width=1\linewidth]{figure/fishT1.pdf}
\caption{Schematic diagram illustrating the interactivity in and between plots for individual level exploration panel of the web app. From the data, the four plots are generated. Each plot has a listener attached which collects user actions. When a user selects a point or line in any of the displays, it makes a change in the data which propagates updates to each of the plots.}
\label{tab1diag}
\end{figure}

\subsection{Models}

This second tab in the app focuses on teasing apart the forest to examine the qualities of each tree. For each tree, information on the variable importance, the projections used at each node, and the OOB error is available.  The data feeding into this tab is a list of models along with the original data frame.
The tree id is displayed when we mouse over the jittered side-by-side plot. This information is useful because, based on the accuracy, some trees could be pruned from the forest outside of the app.

Figure \ref{tab2} is a screenshot of the models tab. There are five plots, with varying levels of interaction: (1) a jittered side-by-side dotplot showing variable importance for the top three nodes of all trees in the forest, (2) a static display of one tree, (3) a boxplot of OOB error for all trees, (4) a faceted density plot of projected data at each node of the tree, with split point indicated by a vertical line, and (5) a mosaic plot showing the confusion matrix for each node of the tree.  The interaction is driven from the variable importance plot -- when the user selects a point in that display, the corresponding tree density displays and mosaic plots are drawn. The tree plot from the \pkg{PPtreeViz} is used to visualize the selected tree structure. Also highlighted are the variable importance values for each variable for each of the top three nodes and the OOB error value for the tree on the boxplot.

\begin{figure}[hbpt]
\centering
\includegraphics[width=1\linewidth]{figure/fish21.png}
\caption{The individual model tab in the web app. Variable importance is displayed as jittered dot plots for three nodes of all trees. This is linked to a display of the PPTree, a boxplot of the error for all trees in the forest, and display of the data showing splits at each of three nodes and confusion tables as mosaic plots. Clicking a point in the jittered dotplot triggers various updates: each of the importance values for the same tree are highlighted (red), the tree that this corresponds to is drawn, the error for the tree is shown on the boxplot (in red), and the data displays are updated to show the tree. \label{tab2}}
\end{figure}

The diagram in Figure~\ref{tab2diag} illustrates the data pipeline for interactive graphics.
The data source is a \pkg{PPforest} object. Interaction is driven by the variable importance plot. Selecting a point triggers a change in the data which cascades to re-draws of the other displays. Each plot has some information available on mouse over.

\begin{figure}[hbpt]
\centering
\includegraphics[width=1\linewidth]{figure/fishT2.pdf}
\caption{Schematic diagram illustrating the interactivity in and between plots for model level exploration panel of the web app. Only the dotplot of variable importance is has click selection, which invokes changes of the tree display, boxplot, density plots and mosaic plots. Selecting a point, makes a change in the data, which propagates the importance values for other variables in this tree to be highlighted (red), draws the tree, highlights the error value of the tree, and shows the projections and confusion matrix for the three top nodes.}
\label{tab2diag}
\end{figure}

\subsection{Performance comparison}

The third tab (Figure \ref{tab3}) examines the PPF fit and compares the result with a RF fit. There are four displays for each type of model: (1) Variable importance for all trees in the forest (same as in the models tab), (2) an receiver operating characteristic curve (ROC) curve comparing sensitivity and specificity for each class, (3) OOB error by number of trees to assess complexity, (4) overall variable importance. There is very little interaction on this tab. Users can select to focus on a subset of classes or choose the importance measure to show. Being able to focus on class can help to better understand how well the model performs across classes and the focus can be especially useful for unbalanced data. Examining the OOB error by trees enables an assessment of how few trees might be used to provide an equally accurate prediction of future data.

The  ROC is used to summarize the trade-off between sensitivity and specificity. The plot shows the sensitivity and specificity when a parameter of classifier is varied \citep{trevor2011elements}. The specificity and sensitivity was computed with the \pkg{pROC} package.
If more than two classes are available a multi-class ROC analysis is needed. Several solutions have been proposed for multi-class ROC. Some of the proposed reduced the multi-class problem to a set of binary problems. The approach used for a multi-class ROC analysis in this paper is called one-against-all ~\citep{allwein2000reducing}.

\begin{figure}[hbpt]
\includegraphics[width=.8\linewidth]{figure/fish31.png}
\caption{Performance comparison tab of the web app. ROC curves displaying sensitivity against specificity for the classes are shown, along with the OOB error by number of trees used to build the forest, and overall variable importance. Displays are shown for the PPF and RF, for comparison purposes. Users can select a class to focus on, using the text entry box. \label{tab3}}
\end{figure}
\newpage

\section{Discussion}\label{fur}

Having better tools to open up black box models will provide for better understanding the data, the model strengths and weaknesses, and how the model will performs on future data. This visualisation app provides a selection of interactive plots to diagnose PPF models. This shell could be used to make an app for other ensemble classifiers. The philosophy underlying the collection of displays is ``show the model in the data space''  explained in \citet{wickham2015visualizing}. It is not easy to do this, and to completely take this on would require plotting the model in the $p$-dimensional data space. In the simplest approach, as taken here, it means to link the model diagnostics to displays of the data. Then it is possible to probe and query to obtain a better understanding such as finding regions in the data that prove difficult to fit, and detract from the predictive accuracy, or that don't adhere to model assumptions.

The app is implemented with new technology for interactive graphics provided by the \pkg{plotly} package. It is one of the first uses of these new tools.

One challenge to use plotly is that  when layers with different data are created in a ggplot2,  it is difficult to  specify the unique keys required for linking with another plot.

There are many possible extensions to the app, that could help it to be a tool for model refinement: (1) Using the diagnostics to weed out under-performing models in the ensemble; (2) Identifying and boosting models that perform well, particularly if they do well for problematic subsets of the data; (3) Problematic cases could be removed, and ensembles re-fit; (4) Classes as a whole could be aggregated or re-organised as suggested by the model diagnostics, to produce a more effective hierarchical approach to the multiple class problem. Working within the R environment makes all of these desires available using command line outside the app, given the unique ids of models and cases can be exported from the app.

The app has helped to identify ways to improve the  PPtree algorithm and consequently, the PPF model. These especially apply to multiclass problems. Multiple splits for the same class would enable nonlinear classifications. Split criteria tend to place boundaries too close to some groups, due to heteroskedasticity being induced by aggregating classes. Forests are not always better than their constituent trees, and if the trees can be built better, the forest will provide stronger predictions.

\bibliographystyle{spr-chicago}
\bibliography{ppforestpaperbib}
\end{document}

### 2021-02-01 PPforestVizupdate.Rnw
update
Include comments to work later
update bib with PPforest paper

@@
 \section*{Abstract}
-This paper describes structuring data and constructing plots to explore forest classification models interactively.
+This article describes structuring data and constructing plots to explore forest classification models interactively.
 A forest classifier is an example of an ensemble since it is produced by bagging multiple trees.
 The process of bagging and combining results from multiple trees produces numerous diagnostics which, with interactive graphics, can provide a lot of insight into class structure in high dimensions.
-Various aspects are explored in this paper, to assess model complexity, individual model contributions, variable importance and dimension reduction, and uncertainty in prediction associated with individual observations.
+Various aspects are explored in this article, to assess model complexity, individual model contributions, variable importance and dimension reduction, and uncertainty in prediction associated with individual observations.
 The ideas are applied to the random forest algorithm and projection pursuit forest, but could be more broadly applied to other bagged ensembles.
@@
 
-This paper describes structuring interactive graphics to facilitate visual exploration of ensemble classifiers using RFs and projection pursuit forests (PPF)~\citep{dasilvappforest} as examples.
+This article describes structuring interactive graphics to facilitate visual exploration of ensemble classifiers using RFs and projection pursuit random forests (PPF)~\citep{da2021projection} as examples.
 The PPF algorithm builds on the projection pursuit tree (PPtree)~\citep{lee2013pptree} algorithm which uses projection pursuit at each tree node to find the best linear combination of variables to separate the classes.
@@
 
-The paper is structured as follows.
+% Should I include some references about interpretable machine learning
+
+The article is structured as follows.
 Section \ref{key} describes the ensemble components to be accessed.
@@
 
-\section{Diagnostics in forest classifiers}\label{key}
+\section{Diagnostics in Forest Classifiers}\label{key}
 
@@
 
-\section{Mapping ensemble diagnostics to visual components}~\label{visen}
+\section{Mapping Ensemble Diagnostics to Visual Components}~\label{visen}
 
@@
 
-\subsection{Individual models: PPtree}
+\subsection{Individual Models: PPtree}
 
 The PPF is composed of individual projection pursuit trees.
-Figure \ref{pptree} shows a visual ensemble of plots of a tree model on the crab data.
+Figure \ref{pptreeviz1} shows a visual ensemble of plots of a tree model on the crab data.
 There are three nodes for the four class problem.
@@
 
-\subsection{Variable importance}
+\subsection{Variable Importance}
 
@@
 
-\subsection{Similarity of cases}
+\subsection{Similarity of Cases}
 
@@
 Multidimensional scaling (MDS) is used to reduce the dimension of this matrix, to view the similarity between observations.
-MDS transforms the data set into a low-dimensional space where the distances are approximately the same as in the full $n$ dimensions.
+MDS transforms the dataset into a low-dimensional space where the distances are approximately the same as in the full $n$ dimensions.
 With $G$ groups, the low-dimensional space should be no more than $G-1$ dimensions.
@@
 
-\subsection{Uncertainty of cases}
+\subsection{Uncertainty of Cases}
 
@@
 
-\section{Interactive web app }\label{app}
+\section{Interactive Web App }\label{app}
+%Todo the PPforestViz pakage
 
@@
 
-\subsection{Individual cases}
+\subsection{Individual Cases}
 
@@
 \subsection{Models}
+%I have to modify the selecction of the tree instead to select the tree based on the importance variable will be better select the tree based on the ooberror.
+Move the boxplot and select using thisone.
+%Include some alternative visualization fro Random forest
+%This is the oly tabset that can not be reproduced using RandomForest because we don't have the same information than in PPforest.
+Also doesn't make sense present the 1D projection for each of the node partitions.
 
@@
 
-\subsection{Performance comparison}
+\subsection{Performance Comparison}
 
@@
 Some of the proposed reduced the multi-class problem to a set of binary problems.
-The approach used for a multi-class ROC analysis in this paper is called one-against-all ~\citep{allwein2000reducing}.
+The approach used for a multi-class ROC analysis in this article is called one-against-all ~\citep{allwein2000reducing}.
 

### 2021-02-02 PPforestVizupdate.Rnw
change variable importance secction and some refs

@@
 It also samples variables at each tree node.
-These produce diagnostics in the form of uncertainty in predictions for each observation, importance of variables for the prediction, predictive error for future samples based on out-of-bag (OOB) case predictions, and similarity of observations based on how often they group together in the trees.
+These produce diagnostics in the form of uncertainty in predictions for each observation, importance of variables for the prediction, predictive error for future samples based on out-of-bag (oob) case predictions, and similarity of observations based on how often they group together in the trees.
 
@@
 Predicting the response for these cases gives a better estimate for the error of the model with future data.
-The OOB error rate is a measure for each model that is combined in the ensemble and is used to provide the overall error of the ensemble.
+The oob error rate is a measure for each model that is combined in the ensemble and is used to provide the overall error of the ensemble.
   \item {\em Uncertainty measure for each observation:} Across individual (classification) models we can compute the proportion of times that a case is predicted to be each class.
@@
 
-The PPF algorithm calculates variable importance in two ways: (1) permuted importance using accuracy,  and (2) importance based on projection coefficients of standardized variables.
+The PPF algorithm calculates variable importance in three ways: permuted importance using accuracy,  and two importance measures based on projection coefficients on standardized variables.
 
-The permuted variable importance is comparable with the measure defined in the classical random forest algorithm.
-It is computed using the OOB sample for the tree $k\;\;(B^{(k)})$ for each $X_j$ predictor variable.
-Then the
-permuted importance of the variable $X_j$ in the tree $k$ can be defined as:
+The first importance measure is comparable with the measure defined in the classical random forest algorithm.
+Importance variable is computed based on the oob cases for the tree $k\;\;(O_k)$ for each $j$ predictor variable.
+Permuted importance of the variable $j$ in the tree $k$ is defined as:
 
 \[
-IMP^{(k)}(X_j) = \frac{\sum_{i \in B^{(k)} } I(y_i=\hat y_i^{(k)})-I(y_i=\hat y_{i,P_j}^{(k)})}{|B^{(k)}|}
+VI^{(1)} _{jk} = \frac{\sum_{i \in O_k } I(y_i=\hat y_{ik})-I(y_i=\tilde y_{ik})}{\# O_k}
 \]
 
-\noindent where $\hat y_i^{(k)}$
-is the predicted class for observation $i$ in tree $k$ and $y_{i,P_j}^{(k)}$ is the predicted class for observation $i$ in tree $k$ after permuting the values for variable $X_j$.
+\noindent where $\hat y_{ik}$ is the predicted class for the observation $i$ in the tree $k$, and $\tilde y_{ik}$ is the predicted class for the observation $i$ in the tree $k$ after permuting the values for variable $j$.
 The global permuted importance measure is the average importance over all the trees in the forest.
-This measure is based on comparing the accuracy of classifying OOB observations using the true class with permuted (nonsense) class.
+This measure is based on comparing the accuracy of classifying oob observations using the true class with permuted (nonsense) class.
 
-For the second importance measure, the coefficients of each projection are examined.
-The magnitude of these values indicates importance, if the variables have been standardized.
-The variable importance for a single tree is computed by a weighted sum of the absolute values of the coefficients across nodes.
-The weights takes the number of classes in each node into account ($cl_{nd}$)~\citep{lee2013pptree}.
-Then the importance of the variable $X_j$ in the PPtree $k$ can be defined as:
+The coefficients of each projection are examined to define the other two importance measures.
+If the variables have been standardized, the magnitude of these values indicates importance.
+For a single PPtree the variable importanceis is computed by a weighted sum of the absolute values of the coefficients across node, then the weights take the number of classes in each node into account($G_s$)~\citep{lee2013pptree}.
 
-\[
-  IMP_{pptree}^{(k)}(X_j)=\sum_{nd = 1}^{nn}\frac{|\alpha_{nd}^{(k)}|}{cl_{nd} }
-\]
-
-Where $\alpha_{nd}^{(k)}$ is the projected coefficient for node $nd$, variable $k$, and $nn$ the total number of node partitions in the tree $k$.
-
-The global variable importance in a PPforest then can be defined in different ways. The most intuitive is the average variable importance from each PPtree across all the trees in the forest.
-
-\[
-IMP_{ppforest1}(X_j)=\frac{\sum_{k=1}^K IMP_{pptree}^{(k)}(X_j)}{K}
-\]
-Alternatively we have defined a global importance measure for the forest as a weighted mean of the absolute value of the projection coefficients across all nodes in every tree.
-The weights are based on the projection pursuit indexes in each node ($Ix_{nd}$), and 1-(OOB-error of each tree)($acc_k$).
-
-\[IMP_{ppforest2}(X_j)=\frac{\sum_{k=1}^K acc_k \sum_{nd = 1}^{nn}\frac{Ix_{nd}|\alpha_{nd}^{(k)}|}{nn }}{K}
-\]
+The global variable importance in a PPforest then can be defined in different ways.
+One way is to average the variable importance from each PPtree across all the trees in the forest.
+Alternatively, can be computed as a weighted mean of the absolute value of the projection coefficients across all nodes in every tree, for more details ~\cite{da2021projection}.
 
@@
 This information is displayed by a side-by-side jittered dot plot.
-The red dots correspond to the absolute coefficient values for the tree model of Figure \ref{pptree}.
+The red dots correspond to the absolute coefficient values for the tree model of Figure \ref{pptreeviz1}.
 The forest was built using random samples of two variables for each node, hence there are two coefficients for each node.
@@
 The importance values are the absolute values of the projection coefficients.
-The red points correspond to these values for the tree shown in Figure \ref{pptree}.
+The red points correspond to these values for the tree shown in Figure \ref{pptreeviz1}.
 Two variables are randomly selected at each node for creating the best projection, and split.
@@
 With $G$ groups, the low-dimensional space should be no more than $G-1$ dimensions.
-Figure \ref{prox1} shows the MDS plots for the 3D space induced by the four groups of the crab data.
+Figure \ref{prox1} shows the MDS plots for the three dimensional space induced by the four groups of the crab data.
 Color indicates the true species and sex.
@@
   @
-\caption{Examining similarity between cases, using pairwise plots of multidimensional scaling into 3D.
+\caption{Examining similarity between cases, using pairwise plots of multidimensional scaling into three dimensional.
 It can be seen that most cases are grouped closely with their class, and particularly that the two species are distinct.
@@
 A ternary plot is a triangular diagram used to display compositional data with three components.
-More generally, compositional data can have any number of components, say $p$, and hence is constrained to a $(p-1)$-D simplex in $p$-space.
+More generally, compositional data can have any number of components, say $p$, and hence is constrained to a $(p-1)$ dimensional simplex in $p$-space.
 The vote matrix is an example of compositional data with $G$ components.
 
-With G classes, the ternary plot idea is generalized to a $(G-1)-D$ simplex~\citep{sutherland2000orca, schloerke}. This is one of the approaches used to visualize the vote matrix.
+With G classes, the ternary plot idea is generalized to a $(G-1)- dimensional$ simplex~\citep{sutherland2000orca, schloerke}. This is one of the approaches used to visualize the vote matrix.
 
@@
 @
-\caption{Generalized ternary plot ((G-1)-D simplex, here it is a tetrahedron) representation of the vote matrix for four classes.
+\caption{Generalized ternary plot ((G-1) dimensional simplex, here it is a tetrahedron) representation of the vote matrix for four classes.
 The tetrahedron is shown pairwise.
@@
 
-Because visualizing the vote matrix with a $(G-1)$-D tetrahedron requires dynamic graphics, a low-dimensional option is also provided.
+Because visualizing the vote matrix with a $(G-1)$ dimensional tetrahedron requires dynamic graphics, a low-dimensional option is also provided.
 For each class, each case has a value between 0-1.
@@
 This second tab in the app focuses on teasing apart the forest to examine the qualities of each tree.
-For each tree, information on the variable importance, the projections used at each node, and the OOB error is available.
+For each tree, information on the variable importance, the projections used at each node, and the oob error is available.
 The data feeding into this tab is a list of models along with the original data frame.
@@
 Figure \ref{tab2} is a screenshot of the models tab.
-There are five plots, with varying levels of interaction: (1) a jittered side-by-side dotplot showing variable importance for the top three nodes of all trees in the forest, (2) a static display of one tree, (3) a boxplot of OOB error for all trees, (4) a faceted density plot of projected data at each node of the tree, with split point indicated by a vertical line, and (5) a mosaic plot showing the confusion matrix for each node of the tree.
+There are five plots, with varying levels of interaction: (1) a jittered side-by-side dotplot showing variable importance for the top three nodes of all trees in the forest, (2) a static display of one tree, (3) a boxplot of oob error for all trees, (4) a faceted density plot of projected data at each node of the tree, with split point indicated by a vertical line, and (5) a mosaic plot showing the confusion matrix for each node of the tree.
 The interaction is driven from the variable importance plot -- when the user selects a point in that display, the corresponding tree density displays and mosaic plots are drawn.
 The tree plot from the \pkg{PPtreeViz} is used to visualize the selected tree structure.
-Also highlighted are the variable importance values for each variable for each of the top three nodes and the OOB error value for the tree on the boxplot.
+Also highlighted are the variable importance values for each variable for each of the top three nodes and the oob error value for the tree on the boxplot.
 
@@
 The third tab (Figure \ref{tab3}) examines the PPF fit and compares the result with a RF fit.
-There are four displays for each type of model: (1) Variable importance for all trees in the forest (same as in the models tab), (2) an receiver operating characteristic curve (ROC) curve comparing sensitivity and specificity for each class, (3) OOB error by number of trees to assess complexity, (4) overall variable importance.
+There are four displays for each type of model: (1) Variable importance for all trees in the forest (same as in the models tab), (2) an receiver operating characteristic curve (ROC) curve comparing sensitivity and specificity for each class, (3) oob error by number of trees to assess complexity, (4) overall variable importance.
 There is very little interaction on this tab.
@@
 Being able to focus on class can help to better understand how well the model performs across classes and the focus can be especially useful for unbalanced data.
-Examining the OOB error by trees enables an assessment of how few trees might be used to provide an equally accurate prediction of future data.
+Examining the oob error by trees enables an assessment of how few trees might be used to provide an equally accurate prediction of future data.
 
@@
 \caption{Performance comparison tab of the web app.
-ROC curves displaying sensitivity against specificity for the classes are shown, along with the OOB error by number of trees used to build the forest, and overall variable importance.
+ROC curves displaying sensitivity against specificity for the classes are shown, along with the oob error by number of trees used to build the forest, and overall variable importance.
 Displays are shown for the PPF and RF, for comparison purposes.

### 2021-02-02 PPforestVizupdate.Rnw
JCGS preamble and \pkg changed by \verb# pkg#

@@
 \documentclass[12pt]{article}
+\usepackage{amsmath}
+\usepackage{graphicx}
+\usepackage{natbib}
+\usepackage{url} % not crucial - just used below for the URL 
 \usepackage{multirow}
@@
 \usepackage{amsfonts}
-\usepackage{amsmath}
 \usepackage{rotating}
@@
 \usepackage{tablefootnote}
-\usepackage{url}
-\usepackage[authoryear]{natbib}
-\topmargin     -1.5cm  % read Lamport p.163
-\oddsidemargin -0.04cm % read Lamport p.163
-\evensidemargin -0.04cm % same as oddsidemargin but for left-hand pages
-\textwidth     16.59cm
-\textheight     22.94cm
-\parskip       7.2pt  % sets spacing between paragraphs
-\parindent       3mm  % sets leading space for paragraphs
 \usepackage{verbatim}
@@
 \usepackage{fancyhdr}
-\fancyhf{}
-\captionsetup{font={stretch=1}}
-\renewcommand{\headrulewidth}{0pt}
-\newcommand{\pkg}[1]{{\fontseries{b}\selectfont #1}}
-\pagestyle{fancy}
-\rfoot{\thepage}
-\linespread{1.5}
-\title{Interactive Graphics for Visually Diagnosing Forest Classifiers in R}
-\author{Natalia da Silva, Dianne Cook \& Eun-Kyung Lee}
+%\pdfminorversion=4
+% NOTE: To produce blinded version, replace "0" with "1" below.
+\newcommand{\blind}{0}
+
+\newcommand{\blA}{\mbox{\boldmath {\bf A}}}
+\newcommand{\blB}{\mbox{\boldmath {\bf B}}}
+\newcommand{\blW}{\mbox{\boldmath {\bf W}}}
+\newcommand{\blX}{\mbox{\boldmath {\bf X}}}
+\newcommand{\blM}{\mbox{\boldmath {\bf M}}}
+\newcommand{\blx}{\mbox{\boldmath {\bf x}}}
+\newcommand{\bla}{\mbox{\boldmath {\bf a}}}
+\newcommand{\blalpha}{\mbox{\boldmath {\alpha}}}
+
+% DON'T change margins - should be 1 inch all around.
+\addtolength{\oddsidemargin}{-.5in}%
+\addtolength{\evensidemargin}{-.5in}%
+\addtolength{\textwidth}{1in}%
+\addtolength{\textheight}{1.3in}%
+\addtolength{\topmargin}{-.8in}%
 
 \begin{document}
+
+\def\spacingset#1{\renewcommand{\baselinestretch}%
+{#1}\small\normalsize} \spacingset{1}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\if0\blind
+{
+  \title{\bf Interactive Graphics for Visually Diagnosing Forest Classifiers in R}
+  \author{Natalia da Silva\hspace{.2cm}\\
+    Instituto de Estadística (IESTA), Universidad de la República\\
+    and \\
+    Dianne Cook \\
+    Department of Econometrics and Business Statistics, Monash University\\
+    and Eun-Kyung Lee\\
+    Department of Statistics,  Ewha Womans University }
   \maketitle
+} \fi
 
-\section*{Abstract}
+\if1\blind
+{
+  \bigskip
+  \bigskip
+  \bigskip
+  \begin{center}
+    {\LARGE\bf Interactive Graphics for Visually Diagnosing Forest Classifiers in R}
+\end{center}
+  \medskip
+} \fi
+
+\bigskip
+\begin{abstract}
 This article describes structuring data and constructing plots to explore forest classification models interactively.
@@
 The ideas are applied to the random forest algorithm and projection pursuit forest, but could be more broadly applied to other bagged ensembles.
-Interactive graphics are built in \pkg{R} using the \pkg{ggplot2}, \pkg{plotly}, and \pkg{shiny} packages.
+Interactive graphics are built in R using the \verb#ggplot2 #, \verb#plotly#, and \verb#shiny# packages.
+\end{abstract}
 
+\noindent%
+{\it Keywords:} Statistical Visualization, Interactive Visualization, Interpretable Machine Learning, Ensemble Model. 
+\vfill
+
+\newpage
+\spacingset{1.5} % DON'T change the spacing!
 \section{Introduction}
+\label{sec:intro}
+
 The random forest (RF) algorithm~\citep{breiman1996bagging} was one of the first ensemble classifiers developed.
@@
 There has been some, but not a lot of, research on visualizing classification models.
-\cite{urbanek2002exploring} presents interactive tree visualization implemented in the java software called \pkg{KLIMT} that include zooming, selection, multiple views, interactive pruning, and tree construction as well as the interactive analysis of forests of trees using treemaps.
-\cite{cutler15raft} developed a java package called \pkg{RAFT} to visualize a forest classifier, that included variable selection, parallel coordinate plots, heat maps, and scatter plots of some diagnostics.
+\cite{urbanek2002exploring} presents interactive tree visualization implemented in the java software called \verb# KLIMT# that include zooming, selection, multiple views, interactive pruning, and tree construction as well as the interactive analysis of forests of trees using treemaps.
+\cite{cutler15raft} developed a java package called \verb# RAFT# to visualize a forest classifier, that included variable selection, parallel coordinate plots, heat maps, and scatter plots of some diagnostics.
 Linking between plots is limited.
-\cite{quach2012interactive} presents interactive forest visualization using the R package \pkg{iPlots eXtreme} \citep{urbanek2011iplots} where several displays are shown in one window with some linking between them available.
+\cite{quach2012interactive} presents interactive forest visualization using the R package \verb# iPlots eXtreme # \citep{urbanek2011iplots} where several displays are shown in one window with some linking between them available.
 \cite{Silva2016} describes visualizing components of an ensemble classifier.
@@
 The PPF algorithm builds on the projection pursuit tree (PPtree)~\citep{lee2013pptree} algorithm which uses projection pursuit at each tree node to find the best linear combination of variables to separate the classes.
-The visualization approach is consistent with the framework in \cite{wickham2015visualizing}, and the implementation is built on the newest interactive graphics available in \pkg{R}\citep{RCore}.
+The visualization approach is consistent with the framework in \cite{wickham2015visualizing}, and the implementation is built on the newest interactive graphics available in R \citep{RCore}.
 The purpose is to provide readily available tools for users to explore and improve ensemble fits and obtain an intuition for the underlying class structure in data.
@@
 The density plot displays the data projection at each node, and the mosaic plot shows the confusion matrix for the nodes.
-The package \pkg{PPtreeViz} provides visual tools to diagnose a PPtree model.
+The package \verb#PPtreeViz# provides visual tools to diagnose a PPtree model.
 The PPF builds on these and modifies a little.
@@
 <<pkg, echo=FALSE, message=FALSE, warning=FALSE>>=
-# require(devtools)
-# install_version("ggplot2", version = "0.9.1", repos = "http://cran.us.r-project.org")
 library(ggplot2)
@@
 library(shiny)
-#library(plotly)
 library(GGally)
@@
 library(ggmosaic)
-#library(PairViz)
 load("ppf.RData")
@@
 Node 2 where orange females are separated from orange males indicates small number of misclassifications.
-Node 3 where blue females are separated from blue males, shows a larger misclassification than for orange specie.
-    \label{pptreeviz1}}
+Node 3 where blue females are separated from blue males, shows a larger misclassification than for orange specie.\label{pptreeviz1}}
 \end{figure}
@@
 
-For the crab data, $G=4$ and the generalized ternary plot will be a tetrahedron. The  \pkg{tourr} package~\citep{wickham2011tourr} can be used to view it (e.g. \url{https://vimeo.com/170522736}).
+For the crab data, $G=4$ and the generalized ternary plot will be a tetrahedron. The  \verb#tourr# package~\citep{wickham2011tourr} can be used to view it (e.g. \url{https://vimeo.com/170522736}).
 
@@
 
-Interaction is added to the plots described in Section \ref{visen} and other plots, and they are organized into an interactive web app using \pkg{shiny}~\citep{chang11shiny} for exploring the ensemble model.
+ Interaction is added to the plots described in Section \ref{visen} and other plots, and they are organized into an interactive web app using \verb#shiny#~\citep{chang11shiny} for exploring the ensemble model.
 The app is organized into three tabs: individual cases; models; and performance comparison, to provide a model diagnostic tool.
 Interaction is provided as mouse-over labeling, mouse-click selection, and brushing, with results linked across multiple plots.
-The app takes advantage of new tools provided in the \pkg{plotly}~\citep{plotly} package, developed as a part of Sievert's PhD thesis research~\citep{sievertthesis}.
+The app takes advantage of new tools provided in the \verb#plotly#~\citep{plotly} package, developed as a part of Sievert's PhD thesis research~\citep{sievertthesis}.
 
-The \pkg{plotly} functions directly translate a static \pkg{ggplot2}  object by  extracting  information  and  storing it in JavaScript Object Notation (JSON).
+The \verb#plotly# functions directly translate a static \verb#ggplot2# object by  extracting  information  and  storing it in JavaScript Object Notation (JSON).
 This information is passed as input to a javascript function to produce a web graphic.
@@
 The interaction is driven from the variable importance plot -- when the user selects a point in that display, the corresponding tree density displays and mosaic plots are drawn.
-The tree plot from the \pkg{PPtreeViz} is used to visualize the selected tree structure.
+The tree plot from the \verb#PPtreeViz# is used to visualize the selected tree structure.
 Also highlighted are the variable importance values for each variable for each of the top three nodes and the oob error value for the tree on the boxplot.
@@
 The diagram in Figure~\ref{tab2diag} illustrates the data pipeline for interactive graphics.
-The data source is a \pkg{PPforest} object.
+The data source is a \verb #PPforest# object.
 Interaction is driven by the variable importance plot.
@@
 The plot shows the sensitivity and specificity when a parameter of classifier is varied \citep{trevor2011elements}.
-The specificity and sensitivity was computed with the \pkg{pROC} package.
+The specificity and sensitivity was computed with the \verb#pROC# package.
 If more than two classes are available a multi-class ROC analysis is needed.
@@
 
-The app is implemented with new technology for interactive graphics provided by the \pkg{plotly} package. It is one of the first uses of these new tools.
+The app is implemented with new technology for interactive graphics provided by the \verb# plotly# package. It is one of the first uses of these new tools.
 
@@
 
-\bibliographystyle{spr-chicago}
+\bibliographystyle{jasa3}
 \bibliography{ppforestpaperbib}

### 2021-06-22 PPforestVizupdate.Rnw
update paper to COST

@@
-\documentclass[12pt]{article}
+\documentclass[smallextended,natbib]{svjour3}  
+\smartqed  % flush right qed marks, e.g. at end of proof
+\usepackage{graphicx}
+\usepackage{array,multirow,graphicx}
+\parskip 2mm           
 \usepackage{amsmath}
@@
 \usepackage{fancyhdr}
-%\pdfminorversion=4
-% NOTE: To produce blinded version, replace "0" with "1" below.
-\newcommand{\blind}{0}
-
-\newcommand{\blA}{\mbox{\boldmath {\bf A}}}
-\newcommand{\blB}{\mbox{\boldmath {\bf B}}}
-\newcommand{\blW}{\mbox{\boldmath {\bf W}}}
-\newcommand{\blX}{\mbox{\boldmath {\bf X}}}
-\newcommand{\blM}{\mbox{\boldmath {\bf M}}}
-\newcommand{\blx}{\mbox{\boldmath {\bf x}}}
-\newcommand{\bla}{\mbox{\boldmath {\bf a}}}
-\newcommand{\blalpha}{\mbox{\boldmath {\alpha}}}
-
-% DON'T change margins - should be 1 inch all around.
-\addtolength{\oddsidemargin}{-.5in}%
-\addtolength{\evensidemargin}{-.5in}%
-\addtolength{\textwidth}{1in}%
-\addtolength{\textheight}{1.3in}%
-\addtolength{\topmargin}{-.8in}%
 
@@
 
-\def\spacingset#1{\renewcommand{\baselinestretch}%
-{#1}\small\normalsize} \spacingset{1}
+\title{Interactive Graphics for Visually Diagnosing Forest Classifiers in R}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\author{Natalia da Silva \and
+        Dianne Cook\and
+        Eun-Kyung Lee}
 
-\if0\blind
-{
-  \title{\bf Interactive Graphics for Visually Diagnosing Forest Classifiers in R}
-  \author{Natalia da Silva\hspace{.2cm}\\
+%\authorrunning{Short form of author list} % if too long for running head
+
+\institute{Natalia da Silva \at
               Instituto de Estadística (IESTA), Universidad de la República\\
-    and \\
-    Dianne Cook \\
+              \email{natalia.dasilva@fcea.edu.uy}
+           \and
+          Dianne Cook \at 
           Department of Econometrics and Business Statistics, Monash University\\
-    and Eun-Kyung Lee\\
-    Department of Statistics,  Ewha Womans University }
+          \email{dicook@monash.edu}\and
+          Eun-Kyung Lee \at
+          Department of Statistics, Ewha Womans University\\
+          \email{lee.eunk@gmail.com}
+}
+
+\date{Received: date / Accepted: date}
+
 \maketitle
-} \fi
 
-\if1\blind
-{
-  \bigskip
-  \bigskip
-  \bigskip
-  \begin{center}
-    {\LARGE\bf Interactive Graphics for Visually Diagnosing Forest Classifiers in R}
-\end{center}
-  \medskip
-} \fi
-
-\bigskip
 \begin{abstract}
@@
 Various aspects are explored in this article, to assess model complexity, individual model contributions, variable importance and dimension reduction, and uncertainty in prediction associated with individual observations.
-The ideas are applied to the random forest algorithm and projection pursuit forest, but could be more broadly applied to other bagged ensembles.
-Interactive graphics are built in R using the \verb#ggplot2 #, \verb#plotly#, and \verb#shiny# packages.
+The ideas are applied to the random forest algorithm and projection pursuit forest but could be more broadly applied to other bagged ensembles.
+Interactive graphics are built in R using the \texttt{ggplot2}, \texttt{plotly}, and \texttt{shiny} packages.
+\keywords{Statistical Visualization \and Interactive Visualization \and Interpretable Machine Learning \and Ensemble Model}
 \end{abstract}
 
-\noindent%
-{\it Keywords:} Statistical Visualization, Interactive Visualization, Interpretable Machine Learning, Ensemble Model. 
-\vfill
+\newpage
 
-\newpage
-\spacingset{1.5} % DON'T change the spacing!
 \section{Introduction}
@@
 The benefits for classification are reduced variability in predictive error, and the suite of diagnostics provides the potential for better understanding the class structure in the high-dimensional data space.
-The use of visualization on these diagnostics, in association with multivariate data plots, completes the process to support a better understanding of the underlying problem.
+The use of visualization on these diagnostics, in association with multivariate data plots, completes the process to better understand the underlying problem.
 
-A conceptual framework for model visualization can be summarized in three strategies: (1) visualize the model in the data space, (2) look all members of a collection of a model and (3) explore the complete process of model fitting~\citep{wickham2015visualizing}.
-The first strategy is to explore how well the model captures the data characteristics (model in the data space), which contrasts with determining if the model assumptions hold (data in the model space).
+A conceptual framework for model visualization can be summarized in three strategies: (1) visualize the model in the data space, (2) look at all members of a collection of a model, and (3) explore the complete process of model fitting~\citep{wickham2015visualizing}.
+The first strategy is to explore how well the model captures the data characteristics (the model in the data space), which contrasts with determining if the model assumptions hold (data in the model space).
 The second strategy is to look at a group of models instead of only the best model.
@@
 There has been some, but not a lot of, research on visualizing classification models.
-\cite{urbanek2002exploring} presents interactive tree visualization implemented in the java software called \verb# KLIMT# that include zooming, selection, multiple views, interactive pruning, and tree construction as well as the interactive analysis of forests of trees using treemaps.
+\cite{urbanek2002exploring} presents interactive tree visualization implemented in the java software called \verb# KLIMT# that includes zooming, selection, multiple views, interactive pruning, and tree construction as well as the interactive analysis of forests of trees using treemaps.
 \cite{cutler15raft} developed a java package called \verb# RAFT# to visualize a forest classifier, that included variable selection, parallel coordinate plots, heat maps, and scatter plots of some diagnostics.
@@
 The purpose is to provide readily available tools for users to explore and improve ensemble fits and obtain an intuition for the underlying class structure in data.
-Interactive plots are a key component for model visualization that help the user see multivariate relationships and be more efficient in the model diagnosis.
-Multiple levels of data are constructed for exploration: observation, model and ensemble summaries.
+Interactive plots are a key component for model visualization that helps the user see multivariate relationships and be more efficient in the model diagnosis.
+Multiple levels of data are constructed for exploration: observation, model, and ensemble summaries.
 
@@
 The oob error rate is a measure for each model that is combined in the ensemble and is used to provide the overall error of the ensemble.
+
   \item {\em Uncertainty measure for each observation:} Across individual (classification) models we can compute the proportion of times that a case is predicted to be each class.
 If a case is always predicted to be the true class, there is no uncertainty about an observation.
-Cases that are proportionately predicted to be multiple classes indicate difficult to classify observations.
+Cases that are proportionately predicted to be multiple classes indicate difficulty to classify observations.
 They may be important by indicating neighborhoods of the data space that would benefit from a more complex model, or more simply, they may be errors in measurements in the data.
+
   \item {\em Variable importance:} Each model uses samples of variables.
@@
 There are several versions of statistics that use this to provide a measure of the variable importance for prediction.
+
   \item {\em Similarity measure for pairs of observations:} In each model, each pair of observations will be either in the same terminal node or not.
@@
 These are illustrated using the Australian crabs data~\citep{CM74}.
-The data has 200 cases, 5 predictors and 4 classes (combinations of species and sex, blue male, blue female, orange male and orange female).
+The data has 200 cases, 5 predictors, and 4 classes (combinations of species and sex, blue male, blue female, orange male, and orange female).
 The predictors are: FL (the size of the frontal lobe length, in mm), RW (rear width, in mm), CL (length of mid-line of the carapace, in mm), CW (maximum width of carapace, in mm), BD (depth of the body; for females, measured after displacement of the abdomen, in mm).
@@
 If the variables have been standardized, the magnitude of these values indicates importance.
-For a single PPtree the variable importanceis is computed by a weighted sum of the absolute values of the coefficients across node, then the weights take the number of classes in each node into account($G_s$)~\citep{lee2013pptree}.
+For a single PPtree the variable importance is computed by a weighted sum of the absolute values of the coefficients across nodes, then the weights take the number of classes in each node into account($G_s$)~\citep{lee2013pptree}.
 
@@
 @
-\caption{Visualising the variable importance of all the trees in the forest, for three nodes.
+\caption{Visualizing the variable importance of all the trees in the forest, for three nodes.
 Each node of each tree has an importance value for each variable.
@@
 Two variables are randomly selected at each node for creating the best projection, and split.
-The plots are right show the variables used and the split made at each of the nodes of this tree.
+The plots at right show the variables used and the split made at each of the nodes of this tree.
 \label{impotree}}
@@
 Each point corresponds to one observation and color is the true class.
-This is close but not a perfect classification, since the colors are concentrated in the corners and there are some mixed colors in each corner.}
+This is close but not a perfect classification since the colors are concentrated in the corners and there are some mixed colors in each corner.}
 \label{tetra}
@@
 For each class, each case has a value between 0-1.
-A side-by-side jittered dotplot is used for the display, where class is displayed on one axis and proportion is displayed on the other.
+A side-by-side jittered dotplot is used for the display, where the class is displayed on one axis and proportion is displayed on the other.
 For each dotplot, the ideal arrangement is that points are concentrated at 0 or 1 and only at 1 for their true class.
@@
 On each dotplot, a single color dominates the top, indicating fairly clear distinctions between classes.
-Crabs from the blue species, both male and female, have more uncertainty in predictions, as seen by more crabs from other classes having higher vote proportions, than is seen in the orange species.}
+Crabs from the blue species, both male and female, have more uncertainty in predictions, as seen by more crabs from other classes having higher vote proportions than is seen in the orange species.}
 \label{sideby}
@@
 \section{Interactive Web App }\label{app}
-%Todo the PPforestViz pakage
 
@@
 This information is passed as input to a javascript function to produce a web graphic.
-Interactions in a single display, and links between different graphics are two key tasks an interactive visualization should accomplish~\citep{xie2014reactive}.
+Interactions in a single display and links between different graphics are two key tasks an interactive visualization should accomplish~\citep{xie2014reactive}.
 
-As \cite{sievertthesis} describes, one of the biggest difficulties for the app in order manage linking between plots is the data structure management for each widget.
+As \cite{sievert2020interactive} describes, one of the biggest difficulties for the app in order manage linking between plots is the data structure management for each widget.
 Each widget has it own data structure and interaction.
@@
 The top plot shows the data, the remaining plots show similarity of cases and uncertainty in predictions.
-All of the plots are linked, so that selecting elements of one will generate changes in the others.
+All of the plots are linked so that selecting elements of one will generate changes in the others.
 \label{tab1}}
@@
 The diagram in Figure \ref{tab1diag} illustrates the data pipeline \citep{BAHM88,wickham2009plumbing} for the interactive graphics in the case level tab.
-Solid lines indicate notifications from the source data to the plots, and dashed lines indicate notification of user action on the plot, that notifies the data source of actions to take.
+Solid lines indicate notifications from the source data to the plots, and dashed lines indicate notification of user action on the plot, which notifies the data source of actions to take.
 The data table is a reactive object that has a listener associated with it.
@@
 Selecting in one of the dotplots notifies the data table of the selection which triggers a re-draw of the other dotplots.
-Mouseovers on the plot pull additional information about the point or line under the cursor but doesn't link between plots.
+Mouseovers on the plot pull additional information about the point or line under the cursor but do not link between plots.
 
@@
 \subsection{Models}
-%I have to modify the selecction of the tree instead to select the tree based on the importance variable will be better select the tree based on the ooberror.
-Move the boxplot and select using thisone.
-%Include some alternative visualization fro Random forest
-%This is the oly tabset that can not be reproduced using RandomForest because we don't have the same information than in PPforest.
-Also doesn't make sense present the 1D projection for each of the node partitions.
 
@@
 Variable importance is displayed as jittered dot plots for three nodes of all trees.
-This is linked to a display of the PPTree, a boxplot of the error for all trees in the forest, and display of the data showing splits at each of three nodes and confusion tables as mosaic plots.
+This is linked to a display of the PPTree, a boxplot of the error for all trees in the forest, and a display of the data showing splits at each of three nodes and confusion tables as mosaic plots.
 Clicking a point in the jittered dotplot triggers various updates: each of the importance values for the same tree are highlighted (red), the tree that this corresponds to is drawn, the error for the tree is shown on the boxplot (in red), and the data displays are updated to show the tree.
@@
 The diagram in Figure~\ref{tab2diag} illustrates the data pipeline for interactive graphics.
-The data source is a \verb #PPforest# object.
+The data source is a \verb#PPforest# object.
 Interaction is driven by the variable importance plot.
@@
 \caption{Schematic diagram illustrating the interactivity in and between plots for model level exploration panel of the web app.
-Only the dotplot of variable importance is has click selection, which invokes changes of the tree display, boxplot, density plots and mosaic plots.
-Selecting a point, makes a change in the data, which propagates the importance values for other variables in this tree to be highlighted (red), draws the tree, highlights the error value of the tree, and shows the projections and confusion matrix for the three top nodes.}
+Only the dotplot of variable importance has click selection, which invokes changes of the tree display, boxplot, density plots, and mosaic plots.
+Selecting a point makes a change in the data, which propagates the importance values for other variables in this tree to be highlighted (red), draws the tree, highlights the error value of the tree, and shows the projections and confusion matrix for the three top nodes.}
 \label{tab2diag}
@@
 \subsection{Performance Comparison}
-
 The third tab (Figure \ref{tab3}) examines the PPF fit and compares the result with a RF fit.
-There are four displays for each type of model: (1) Variable importance for all trees in the forest (same as in the models tab), (2) an receiver operating characteristic curve (ROC) curve comparing sensitivity and specificity for each class, (3) oob error by number of trees to assess complexity, (4) overall variable importance.
+There are four displays for each type of model: (1) Variable importance for all trees in the forest (same as in the models tab), (2) an receiver operating characteristic curve (ROC) curve comparing sensitivity and specificity for each class, (3) oob error by the number of trees to assess complexity, (4) overall variable importance.
 There is very little interaction on this tab.
@@
 The  ROC is used to summarize the trade-off between sensitivity and specificity.
-The plot shows the sensitivity and specificity when a parameter of classifier is varied \citep{trevor2011elements}.
-The specificity and sensitivity was computed with the \verb#pROC# package.
+The plot shows the sensitivity and specificity when a parameter of the classifier is varied \citep{trevor2011elements}.
+The specificity and sensitivity were computed with the \verb#pROC# package.
 If more than two classes are available a multi-class ROC analysis is needed.
@@
 \caption{Performance comparison tab of the web app.
-ROC curves displaying sensitivity against specificity for the classes are shown, along with the oob error by number of trees used to build the forest, and overall variable importance.
+ROC curves displaying sensitivity against specificity for the classes are shown, along with the oob error by the number of trees used to build the forest, and overall variable importance.
 Displays are shown for the PPF and RF, for comparison purposes.
@@
 
-\bibliographystyle{jasa3}
+\bibliographystyle{bibstyle}
 \bibliography{ppforestpaperbib}

### 2022-02-01 PPforestVizupdate.Rnw
reviewer minor comments

@@
 If a case is always predicted to be the true class, there is no uncertainty about an observation.
-Cases that are proportionately predicted to be multiple classes indicate difficulty to classify observations.
+Cases that are predicted to be multiple classes, with similar class probabilities, indicate difficulty to classify observations.
 They may be important by indicating neighborhoods of the data space that would benefit from a more complex model, or more simply, they may be errors in measurements in the data.
@@
   \item {\em Variable importance:} Each model uses samples of variables.
-So that, the accuracy of the models can be compared when the variable is included or omitted.
+Thus, the accuracy of the models can be compared when the variable is included or omitted.
 There are several versions of statistics that use this to provide a measure of the variable importance for prediction.
@@
 Node 2 where orange females are separated from orange males indicates small number of misclassifications.
-Node 3 where blue females are separated from blue males, shows a larger misclassification than for orange specie.\label{pptreeviz1}}
+Node 3 where blue females are separated from blue males, shows a larger misclassification than for orange species.\label{pptreeviz1}}
 \end{figure}
@@
 If the variables have been standardized, the magnitude of these values indicates importance.
-For a single PPtree the variable importance is computed by a weighted sum of the absolute values of the coefficients across nodes, then the weights take the number of classes in each node into account($G_s$)~\citep{lee2013pptree}.
+For a single PPtree the variable importance is computed by a weighted sum of the absolute values of the coefficients across nodes, then the weights take the number of classes in each node into account~\citep{lee2013pptree}.
 
@@
 One way is to average the variable importance from each PPtree across all the trees in the forest.
-Alternatively, can be computed as a weighted mean of the absolute value of the projection coefficients across all nodes in every tree, for more details ~\cite{da2021projection}.
+Alternatively, it can be computed as a weighted mean of the absolute value of the projection coefficients across all nodes in every tree, for more details ~\cite{da2021projection}.
 
@@
 Node 2 uses CL and RW, and RW contributes the most to the separation.
-The plot at right shows the boundary that is induced.
+The plot in the right shows the boundary that is induced.
 Node 3 uses FL and RW, and this is a much more even contribution by the two variables.
@@
   @
-\caption{Examining similarity between cases, using pairwise plots of multidimensional scaling into three dimensional.
+\caption{Examining similarity between cases, using pairwise plots of multidimensional scaling into a three-dimensional space.
 It can be seen that most cases are grouped closely with their class, and particularly that the two species are distinct.
@@
 With well-separated classes, the colored points will each concentrate into one of the vertices.
-This is close but not perfect, indicating some crabs are commonly incorrectly predicted.
+This is close but not perfect, indicating some crabs are commonly incorrectly classified.
 
@@
 
-As \cite{sievert2020interactive} describes, one of the biggest difficulties for the app in order manage linking between plots is the data structure management for each widget.
+As \cite{sievert2020interactive} describes, one of the biggest difficulties when it comes to the app managing linking between plots is the data structure management for each widget.
 Each widget has it own data structure and interaction.
@@
 This tab is designed to examine uncertainty in the classification of observations and to explore the similarity between pairs of observations.
-The data feeding the display is an $n\times p$ data frame, containing the original data, the model statistics generated from the full $n\times G$ vote matrix along with its generalized ternary coordinates, and the first two MDS projections of the proximity matrix.
+The data feeding the display is an $n\times p$ data frame; containing the original data; the model statistics generated from the full $n\times G$ vote matrix along with its generalized ternary coordinates; and the first two MDS projections of the proximity matrix.
 Figure \ref{tab1} shows the arrangement of plots.
@@
 This invokes a note to other plots to re-draw themselves.
-The linking between plots is effectively one-to-one, based on the row id of the data.
+The linking between plots is effectively one-to-one, based on the row ID of the data.
 The side-by-side jittered dotplot has $n\times G$ points, but selection can only be done within a dotplot.
@@
 Parallel draws the classic PCP and enhanced draws a modified version where variables are repeated \citep{hurley2011eulerian}.
-Because reading a PCP is really only possible for neighboring variables, the variables are repeated so that all variables are neighboring.
+Because reading a PCP is really only possible for neighboring variables, the variables are repeated so that all variable pairs are shown next to each other once.
 
@@
 The data feeding into this tab is a list of models along with the original data frame.
-The tree id is displayed when we mouse over the jittered side-by-side plot. This information is useful because, based on the accuracy, some trees could be pruned from the forest outside of the app.
+The tree ID is displayed when we mouse over the jittered side-by-side plot. This information is useful because, based on the accuracy, some trees could be removed from the forest outside of the app.
 
@@
 The third tab (Figure \ref{tab3}) examines the PPF fit and compares the result with a RF fit.
-There are four displays for each type of model: (1) Variable importance for all trees in the forest (same as in the models tab), (2) an receiver operating characteristic curve (ROC) curve comparing sensitivity and specificity for each class, (3) oob error by the number of trees to assess complexity, (4) overall variable importance.
+There are four displays for each type of model: (1) Variable importance for all trees in the forest (same as in the models tab), (2) a receiver operating characteristic curve (ROC) curve comparing sensitivity and specificity for each class, (3) oob error by the number of trees to assess complexity, (4) overall variable importance.
 There is very little interaction on this tab.
@@
 
-Having better tools to open up black box models will provide for better understanding the data, the model strengths and weaknesses, and how the model will performs on future data.
+Having better tools to open up black box models will provide for better understanding the data, the model strengths and weaknesses, and how the model will perform on future data.
 This visualisation app provides a selection of interactive plots to diagnose PPF models.
@@
 There are many possible extensions to the app, that could help it to be a tool for model refinement: (1) Using the diagnostics to weed out under-performing models in the ensemble; (2) Identifying and boosting models that perform well, particularly if they do well for problematic subsets of the data; (3) Problematic cases could be removed, and ensembles re-fit; (4) Classes as a whole could be aggregated or re-organised as suggested by the model diagnostics, to produce a more effective hierarchical approach to the multiple class problem.
-Working within the R environment makes all of these desires available using command line outside the app, given the unique ids of models and cases can be exported from the app.
+Working within the R environment makes all of these desires available using command line outside the app, given the unique IDs of models and cases can be exported from the app.
 

### 2022-02-01 response_reviewers.Rmd
reviewer response

---
title: "Authors Response to Review"
output: pdf_document
header-includes:  \usepackage{xcolor}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Our responses to the review are written in \textcolor{violet}{violet}.

## Response to Editor

Dear Mrs. da Silva:

We have received the reports from our advisors on your manuscript, "Interactive Graphics for Visually Diagnosing Forest Classifiers in R", 
Based on the advice received, the Editor feels that your manuscript could be accepted for publication should you be prepared to incorporate minor revisions.
When preparing your revised manuscript, you are asked to carefully consider the reviewer comments which are attached, and submit a list of responses to the comments.
Your list of responses should be uploaded as a file in addition to your revised manuscript.

TO VIEW REVIEWER ATTACHMENTS (should there be any), please login to the journal site as "Author" and access "Submission Needing Revision" from the Author Main Menu.
On the next page display, navigate on the "Action Links" and select "View Reviewer Attachments" from the selection box. This will redirect you to the page that will allow you to "Download" and view the reviewer report attachments. Then, proceed with revising your manuscript.

PLEASE NOTE: YOUR REVISED VERSION CANNOT BE IN .PS OR .PDF AS TYPESETTING CANNOT MANIPULATE THE FILES.
IN ORDER TO PROCEED WITH PUBLICATION YOUR ORIGINAL SOURCE FILES ARE REQUIRED.
Submissions without original source files will be returned for these prior to final acceptance.

To submit the revision please log on to the Journal's website.

Your username is: ********

If you forgot your password, you can click the 'Send Login Details' link on the EM Login page at https://www.editorialmanager.com/cost/.

Click "Author Login" to submit your revision.

We look forward to receiving your revised manuscript.

Thank you very much.

Best regards,

Springer Journals Editorial Office
%JOURNALFULLTITLE

\textcolor{violet}{Thank you. We appreciate the chance to do a further revision, and hope that these changes satisfy the AE and reviewer.}

## Response to Reviewer 2

\textcolor{violet}{We appreciate all of your feedback. 
 }

Summary of the paper

The paper introduces an app for visual exploration of tree-based ensemble models in three steps. First, the app enables the user to investigate how well the model performs when it comes to predicting each class in its observation-level functionality. Second, the tree-level section allows the user to select a tree based on a variable importance measure in one of the first three nodes and visualize the way the first three splits are made in that tree. Finally, the app visually compares the predictive performance of the model to a benchmark one, which is illustrated by comparing a projection-pursuit-based random forest to a standard one.

The paper and its accompanying app seem to be a good fit for Computational Statistics, as they apply interesting visualization techniques to a tree-based ensemble model that will surely prove useful for practitioners. However, for this potential to materialize the authors should put more emphasis on the limitations of their approach. In particular, the app focuses on the first three nodes of the trees in question, which would be much less informative in settings requiring deep trees. Furthermore, plots such as the parallel coordinate plot would lose their appeal in settings with thousands of variables. A brief discussion of circumstances limiting the usability of the app would address these concerns and allow users to easily determine whether the app is suitable for their application.

Substantive comments

\begin{enumerate}
\item p. 8, line 42: The issue of visualizing the tetrahedron is arguably not of first-order importance as it is not as helpful for classification problems with more than four classes. It might be more useful to focus exclusively on the side-by-side jittered dotplot as this approach generalizes in a more straightforward way to problems with more classes.

\textcolor{violet}{As we mentioned in page 7 the ternary plot can be generalized to more than three dimensional space, with G classes, the ternary plot idea is generalized to a G-1 dimensional simplex. Then if we have 4 classes the generalized ternary plot is a tetrahedron as shown in Figure 4 but if we have more than four classes we can also generalize the ternary plot idea. In the fishcatch data used in this paper to illustrate the shiny app there are 7 classes and we generalized the ternary plot to a 6-dimensional simplex as is shown in the last panel of  Figure 6}

\item p. 11, line 16: As section 3 discusses all other plots produced by the app it would be useful if it also included an overview of the parallel coordinate plot that only appears in section 4, so the meaning of that plot is not immediately clear (e.g. what is displayed on the vertical axis).

\item p. 13, line 38: I find the choice of the interaction driver for the model tab puzzling. I would argue that the key statistic distinguishing the trees in the ensemble is OOB error so a more natural design choice would be to include jittered points in the OOB error boxplot so that the user can explore the well or poorly performing trees in more detail. I do not have a good intuition for why the user would be interested in exploring trees based on variable importance.

\item p. 15, line 35: The OOB error plot is somewhat hard to read due to some classes being harder to classify than others, and a potential solution to that would be using a different scale on the vertical axis (e.g. percent of highest OOB error for a class).

\end{enumerate}

Minor comments

\begin{enumerate}
\item p. 6, line 17: introducing the $G\_s$ notation is unnecessary as it is not used in the remainder of the paper (also, the meaning of the subscript s is not explained).

\textcolor{violet}{I have removed $G_s$, as you notices it is unnecessary here}

\item p.6, line 21: it should be "Alternatively, it can be computed" instead of "Alternatively, can be computed".

\textcolor{violet}{Done}

\item p. 6, line 33: it should be "The plot in the right" instead of "The plot at right".

\textcolor{violet}{Done}

\item p. 8, line 20: it should be "into a three-dimensional space" instead of "into three dimensional".

\textcolor{violet}{Done}

\item p.8, line 49: crabs are "classified", not "predicted".

\textcolor{violet}{Done}

\item p. 10, line 40: the phrase "one of the biggest difficulties for the app in order manage linking" should be rewritten to, for example, "one of the biggest difficulties when it comes to the app managing linking".

\textcolor{violet}{Done}

\item p. 12, line 46: using the phrase "some trees could be removed from the forest" instead of "some trees could be pruned from the forest" would make this sentence clearer.

\textcolor{violet}{Done}

\item p. 14, line 49: "a receiver" instead of "an receiver".

\textcolor{violet}{Done}

\item p. 16, line 32: it would be useful to include a legend of the class colors.

\textcolor{violet}{TO DO}

\item p. 17, line 6: " will perform" instead of "will performs".

\textcolor{violet}{Done}

\end{enumerate}

## Response to Reviewer 4

 I think the paper overall is good. The comments below are a combination of comments on the paper itself and some suggestions for the plotting in the software (which don't have to be changed for the paper). I think also that some of the details of the plots need to be explained better.

MAIN COMMENTS:

Section 4 in general: since the big point here is that the software is interactive, I think the discussion in this section would be a lot more useful if you illustrated step by step with screenshots an example where your interactive drill-down allows you to quickly discover things. For instance, say there are some outliers in plot A. Then you click on them and see some further details in plot B that help you understand what is going on. In this section you show a bunch of screenshots and label their capabilities, but you don't actually demonstrate an end-to-end example of the interactivity. For instance, in Figure 9 you give an example of a set of actions, why don't you actually show us the results and the changes that occur in a sequence of plots? Furthermore, I would show some actual examples of screenshots of what shows when you hover over observations.

CONTENT COMMENTS:

\begin{enumerate}
\item Page 2: In describing Random Forests, in addition to randomly sampling variables to try to split on in each node, don't they also initially sample a subset of variables for each tree? If so, you should note what settings you use in this paper (how many variables are selected for each tree). Also, if the variables are randomly sampled at each node, rather than simply choosing the best projection possible for, say, any pair of variables, what effect does this have on the importance measures? What effect is there to determining the number of variables to use in each projection (e.g. 3 instead of 2)? Is this something you have experimented with?
I think the paper could also benefit from a simple preliminary example of, say, two features and two classes and showing the plot of the linear combination resulting from projection pursuit. You could even include an image from the original paper.

\item Page 4 line 48: should $O_k$ also have a subscript for the variable j? Also $\hat{y}_{ikj}$

\item Page 5 Figure 1: You mention that the variables should be standardized for the coefficients to be interpretable as importances and be comparable. But in the upper right corner, the plots of the variables distributions, for instance for node 2, are not in the standardized range (node 2 has an x-axis that goes up to 5 and 6). Is the distribution being plotted here the value of the linear combination for each class, rather than a single variable? Is there something going on here with the default x limits for doing KDE? As a side note, you show that the first cut perfectly separates the blue and orange species. You may want to consider having the bounds for the KDE be the observed range of the linear combination for each class so that the density plots won't overlap in the plot for node 1, which they seem to do more than they should if the separation is perfect.
Furthermore, if you have any control over the color scheme (if you don't I would recommend adding this as a variable), it might be a good idea in this paper to have the colors for the blue and orange species be blue and orange, and have differing shading patterns to distinguish male from female, or something similar; perhaps you can choose more similar colors within both species. Or, better yet, can you use different plotting symbols, say triangles vs circles to distinguish the species, and shaded vs hollow to distinguish the sex?
Also, can you include labeling in your plot that the sub-plots are node ID (e.g., put the label "node" above the 3 density plots).
What do the numbers in the grey boxes at the terminal nodes represent?

\item Figures 1 and 2: I see that the potential range for coefficient values for the projection pursuit seem to be between -1 and 1. I imagine this is a feature of the algorithm. Perhaps include in your description of projection pursuit a note about this fact, for those unfamiliar, and mention what optimization criteria is used in it.

\item Page 6: If you use a notation for VI for global variable importance (you define it as the average across k trees), I'd just define it here as an equation and use it in the text (e.g in line 19).

\item Page 6 line 48: does MDS need a citation? Also, can you connect the results in this discussion to the RF graphics, e.g. in terms of same variables with high RF importance being important in MDS? Otherwise, this whole discussion seems unnecessary. If you are doing it simply to understand the data, I would move this section to right after you introduce the data, before you show the other graphics. Here, it seems like a tangent. Also, are these MDS and tetrahedral plots part of your software or do you perform them separately? I would say so explicitly in either case.

\item Page 7 Figure 2: why does the left plot need to be a dot plot rather than, say, a single box or violin plot for each variable? I don't understand what role the x axis for each variable plays here. If it doesn't play any role (e.g. is just for the jittering), then don't use it.

\item Page 9, Figure 4: is the dimension of the simplex always at most 3 or G-1? What would it look like with more classes? I'd clarify whether this only works for your example or whether it can always be drawn.

\item Page 10, Figure 5: similar remark as before. I think a violin plot or something similar for each class, rather than a jittered dot plot, is more correct. The x-axis for each class is a "fake" axis. I think jittered dot plots should only be used with scatterplots to make sure individual dots can be distinguished.

\item Page 10, line 40: This sentence is muddled "As Sievert (2020) describes, one of the biggest difficulties for the app in order manage linking between plots is the data structure management for each widget". Furthermore, what does "widget" refer to here? Use a more descriptive term.

\item Page 10, line 45: If you are just going to demonstrate the interactivity, wouldn't it be better to just continue with the original crab dataset, or have one dataset throughout? Unless there is something particularly interesting about this dataset.

\item Page 12 Figure 6: in the enhanced PCP, you say all variable pairs are shown at least once as neighbors. Is there a way, either through ordering or adding dividing lines, etc. to help a reader find the pair they are interested in? Or is there some sorting done so that, the pairs are ordered from left to right in terms of some metric of "interesting-ness"? There seems to be no order in the given plot. Do you automatically pick the number of dimensions to show in MDS? Should the tetrahedral plots here have some kind of axis labels? See Figure 7 where they do.

\item Page 13 line 46: I wasn't aware that there was an option in RF to prune under-performing trees (as opposed to nodes within a tree) in an ensemble. Please elaborate or include a reference. Also, in this case, maybe a different word than "prune", such as "discard".

\item Page 16 Figure 10: in the upper LH corner, where you select the classes, do these appear as text that you type or delete, or as buttons you can toggle on and off? Seems like the latter would be better.
Also, the bottom two plots have different x axes, so it is hard to compare them (if they should be compared, the axes should be the same; perhaps you could even plot them on the same plot?). Furthermore, there are no titles for either, and they have the same axis labels so I have no idea what the difference is between them.
\end{enumerate}

STYLISTIC COMMENTS

\begin{enumerate}
\item Should out of bag be written OOB rather than oob?

\textcolor{violet}{Not necessarily, in Breiman Random Forest original paper always is written as `out-of-bag' without abbreviation. I think if we are consistent with the abbreviation we can use oob or OOB indistinctly.}

\item Page 3 line 29: when you say "proportionately predicted", do you mean something specific, like that the class probabilities are proportionate to their frequency in the data?

\textcolor{violet}{Cases that are predicted to be multiple classes, without a clear majority in any of them, indicate that those cases are difficult to classify. The paragraph was not clear and we modify it to:" Cases that are predicted to be multiple classes, with similar class probabilities, indicate difficulty to classify observations"  }

\item Page 3 line 38: These two sentences don't seem to work together: "Each model uses samples of variables. So that, the accuracy of the models can be compared when the variable is included or
omitted." Do you mean "Thus" rather than "So that"?

\textcolor{violet}{Yes, should be Thus. It is done}

\item Page 4: "Specie" in Figure 1 caption is misspelled.

\textcolor{violet}{Done}

\item Page 11 line 12: For the sentence beginning "The data feeding…", separate each clause with semicolons rather than commas.

\textcolor{violet}{Done}

\item Page 11 line 39: say "ID" or "index" rather than "id" ; same for, say, page 13 line 44

\textcolor{violet}{Done}

\item Page 11 line 49: I would modify the final sentence to say that "all variable pairs are shown next to each other once" rather than "all variables are neighbors".

\textcolor{violet}{Done}

\end{enumerate}

<!-- ******** -->

<!-- Please note that this journal is a Transformative Journal (TJ). Authors may publish their research with us through the traditional subscription access route or make their paper immediately open access through payment of an article-processing charge (APC). Authors will not be required to make a final decision about access to their article until it has been accepted. -->

<!-- Authors may need to take specific actions to achieve compliance with funder and institutional open access mandates. If your research is supported by a funder that requires immediate open access (e.g. according to Plan S principles) then you should select the gold OA route, and we will direct you to the compliant route where possible. For authors selecting the subscription publication route our standard licensing terms will need to be accepted, including our self-archiving policies. Those standard licensing terms will supersede any other terms that the author or any third party may assert apply to any version of the manuscript. -->

<!-- Find out more about compliance -->

<!-- This letter contains confidential information, is for your own use, and should not be forwarded to third parties. -->

<!-- Recipients of this email are registered users within the Editorial Manager database for this journal. We will keep your information on file to use in the process of submitting, evaluating and publishing a manuscript. For more information on how we use your personal details please see our privacy policy at https://www.springernature.com/production-privacy-policy. If you no longer wish to receive messages from this journal or you have questions regarding database management, please contact the Publication Office at the link below. -->

### 2022-02-03 response_reviewers.Rmd
response update

@@
 
+\textcolor{violet}{We included a description of parallel coordinate plot in section 4 were this figure appears and the corresponding reference for further details.}
+
 \item p.
@@
 
+\textcolor{violet}{TO DO, change selection to boxplot trees}
+
 \item p.
@@
 percent of highest OOB error for a class).
+
+\textcolor{violet}{TO DO, not sure about the solutions}
 
@@
 
+\textcolor{violet}{TO DO}
+
 \item Page 4 line 48: should $O_k$ also have a subscript for the variable j? Also $\hat{y}_{ikj}$
+
+\textcolor{violet}{Subscript j for $O_k$ is not needed because for every variable j $O_k$ will be the same and not change depending on j, the same for $\hat{y}_{ik}$.
+In the only case subscript $j$ can be included is in  $\tilde y_{ik}$ to indicate the permuted variable but for simplicity we just keep the formula as it is in the original PPForest paper (https://www.tandfonline.com/doi/abs/10.1080/10618600.2020.1870480) }
 
@@
 
+\textcolor{violet}{TO DO}
+
 \item Figures 1 and 2: I see that the potential range for coefficient values for the projection pursuit seem to be between -1 and 1.
@@
 
+\textcolor{violet}{TO DO}
+
 \item Page 6: If you use a notation for VI for global variable importance (you define it as the average across k trees), I'd just define it here as an equation and use it in the text (e.g in line 19).
+
+\textcolor{violet}{Formulas for all VI were defined and is more clear now.}
 

### 2022-02-03 PPforestVizupdate.Rnw
update some reviewer 4 comments

@@
 
-The global variable importance in a PPforest then can be defined in different ways.
-One way is to average the variable importance from each PPtree across all the trees in the forest.
-Alternatively, it can be computed as a weighted mean of the absolute value of the projection coefficients across all nodes in every tree, for more details ~\cite{da2021projection}.
+The global variable importance in a PPforest then can be defined in different ways. One way is to average the variable importance from each PPtree across all the trees in the forest.
+
+\[
+VI^{(2)}_j=\frac{1}{B}\sum_{k=1}^B \sum_{s= 1}^{S}\frac{|a^{**}_{sjk}|}{G_s }
+\]
+
+\noindent where $a^{**}_{sjk}$ is the projected coefficient for node $s$ and variable $j$ in the $k^{th}$ PPtree, and $S$ is the total number of node partitions in the $k^{th}$ tree.
+Note that $\sum_{s= 1}^{S}\frac{|a^{**}_{sjk}|}{G_s }$ is the importance of the variable $j$ in the PPtree $k$.
+
+Alternatively, it can be computed as a weighted mean of the absolute value of the projection coefficients across all nodes in every tree .
+
+\[VI^{(3)}_j=\frac{1}{B}\frac{1}{G-1}\sum_{k=1}^B\sum_{s = 1}^{S} (1-e_k) I_{sjk}|a^{**}_{sjk}|
+\]
+
+\noindent where $e_k$ is the oob error rate of tree $k$, $I_{sjk}$ is the projection pursuit index value of node $s$ ~\cite{da2021projection}.
 
@@
 
-Multidimensional scaling (MDS) is used to reduce the dimension of this matrix, to view the similarity between observations.
+Multidimensional scaling (MDS) is used to reduce the dimension of this matrix, to view the similarity between observations ~\citep{kruskal1964multidimensional}.
 MDS transforms the dataset into a low-dimensional space where the distances are approximately the same as in the full $n$ dimensions.
@@
 Theoretically, either of these could be substituted or added.
+Parallel coordinate plots are useful to visualize multivariate numerical data.
+As is shown in the first plot in Figure \ref{tab1}, in a PCP each variable is placed equidistant and perpendicular to the x-axis, on the y-axis the values for each variable are standardized to easily compare them.
+Each observation is represented as a line connected across  all the variables.
+In a PCP many variables can be compared together and seeing the relationships between them, identify and characterize groups of observations.
+The features can be ordered so that there are not too many lines intersecting resulting in an unreadable chart \citet{inselberg2009parallel}.
 

### 2022-02-08 response_reviewers.Rmd
woeking on reviewer response

@@
 
+\textcolor{violet}{MDS citation was included.
+In  Section 3 we describe the mapping of diagnostics to visualizations.
+MDS is presented in Subsection 3.3 (Similarity of cases) the key connection with the forest is that we are using the proximity matrix information which is one of the diagnostics described for PPforest or RF and we use this information to do the MDS.
+This visualization can be useful to identify in which cases the model is working well and when is having problems to classify cases.
+MDS and generalized ternary plot (tetrahedral plot in case there are  four clases) are both implemented in the app and we included a comment in the paper to clarify this point. }
+
 \item Page 7 Figure 2: why does the left plot need to be a dot plot rather than, say, a single box or violin plot for each variable?
@@
 
+\textcolor{violet}{We selected a jittered side-by-side dot plot in this case instead of a box plot because we want to see the values for each tree.
+Jittering implys to add random noise to data in order to prevent overplotting.
+The x axis represents each variable but instead of using the same x position for all the points we use jittered values in an interval.
+This plot have more sence when is presented in the app conected with the others in the model level analysis.
+}
+
 \item Page 9, Figure 4: is the dimension of the simplex always at most 3 or G-1?
@@
 I'd clarify whether this only works for your example or whether it can always be drawn.
+
+\textcolor{violet}{In Figure 4 the dimension of the simplex is G-1, since in crab data there are 4 clases in this case we work with a three dimensional simplex.
+The ternary plot can be generalized to more dimensions and in the interactive web app (Section 4) we present a problem with 7 clases (fishcatch data) and in Figure 6 shows how a generalized ternary plot look like with more classes (6 dimensional simplex).
+}
 
@@
 
+\textcolor{violet}{We are using points because it is important to see the individual cases and the jitter is to overcome overploting.
+It is not recommended using jittered scatterplots because you are changing the real values of your data.
+In this case we are not changing the values of a continuous variable just the position to overcome overplotting in a categorical one.
+}
+
 \item Page 10, line 40: This sentence is muddled "As Sievert (2020) describes, one of the biggest difficulties for the app in order manage linking between plots is the data structure management for each widget".
@@
 
+\textcolor{violet}{Need help not sure what is more descriptive for widget}
+
 \item Page 10, line 45: If you are just going to demonstrate the interactivity, wouldn't it be better to just continue with the original crab dataset, or have one dataset throughout?
 Unless there is something particularly interesting about this dataset.
+
+\textcolor{violet}{Yes fishcatch data are more difficutl to classify and then more interesting to show some of the results.
+For simplicity to describe the main plot of the app we use crab data but to describe the app we prefer to present a more complex problem, more clases, umbalanced data...}
 
@@
 
+\textcolor{violet}{TO DO,  I just selected two dimensions in the MDS.
+For the generalized ternary plot we are ploting diferent rotations of a 6-dimensional simplex but I can add similar labels as we did in Figure 7 }
+
 \item Page 13 line 46: I wasn't aware that there was an option in RF to prune under-performing trees (as opposed to nodes within a tree) in an ensemble.
@@
 Also, in this case, maybe a different word than "prune", such as "discard".
+
+\textcolor{violet}{This comment is in page 12, In the original RF there is not an option to prune under-performing trees.
+This comment is just to mention a possible use for the information we get from the app, also we are presenting the app using PPforest and this can be a possible extension for our algorithm.}
 
@@
 Furthermore, there are no titles for either, and they have the same axis labels so I have no idea what the difference is between them.
+
+\textcolor{violet}{TO DO}
+
 \end{enumerate}

### 2022-04-22 PPforestVizupdate.Rnw
woeking on reviewer response
include some modifications in text because of changes in app tab 2

@@
 Section \ref{visen} maps the ensemble components to the visual elements.
-The web app is described in \ref{app} and further work is discussed in section \ref{fur}.
+The web app is described in \ref{app} and further work is discussed in Section \ref{fur}.
 
@@
 
+\newpage
+
 \subsection{Similarity of Cases}
@@
 Some crabs are clearly more similar to a different group, especially in examining the sex differences.
+MDS is included in the interactive web app presented in Section \ref{app}.
 
@@
 
-With G classes, the ternary plot idea is generalized to a $(G-1)- dimensional$ simplex~\citep{sutherland2000orca, schloerke}. This is one of the approaches used to visualize the vote matrix.
+With G classes, the ternary plot idea is generalized to a $(G-1)- dimensional$ simplex~\citep{sutherland2000orca, schloerke}.
+This is one of the approaches used to visualize the vote matrix and it is included in the interactive web app presented in Section \ref{app}.
 
@@
 Figure \ref{tab2} is a screenshot of the models tab.
-There are five plots, with varying levels of interaction: (1) a jittered side-by-side dotplot showing variable importance for the top three nodes of all trees in the forest, (2) a static display of one tree, (3) a boxplot of oob error for all trees, (4) a faceted density plot of projected data at each node of the tree, with split point indicated by a vertical line, and (5) a mosaic plot showing the confusion matrix for each node of the tree.
+There are five plots, with varying levels of interaction: (1) a boxplot of oob error for all trees including the data as jittered points, (2) a jittered side-by-side dotplot showing variable importance for the top three nodes of all trees in the forest, (3) a static display of one PPtree, (4) a faceted density plot of projected data at each node of the tree, with split point indicated by a vertical line, and (5) a mosaic plot showing the confusion matrix for each node of the tree.
 The interaction is driven from the variable importance plot -- when the user selects a point in that display, the corresponding tree density displays and mosaic plots are drawn.
@@
 \caption{The individual model tab in the web app.
-Variable importance is displayed as jittered dot plots for three nodes of all trees.
-This is linked to a display of the PPTree, a boxplot of the error for all trees in the forest, and a display of the data showing splits at each of three nodes and confusion tables as mosaic plots.
-Clicking a point in the jittered dotplot triggers various updates: each of the importance values for the same tree are highlighted (red), the tree that this corresponds to is drawn, the error for the tree is shown on the boxplot (in red), and the data displays are updated to show the tree.
+The oob error is displayed in a boxplot  including jittered points for all trees in the forest.
+Variable importance is displayed as jittered dot plots for selected nodes of all trees.
+The boxplot is linked to a display of the PPTree, the variable importance plot, and a density plot of the data projections showing splits at each selected nodes and confusion tables as mosaic plots.
+Clicking a point in the oob error boxplot triggers various updates: each of the importance values for the same tree are highlighted (red), the tree that this corresponds to is drawn, the error for the tree is shown on the boxplot (in red), and the density plot and the mosaic plot are also updated.
 \label{tab2}}

### 2022-04-22 response_reviewers.Rmd
add some reviewer 2 answers

@@
 
-\textcolor{violet}{We included a description of parallel coordinate plot in section 4 were this figure appears and the corresponding reference for further details.}
+\textcolor{violet}{We have included a description of parallel coordinate plot in Section 4 were this figure appears and the corresponding reference for further details.
+In the vertical axis the values for each variable are standardized to easily compare them.}
 
@@
 
-\textcolor{violet}{TO DO, change selection to boxplot trees}
+\textcolor{violet}{The second tab was modifyed in different ways taking into account your valuable comments.
+We have included three main changes in the second tabset, first we have included jittered points in the OOB error boxplot where the user can select trees based in their performance instead of selecting trees based on importance variable.
+Second, we have included a panel to select the nodes to see in the importance variable plot, the mosaic plot and the density plot.
+The panel allows you to select more or less than three nodes and the mentioned plots are updated acordingly.
+Finally, we rearrange some of the plots to make the selection and interaction more intuitive  presented in Figure 8.}
 
@@
 
-\textcolor{violet}{TO DO, not sure about the solutions}
+\textcolor{violet}{We explore different alternatives as you suggested but the result was not clearly better.
+We have removed the points and just keep thiner lines which improve the visualization.
+Also in the app, because the interactivity, it is easy to see the oob error values mousing over the plot which in the static version it is not possible.}
 
@@
 
-\textcolor{violet}{TO DO}
+\textcolor{violet}{Class color legends were not included in the interactive web app because the interactivity allows you to get the class name when you mouse over each plot.}
 

### 2022-04-22 PPforestVizupdate.Rnw
Figure 1 and Figure 2 update based in reviewer 4 comments

@@
 library(ggmosaic)
-load("ppf.RData")
+set.seed(123)
+#  ppf2022 <-PPforest(data = crab, class= "Type", std = TRUE, size.tr = 1 , m = 500, PPmethod = 'LDA', size.p =.33,
+#   lambda = .1, parallel = FALSE, cores = 2, rule = 1)
+# save(ppf2022, file = "ppf2022.RData")
+load("ppf2022.RData")
 load("rf_fish.RData")
+ ppf<- ppf2022
 @
@@
 <<trcomb, echo=FALSE, warning=FALSE, message=FALSE, fig.height=4, fig.width=6.5, results='hide', error=FALSE>>=
-tree <- 18
+
+tree <- 11
 
@@
 
-  p1 <- dat_pl %>% filter(node.id %in%nodes) %>%
+  p1 <- dat_pl %>% filter(node.id %in%nodes) %>% mutate(node.id = paste("Node",node.id)) %>% 
     ggplot( aes(  x = proj.data, group = Class, fill = Class ) ) +
@@
                                                       linetype = "dashed",
-                                                      color = 2) + labs(x="")
+                                                      color = 2) + labs(x="") +
+    scale_x_continuous( breaks = c(-0.5,0,0.5),
+                    limits=c(-0.5,0.5)) +theme(axis.text.x = element_text(size=8))
 
@@
 
-p1b <- PPtree_dens(ppf, tree, nodes =c(1,2,3), pos="none")
+p1b <- PPtree_dens(ppf, tree, nodes = c(1,2,3), pos="none")
 
@@
 
-  p1 <- dat_mosaic %>% filter(node.id %in%nodes) %>% ggplot() +
+  p1 <- dat_mosaic %>% filter(node.id %in%nodes) %>% mutate(node.id = paste("Node",node.id))%>% ggplot() +
     geom_mosaic( aes(weight = Freq, x = product(Class,Dir) ,fill = Class),  offset=.005)+facet_grid(~node.id)+
@@
 colnames(coefs)  <- c("node", colnames(ppf$train[, -1]))
-coefs <- coefs %>% gather(variable, value,-node)
+coefs <- coefs %>% gather(variable, value,-node) %>% mutate(node = paste("Node",node))
 
@@
 labpl <-PPtree_dens(ppf, tree, nodes =c(1,2,3), pos="bottom")
+ppf[["output.trees"]][[tree]]$origclass <-ppf[["output.trees"]][[tree]]$origclass %>% recode(`1` ="BF", `2`="BM", `3`="OF",`4`="OM" )
 
@@
 
-aux <- bestnode %>% filter(node %in% nodes) %>% mutate(ids = rep(1:ppf$n.tree, each = length(nodes))) %>% gather(var, value, -ids, -node)
+aux <- bestnode %>% filter(node %in% nodes) %>% mutate(ids = rep(1:ppf$n.tree, each = length(nodes))) %>% mutate(node = paste("Node",node))%>% gather(var, value, -ids, -node)
 

### 2022-05-06 response_reviewers.Rmd
update

@@
 The paper and its accompanying app seem to be a good fit for Computational Statistics, as they apply interesting visualization techniques to a tree-based ensemble model that will surely prove useful for practitioners.
-However, for this potential to materialize the authors should put more emphasis on the limitations of their approach.
+
+However, for this potential to materialize the authors should put more emphasis on the limitations of their approach. \textcolor{violet}{TO DO}
+
 In particular, the app focuses on the first three nodes of the trees in question, which would be much less informative in settings requiring deep trees.
+\textcolor{violet}{The app was modified to adress this issue.
+Now the user can select the nodes to see, three or more.
+If there are a lot of nodes will be better to select at most 4 to see better the complete display}
+
 Furthermore, plots such as the parallel coordinate plot would lose their appeal in settings with thousands of variables.
 A brief discussion of circumstances limiting the usability of the app would address these concerns and allow users to easily determine whether the app is suitable for their application.
+
+\textcolor{violet}{TO DO}
 
@@
 Then if we have 4 classes the generalized ternary plot is a tetrahedron as shown in Figure 4 but if we have more than four classes we can also generalize the ternary plot idea.
-In the fishcatch data used in this paper to illustrate the shiny app there are 7 classes and we generalized the ternary plot to a 6-dimensional simplex as is shown in the last panel of  Figure 6}
+In the fishcatch data used in this paper to illustrate the shiny app there are 7 classes and we generalized the ternary plot to a 6-dimensional simplex as is shown in the last panel of  Figure 6 (6- didmensional simplex shown pairwise), using tourr package you can see better the 6 dimensional simplex in different rotations https://vimeo.com/707047875.
+The structure is complexbecausethereare7 vertices,21 eddges and 35 faces.}
 
@@
 
-\textcolor{violet}{TO DO}
+\textcolor{violet}{Random Forest presents two sources of randomness in the tree: random variable selection and random bootstrap samples.
+The random variables selection used to node partition, each split are identified from a random subset of all variables.
+There is not an initial sample of variables for each tree, the sample is for each tree in each  node partition, in each case using all the variables.
+We did not explore formally the sensitivity of the random variable sample size in the importance variable measure.
+We have included an image from the original paper with the boundary that would result from a classification tree fitted using the PPtree algorithm to take into account your last request.
+The plot shows a simple example from simulated data with two features and three classes.
+The partitions generated by PPtree algorithm are oblique to the axis, incorporating the association between the two variables.
+}
 
@@
 \textcolor{violet}{Subscript j for $O_k$ is not needed because for every variable j $O_k$ will be the same and not change depending on j, the same for $\hat{y}_{ik}$.
-In the only case subscript $j$ can be included is in  $\tilde y_{ik}$ to indicate the permuted variable but for simplicity we just keep the formula as it is in the original PPForest paper (https://www.tandfonline.com/doi/abs/10.1080/10618600.2020.1870480) }
+In the only case subscript $j$ can be included is in  $\tilde y_{ik}$ to indicate the permuted variable but for simplicity we just keep the formula as it is in the original PPforest paper (https://www.tandfonline.com/doi/abs/10.1080/10618600.2020.1870480) }
 
@@
 But in the upper right corner, the plots of the variables distributions, for instance for node 2, are not in the standardized range (node 2 has an x-axis that goes up to 5 and 6).
+\textcolor{violet}{As you mentioned the data should be standardized for the coefficients to be interpretable as importances, we fixed the density plot and the x axis are in standardized range now.}
+
 Is the distribution being plotted here the value of the linear combination for each class, rather than a single variable?
+
+\textcolor{violet}{The density plots shows the data projection at each node. }
+
 Is there something going on here with the default x limits for doing KDE? 
+\textcolor{violet}{The x limits were fixed.}
+
 As a side note, you show that the first cut perfectly separates the blue and orange species.
 You may want to consider having the bounds for the KDE be the observed range of the linear combination for each class so that the density plots won't overlap in the plot for node 1, which they seem to do more than they should if the separation is perfect.
+
+\textcolor{violet}{We have changed the example but even in the previous one in the first node partition the separation was not perfect as you mentioned.
+We explore your suggestion (changing the bound for the KDE to observe range) but the result was not better and we keep the original.}
+
 Furthermore, if you have any control over the color scheme (if you don't I would recommend adding this as a variable), it might be a good idea in this paper to have the colors for the blue and orange species be blue and orange, and have differing shading patterns to distinguish male from female, or something similar; perhaps you can choose more similar colors within both species.
 Or, better yet, can you use different plotting symbols, say triangles vs circles to distinguish the species, and shaded vs hollow to distinguish the sex?
+
+\textcolor{violet}{We have complete control over the color palette and the selected palette (Dark2) is a qualitative palette which is suited to represent categorical data and is also colorblind friendly.
+In this case we are not changing the color palette as you suggested  because maybe for the density plot your idea will work but in the other visualizations using different shades to differentiate sexes will not be easy to identify when the geometric element for example are points.
+Also in terms of perception using different colors is better than using shapes to differentiate classes.}
+
 Also, can you include labeling in your plot that the sub-plots are node ID (e.g., put the label "node" above the 3 density plots).
+\textcolor{violet}{We have included in each sub-plots "Node" before the number and also update this in the app}
+
 What do the numbers in the grey boxes at the terminal nodes represent?
 
-\textcolor{violet}{TO DO}
+\textcolor{violet}{The numbers  in the gray boxes are the predicted class, we fixed the plot to show the class names}
 
@@
 
-\textcolor{violet}{Formulas for all VI were defined and is more clear now.}
+\textcolor{violet}{Formulas for all VI were defined in equations and mentionedd in the text. It  is more clear now.}
 
@@
 
-\textcolor{violet}{Need help not sure what is more descriptive for widget}
+\textcolor{violet}{NEED HELP, not sure how to improve this sentence and I do not know what is more descriptive for widget.
+At their core Shiny widgets are mini-applications created using the shinyApp function.}
 
@@
 
-\textcolor{violet}{TO DO,  I just selected two dimensions in the MDS.
-For the generalized ternary plot we are ploting diferent rotations of a 6-dimensional simplex but I can add similar labels as we did in Figure 7 }
+\textcolor{violet}{PDP can be order in different ways, we have modified the PDP and included variable order based on the mean value.
+As you suggested, we have also change the grid lines and just keep the vertical grid in gray, it is more clear now.
+In the MDS we just selected two dimensions but can be modify if it is needed by the problem.
+We have included labels in the generalized ternary plot as we did in Figure 7.}
 
@@
 
-\textcolor{violet}{TO DO}
+\textcolor{violet}{In  the upper LH corner we are using the function `selectizeInput()` from shiny R package which creates a select list that can be used to choose a single or multiple items from a list of values, you do not have to type the class names you just select the names from a list.
+For the bottom two plots (variable importance plots) we have fixed the axis and  include titles in each plot, now is easy to compare them.}
 

### 2022-05-06 PPforestVizupdate.Rnw
update

@@
 The PPF algorithm builds on the projection pursuit tree (PPtree)~\citep{lee2013pptree} algorithm which uses projection pursuit at each tree node to find the best linear combination of variables to separate the classes.
+Figure \ref{bounds} shows the boundary that would result from a classification tree fitted using the PPtree algorithm (Example from \citep{da2021projection}).
+
 The visualization approach is consistent with the framework in \cite{wickham2015visualizing}, and the implementation is built on the newest interactive graphics available in R \citep{RCore}.
@@
 Multiple levels of data are constructed for exploration: observation, model, and ensemble summaries.
-
-% Should I include some references about interpretable machine learning
-
-The article is structured as follows.
-Section \ref{key} describes the ensemble components to be accessed.
-Section \ref{visen} maps the ensemble components to the visual elements.
-The web app is described in \ref{app} and further work is discussed in Section \ref{fur}.
-
-\section{Diagnostics in Forest Classifiers}\label{key}
-
-The diagnostics typically available are:
-
-\begin{itemize} \itemsep 0in
-  \item {\em Out-of-bag error:} For each model in the ensemble, some cases of the original data are not used.
-Predicting the response for these cases gives a better estimate for the error of the model with future data.
-The oob error rate is a measure for each model that is combined in the ensemble and is used to provide the overall error of the ensemble.
-
-  \item {\em Uncertainty measure for each observation:} Across individual (classification) models we can compute the proportion of times that a case is predicted to be each class.
-If a case is always predicted to be the true class, there is no uncertainty about an observation.
-Cases that are predicted to be multiple classes, with similar class probabilities, indicate difficulty to classify observations.
-They may be important by indicating neighborhoods of the data space that would benefit from a more complex model, or more simply, they may be errors in measurements in the data.
-
-  \item {\em Variable importance:} Each model uses samples of variables.
-Thus, the accuracy of the models can be compared when the variable is included or omitted.
-There are several versions of statistics that use this to provide a measure of the variable importance for prediction.
-
-  \item {\em Similarity measure for pairs of observations:} In each model, each pair of observations will be either in the same terminal node or not.
-This is used to compute a proximity matrix.
-Cluster analysis on this matrix can be used to follow up the classification to assess the original labeling.
-It may suggest improvements or mistakes in original labels.
-\end{itemize}
-
-In addition to these overall ensemble statistics, each component model has its own diagnostics, measuring error, variables utilized, and class predictions.
-Visualization will enable the individual models to be examined, relate these to the data and their contribution to the ensemble.
-
-\section{Mapping Ensemble Diagnostics to Visual Components}~\label{visen}
-
-This section describes the mapping of diagnostics to visualizations.
-These are illustrated using the Australian crabs data~\citep{CM74}.
-The data has 200 cases, 5 predictors, and 4 classes (combinations of species and sex, blue male, blue female, orange male, and orange female).
-The predictors are: FL (the size of the frontal lobe length, in mm), RW (rear width, in mm), CL (length of mid-line of the carapace, in mm), CW (maximum width of carapace, in mm), BD (depth of the body; for females, measured after displacement of the abdomen, in mm).
-This is old data, but it provides a good illustration of the visual methods.
-
-\subsection{Individual Models: PPtree}
-
-The PPF is composed of individual projection pursuit trees.
-Figure \ref{pptreeviz1} shows a visual ensemble of plots of a tree model on the crab data.
-There are three nodes for the four class problem.
-The nodes of this tree are based on projections of the data; the coefficients of which form the building block to calculate the variable importance.
-The density plot displays the data projection at each node, and the mosaic plot shows the confusion matrix for the nodes.
-The package \verb#PPtreeViz# provides visual tools to diagnose a PPtree model.
-The PPF builds on these and modifies a little.
-The PPtree model is simpler than a regular classification tree because the classes are mostly separated by combinations of variables -- just three projections are needed to see the differences between the four classes.
 
@@
 
+\begin{figure}[!t]
+<<boundss, cache=FALSE, echo=FALSE, fig.height=5, fig.width=8, message=FALSE>>=
+grid_arrange_shared_legend <- function(..., ncol = length(list(...)), nrow = 1, position = c("bottom", "right")) {
+
+  plots <- list(...)
+  position <- match.arg(position)
+  g <- ggplotGrob(plots[[1]] + theme(legend.position = position))$grobs
+  legend <- g[[which(sapply(g, function(x) x$name) == "guide-box")]]
+  lheight <- sum(legend$height)
+  lwidth <- sum(legend$width)
+  gl <- lapply(plots, function(x) x + theme(legend.position="none"))
+  gl <- c(gl, ncol = ncol, nrow = nrow)
+
+  combined <- switch(position,
+       "bottom" = arrangeGrob(do.call(arrangeGrob, gl),
+                   legend,
+                   ncol = 1,
+                   heights = unit.c(unit(1, "npc") - lheight, lheight)),
+       "right" = arrangeGrob(do.call(arrangeGrob, gl),
+                   legend,
+                   ncol = 2,
+                   widths = unit.c(unit(1, "npc") - lwidth, lwidth)))
+
+  grid.draw(combined)
+}
+
+simu3 <- function(mux1, mux2, muy1, muy2, muz1, muz2,
+                  cor1, cor2, cor3) {
+  bivn <- MASS::mvrnorm(100, mu = c(mux1, mux2),
+                  Sigma = matrix(c(1, cor1, cor1, 1), 2))
+  bivn2 <- MASS::mvrnorm(100, mu = c(muy1, muy2),
+                   Sigma = matrix(c(1, cor2, cor2, 1), 2))
+  bivn3 <- MASS::mvrnorm(100, mu = c(muz1, muz2),
+                   Sigma = matrix(c(1, cor3, cor3, 1), 2))
+
+  d1 <- data.frame(Sim = "sim1", bivn, stringsAsFactors = TRUE)
+  d2 <- data.frame(Sim = "sim2", bivn2, stringsAsFactors = TRUE)
+  d3 <- data.frame(Sim = "sim3", bivn3, stringsAsFactors = TRUE)
+  rbind(d1, d2, d3)
+}
+
+set.seed(666)
+dat.pl2 <- simu3(-1,0.6,0,-0.6, 2,-1,0.95, 0.95, 0.95)
+dat.plnew <- simu3(0,0.6,0,-0.6, 0,-1,-0.1, -0.1, -0.1)
+
+grilla <- expand.grid(X1 = seq(-4,4.8,,100),
+                            X2 = seq(-4.3,3.3,,100))
+
+pptree <- PPtreeViz::PPTreeclass(Sim~., data = dat.pl2, "LDA")
+ppred.sim <- PPtreeViz::PPclassify(pptree, test.data = grilla,
+                                   Rule = 6)
+grilla$ppred<-ppred.sim[[2]]
+
+rpart.crab <- rpart::rpart(Sim ~ X1 + X2, data = dat.pl2)
+rpart.pred <- predict(rpart.crab, newdata = grilla, type="class")
+
+bnds <- data.frame(
+  a = c(pptree$splitCutoff.node[[1]] / pptree$projbest.node[[3]],
+      pptree$splitCutoff.node[[2]] / pptree$projbest.node[[4]]),
+  b = c(-pptree$projbest.node[[1]] / pptree$projbest.node[[3]],
+      -pptree$projbest.node[[2]] / pptree$projbest.node[[4]]))
+p <- ggplot(data = grilla ) +
+       geom_point(aes(x = X1, y = X2,
+                  color = as.factor(ppred),
+                  shape=as.factor(ppred)),
+                  alpha = .20) +
+      scale_colour_brewer(name="Class", type="qual",
+                      palette = "Dark2") +
+
+      scale_shape_discrete(name = 'Class') +
+      theme_bw() +
+      theme(aspect.ratio = 1, legend.position="none")
+
+pl.pp <- p + geom_point(data = dat.pl2,
+                        aes(x = X1, y = X2,
+                            group = Sim,
+                            shape = Sim, color=Sim), size = 3) +
+  scale_y_continuous(expand = c(0,0)) +
+  scale_x_continuous(expand = c(0,0))
+
+pl.pp
+
+@
+ \vspace*{-0.3cm}
+ \caption{Decision boundaries for PPtree algorithm on 2D simulated data.
+The partitions generated by PPtree algorithm are oblique to the axis, incorporating the association between the two variables.\label{bounds}}
+\end{figure}
+
+% Should I include some references about interpretable machine learning
+
+The article is structured as follows.
+Section \ref{key} describes the ensemble components to be accessed.
+Section \ref{visen} maps the ensemble components to the visual elements.
+The web app is described in \ref{app} and further work is discussed in Section \ref{fur}.
+
+\section{Diagnostics in Forest Classifiers}\label{key}
+
+The diagnostics typically available are:
+
+\begin{itemize} \itemsep 0in
+  \item {\em Out-of-bag error:} For each model in the ensemble, some cases of the original data are not used.
+Predicting the response for these cases gives a better estimate for the error of the model with future data.
+The oob error rate is a measure for each model that is combined in the ensemble and is used to provide the overall error of the ensemble.
+
+  \item {\em Uncertainty measure for each observation:} Across individual (classification) models we can compute the proportion of times that a case is predicted to be each class.
+If a case is always predicted to be the true class, there is no uncertainty about an observation.
+Cases that are predicted to be multiple classes, with similar class probabilities, indicate difficulty to classify observations.
+They may be important by indicating neighborhoods of the data space that would benefit from a more complex model, or more simply, they may be errors in measurements in the data.
+
+  \item {\em Variable importance:} Each model uses samples of variables.
+Thus, the accuracy of the models can be compared when the variable is included or omitted.
+There are several versions of statistics that use this to provide a measure of the variable importance for prediction.
+
+  \item {\em Similarity measure for pairs of observations:} In each model, each pair of observations will be either in the same terminal node or not.
+This is used to compute a proximity matrix.
+Cluster analysis on this matrix can be used to follow up the classification to assess the original labeling.
+It may suggest improvements or mistakes in original labels.
+\end{itemize}
+
+In addition to these overall ensemble statistics, each component model has its own diagnostics, measuring error, variables utilized, and class predictions.
+Visualization will enable the individual models to be examined, relate these to the data and their contribution to the ensemble.
+
+\section{Mapping Ensemble Diagnostics to Visual Components}~\label{visen}
+
+This section describes the mapping of diagnostics to visualizations.
+These are illustrated using the Australian crabs data~\citep{CM74}.
+The data has 200 cases, 5 predictors, and 4 classes (combinations of species and sex, blue male, blue female, orange male, and orange female).
+The predictors are: FL (the size of the frontal lobe length, in mm), RW (rear width, in mm), CL (length of mid-line of the carapace, in mm), CW (maximum width of carapace, in mm), BD (depth of the body; for females, measured after displacement of the abdomen, in mm).
+This is old data, but it provides a good illustration of the visual methods.
+
+\subsection{Individual Models: PPtree}
+
+The PPF is composed of individual projection pursuit trees.
+Figure \ref{pptreeviz1} shows a visual ensemble of plots of a tree model on the crab data.
+There are three nodes for the four class problem.
+The nodes of this tree are based on projections of the data; the coefficients of which form the building block to calculate the variable importance.
+The density plot displays the data projection at each node, and the mosaic plot shows the confusion matrix for the nodes.
+The package \verb#PPtreeViz# provides visual tools to diagnose a PPtree model.
+The PPF builds on these and modifies a little.
+The PPtree model is simpler than a regular classification tree because the classes are mostly separated by combinations of variables -- just three projections are needed to see the differences between the four classes.
+
 \begin{figure}[hbpt]
@@
 Importance variable is computed based on the oob cases for the tree $k\;\;(O_k)$ for each $j$ predictor variable.
-Permuted importance of the variable $j$ in the tree $k$ is defined as:
-
-\[
+Permuted importance of the variable $j$ in the tree $k$ is defined in Equation \ref{vi1}.
+
+\begin{equation}
 VI^{(1)} _{jk} = \frac{\sum_{i \in O_k } I(y_i=\hat y_{ik})-I(y_i=\tilde y_{ik})}{\# O_k}
-\]
+\label{vi1}
+\end{equation}
 
@@
 
-The global variable importance in a PPforest then can be defined in different ways. One way is to average the variable importance from each PPtree across all the trees in the forest.
-
-\[
+The global variable importance in a PPforest then can be defined in different ways.
+One way is to average the variable importance from each PPtree across all the trees in the forest (Equation \ref{vi2}).
+
+\begin{equation}
 VI^{(2)}_j=\frac{1}{B}\sum_{k=1}^B \sum_{s= 1}^{S}\frac{|a^{**}_{sjk}|}{G_s }
-\]
+\label{vi2}
+\end{equation}
 
@@
 
-Alternatively, it can be computed as a weighted mean of the absolute value of the projection coefficients across all nodes in every tree .
-
-\[VI^{(3)}_j=\frac{1}{B}\frac{1}{G-1}\sum_{k=1}^B\sum_{s = 1}^{S} (1-e_k) I_{sjk}|a^{**}_{sjk}|
-\]
+Alternatively, it can be computed as a weighted mean of the absolute value of the projection coefficients across all nodes in every tree as shown in Equation \ref{vi3}.
+
+\begin{equation}
+VI^{(3)}_j=\frac{1}{B}\frac{1}{G-1}\sum_{k=1}^B\sum_{s = 1}^{S} (1-e_k) I_{sjk}|a^{**}_{sjk}|
+\label{vi3}
+\end{equation}
 
@@
 
+#vimeoternary fish code
+# library(colorspace)
+# library(RColorBrewer)
+# library(tourr)
+# projct <- t(f.helmert(7)[-1,] )
+# 
+# dat7 <- data.frame(class=ppf[[1]], ppf$votes%*%projct)
+# colnames(dat7)[-1] <- paste("x", 1:7,sep="")
+# pal <- brewer.pal(7,"Dark2") 
+# col <- pal[as.numeric(dat7[, 1])]
+# 
+# edges <- rbind(c(1,0,0,0,0,0,0), c(0,1,0,0,0,0,0), c(0,0,1,0,0,0,0), c(0,0,0,1,0,0,0), 
+#                c(0,0,0,0,1,0,0), c(0,0,0,0,0,1,0),c(0,0,0,0,0,0,1) )%*%projct
+# colnames(edges) <- colnames(dat7)[-1]
+# dd7 <- rbind(dat7, data.frame(class=levels(dat7$class),edges ))
+# 
+# x11()
+# animate_xy(dd7[,2:7], col = col, pch=19,axes ="off", edges=as.matrix(expand.grid(160:166,160:166) )[c(2:7,10:14,18:21, 26:28,34:35,42),])
+#
+
 @

### 2022-05-09 PPforestVizupdate.Rnw
include LDA and PDA index

@@
 
+\newcommand{\blind}{0}
+
+\newcommand{\blA}{\mbox{\boldmath {\bf A}}}
+\newcommand{\blB}{\mbox{\boldmath {\bf B}}}
+\newcommand{\blW}{\mbox{\boldmath {\bf W}}}
+\newcommand{\blX}{\mbox{\boldmath {\bf X}}}
+\newcommand{\blM}{\mbox{\boldmath {\bf M}}}
+\newcommand{\blx}{\mbox{\boldmath {\bf x}}}
+\newcommand{\bla}{\mbox{\boldmath {\bf a}}}
+\newcommand{\blalpha}{\mbox{\boldmath {\alpha}}}
+
 \begin{document}
@@
 
-The PPF is composed of individual projection pursuit trees.
-Figure \ref{pptreeviz1} shows a visual ensemble of plots of a tree model on the crab data.
+The PPF is composed of individual projection pursuit trees, PPtree algorithm uses a multi-step approach to fit a multi-class model by finding linear combinations to split on based on prjection pursuit algorithm.
+Two projection pursuit indexes are used to find projections that separate classes,  LDA  \citep{lee2005projection} and PDA \citep{lee2010projection}.
+
+The LDA index is defined as follows:
+
+\begin{equation}
+ \mathbb{I}_{LDA}(\blA) = \left\{
+  \begin{array}{l l}
+    1-\frac{|\blA^T \blW\blA|}{|\blA^T(\blW+\blB)\blA|} &  \text{for}~|\blA^T(\blW+\blB)\blA|\neq 0\\
+    0 &  \text{for}~|\blA^T(\blW+\blB)\blA|= 0
+  \end{array} \right.
+\end{equation}
+
+\noindent where $p\times q$ matrix $\blA = [\bla_1, \bla_2 , .
+.
+.
+, \bla_q ]$ defines an orthonormal projection from $p$-dimensions onto a $q$-dimensional subspace, $\blB=\sum_{g=1}^G n_g(\bar{\mathbf{x}}_{g}-\bar{\mathbf{x}})(\bar{\mathbf{x}}_{g}-\bar{\mathbf{x}})^{T}$ is the $p\times p$ between-group sum of squares,
+$\blW=\sum_{g=1}^{G}\sum_{i\in H_g}(\mathbf{x}_{i}-\bar{\mathbf{x}}_{g})(\mathbf{x}_{i}-\bar{\mathbf{x}}_{g})^T$ is the $p\times p$  within-group sum of squares where $H_g=\{i| y_i = g, i = 1, ..., n\}$, $\bar{\mathbf{x}}_{g}$ is the group mean vector and $\bar{\mathbf{x}}$ the overall mean vector.
+If the LDA index value is high, there is a large difference between classes.
+
+The PDA index is useful when $n$ is small relative to $p$ and when the variables are highly correlated.
+In these situations the maximum likelihood variance-covariance matrix estimator will be close to singular, affecting the inverse calculation.
+The PDA index adjusts the variance-covariance matrix calculation, as follows:
+
+\begin{equation}
+\mathbb{I}_{PDA}(\blA,\lambda)=1-\frac{|\blA^T \blW_{PDA}(\lambda)\blA|}{|\blA^T (\blW_{PDA}(\lambda)+\blB) \blA|}
+\end{equation}
+
+\noindent where notation is the same as for the LDA index, with the addition of $\lambda \in [0,1)$ is a shrinkage parameter, and a different within group sum of squares, $\blW_{PDA}(\lambda)=\mbox{diag}(\blW)+(1-\lambda)\mbox{offdiag}(\blW)$.
+
+Figure \ref{pptreeviz1} shows a visual ensemble of plots of a tree model on the crab data using LDA index.
 There are three nodes for the four class problem.
@@
 
+\begin{figure}[hbpt]
+\centering
+\includegraphics[width=1\linewidth]{figure/fish11comp.png}
+\caption{Entry page for the web app, focusing on examining cases, for selected points. \label{tab1comp}}
+\end{figure}
+
 The diagram in Figure \ref{tab1diag} illustrates the data pipeline \citep{BAHM88,wickham2009plumbing} for the interactive graphics in the case level tab.

### 2022-05-09 response_reviewers.Rmd
new reviewer comments

@@
 
-However, for this potential to materialize the authors should put more emphasis on the limitations of their approach. \textcolor{violet}{TO DO}
-
+However, for this potential to materialize the authors should put more emphasis on the limitations of their approach.
 In particular, the app focuses on the first three nodes of the trees in question, which would be much less informative in settings requiring deep trees.
 \textcolor{violet}{The app was modified to adress this issue.
-Now the user can select the nodes to see, three or more.
-If there are a lot of nodes will be better to select at most 4 to see better the complete display}
+Now the user can select specific nodes to see (three or more nodes).
+If there are a lot of nodes will be better to select 4 or less each time to see better the complete display.}
 
@@
 Then if we have 4 classes the generalized ternary plot is a tetrahedron as shown in Figure 4 but if we have more than four classes we can also generalize the ternary plot idea.
-In the fishcatch data used in this paper to illustrate the shiny app there are 7 classes and we generalized the ternary plot to a 6-dimensional simplex as is shown in the last panel of  Figure 6 (6- didmensional simplex shown pairwise), using tourr package you can see better the 6 dimensional simplex in different rotations https://vimeo.com/707047875.
-The structure is complexbecausethereare7 vertices,21 eddges and 35 faces.}
+In the fishcatch data used in this paper to illustrate the shiny app there are 7 classes and we generalized the ternary plot to a 6-dimensional simplex as is shown in the last panel of Figure 6 (6-didmensional simplex shown pairwise), using tourr package you can see better the 6-dimensional simplex in different rotations https://vimeo.com/707047875.
+The structure is complex because a 6-dimensional simplex has 7 vertices, 21 eddges and 35 faces.}
 
@@
 
-\textcolor{violet}{TO DO}
+\textcolor{violet}{Yes, the coefficient values are between -1 and 1 by construction.
+We find the linear combinations, \textbf{a} which maximize $\frac{a^TBa}{a^TWa}$, where B y W are the between and within group sums of squares.
+We get $\hat a_1=e_1$ where $e_1...e_s$ are eigenvectors of $W^{-1}B$.
+In this case we find the 1-D projection optimizing the LDA index.
+We have included in Section 3 LDA and PDA indexes which are optimized to get the linear projections and it is more clear now how the coefficients are computed.
+}
 
@@
 \textcolor{violet}{In Figure 4 the dimension of the simplex is G-1, since in crab data there are 4 clases in this case we work with a three dimensional simplex.
-The ternary plot can be generalized to more dimensions and in the interactive web app (Section 4) we present a problem with 7 clases (fishcatch data) and in Figure 6 shows how a generalized ternary plot look like with more classes (6 dimensional simplex).
-}
+Ternary plot can be generalized to more than three dimensional space, with G classes, the ternary plot idea is generalized to a G-1 dimensional simplex.
+Then if we have 4 classes the generalized ternary plot is a tetrahedron as shown in Figure 4 but if we have more than four classes we can also generalize the ternary plot idea.
+In the fishcatch data used in this paper to illustrate the shiny app there are 7 classes and we generalized the ternary plot to a 6-dimensional simplex as it is shown in the last panel of Figure 6 (6-didmensional simplex shown pairwise), using tourr package you can see better the 6-dimensional simplex in different rotations https://vimeo.com/707047875.
+The structure is complex because a 6-dimensional simplex has 7 vertices, 21 eddges and 35 faces.}
 

### 2022-05-19 PPforestVizupdate.Rnw
small changes

@@
 Various aspects are explored in this article, to assess model complexity, individual model contributions, variable importance and dimension reduction, and uncertainty in prediction associated with individual observations.
-The ideas are applied to the random forest algorithm and projection pursuit forest but could be more broadly applied to other bagged ensembles.
+The ideas are applied to the random forest algorithm and projection pursuit forest but could be more broadly applied to other bagged ensembles helping in the interpretability deficit of these methods.
 Interactive graphics are built in R using the \texttt{ggplot2}, \texttt{plotly}, and \texttt{shiny} packages.
@@
 The random forest (RF) algorithm~\citep{breiman1996bagging} was one of the first ensemble classifiers developed.
-It combines the predictions from individual classification and regression trees (CART)~\citep{breiman1984cl} built by bagging observations~\citep{breiman1996bagging}.
-It also samples variables at each tree node.
+It combines the predictions from individual classification and regression trees (CART)~\citep{breiman1984cl} built by bagging observations~\citep{breiman1996bagging} and random predictor sample at each tree node.
 These produce diagnostics in the form of uncertainty in predictions for each observation, importance of variables for the prediction, predictive error for future samples based on out-of-bag (oob) case predictions, and similarity of observations based on how often they group together in the trees.
@@
 Multiple levels of data are constructed for exploration: observation, model, and ensemble summaries.
+The visualization tools presented in this article help open up black-box models and contribute to the interpretability of these methods.
 
@@
 
-% Should I include some references about interpretable machine learning
-
 The article is structured as follows.
@@
 
+\newpage
+
 \section{Diagnostics in Forest Classifiers}\label{key}
@@
 \begin{itemize} \itemsep 0in
-  \item {\em Out-of-bag error:} For each model in the ensemble, some cases of the original data are not used.
+\item {\em \textbf{Out-of-bag error:}} For each model in the ensemble, some cases of the original data are not used.
 Predicting the response for these cases gives a better estimate for the error of the model with future data.
-The oob error rate is a measure for each model that is combined in the ensemble and is used to provide the overall error of the ensemble.
-
-  \item {\em Uncertainty measure for each observation:} Across individual (classification) models we can compute the proportion of times that a case is predicted to be each class.
+The oob error rate is a measure for each model that is combined in the ensemble and is used to provide the overall error of the ensemble.\\
+
+\item {\em \textbf{Uncertainty measure for each observation:}} Across individual (classification) models we can compute the proportion of times that a case is predicted to be each class.
 If a case is always predicted to be the true class, there is no uncertainty about an observation.
 Cases that are predicted to be multiple classes, with similar class probabilities, indicate difficulty to classify observations.
-They may be important by indicating neighborhoods of the data space that would benefit from a more complex model, or more simply, they may be errors in measurements in the data.
-
-  \item {\em Variable importance:} Each model uses samples of variables.
+They may be important by indicating neighborhoods of the data space that would benefit from a more complex model, or more simply, they may be errors in measurements in the data.\\
+
+\item {\em \textbf{Variable importance:}} Each model uses samples of variables.
 Thus, the accuracy of the models can be compared when the variable is included or omitted.
-There are several versions of statistics that use this to provide a measure of the variable importance for prediction.
-
-  \item {\em Similarity measure for pairs of observations:} In each model, each pair of observations will be either in the same terminal node or not.
+There are several versions of statistics that use this to provide a measure of the variable importance for prediction.\\
+
+\item {\em \textbf{Similarity measure for pairs of observations:}} In each model, each pair of observations will be either in the same terminal node or not.
 This is used to compute a proximity matrix.
@@
 \end{itemize}
-
 In addition to these overall ensemble statistics, each component model has its own diagnostics, measuring error, variables utilized, and class predictions.
@@
 
-\section{Mapping Ensemble Diagnostics to Visual Components}~\label{visen}
+\section{Mapping Ensemble Diagnostics to Visual Components}
+\label{visen}
 
@@
 
-The PPF is composed of individual projection pursuit trees, PPtree algorithm uses a multi-step approach to fit a multi-class model by finding linear combinations to split on based on prjection pursuit algorithm.
+The PPF is composed of individual projection pursuit trees, PPtree algorithm uses a multi-step approach to fit a multi-class model by finding linear combinations to split on based on projection pursuit algorithm.
 Two projection pursuit indexes are used to find projections that separate classes,  LDA  \citep{lee2005projection} and PDA \citep{lee2010projection}.
@@
 \begin{equation}
+\label{lda}
  \mathbb{I}_{LDA}(\blA) = \left\{
@@
 
-Figure \ref{pptreeviz1} shows a visual ensemble of plots of a tree model on the crab data using LDA index.
+Figure \ref{pptreeviz1} shows a visual ensemble of plots of a tree model on the crab data using LDA index(Equation \ref{lda}) and two variables in each node split.
 There are three nodes for the four class problem.
@@
 
-For the crab data, $G=4$ and the generalized ternary plot will be a tetrahedron. The  \verb#tourr# package~\citep{wickham2011tourr} can be used to view it (e.g. \url{https://vimeo.com/170522736}).
-
+For the crab data, $G=4$ and the generalized ternary plot will be a tetrahedron.
 Figure \ref{tetra} shows the tetrahedron structure for the crab vote matrix shown in three pairwise views.
@@
 This is close but not perfect, indicating some crabs are commonly incorrectly classified.
+The \verb#tourr# package~\citep{wickham2011tourr}, tour methods for multivariate data visualization, can be used to view the animated version (e.g.
+\url{https://vimeo.com/170522736}).
 

### 2022-06-13 response_reviewers.Rmd
updated figures and commented diagrams

@@
 We did not explore formally the sensitivity of the random variable sample size in the importance variable measure.
+In the algorithm you can select the proportion of variables used in each node partition.
+In the examples presented in the paper for crab and fishcatch data we use 2 variables in each node partition.
 We have included an image from the original paper with the boundary that would result from a classification tree fitted using the PPtree algorithm to take into account your last request.
@@
 
-\textcolor{violet}{Yes fishcatch data are more difficutl to classify and then more interesting to show some of the results.
+\textcolor{violet}{Yes fishcatch data are more difficult to classify and then more interesting to show some of the results.
 For simplicity to describe the main plot of the app we use crab data but to describe the app we prefer to present a more complex problem, more clases, umbalanced data...}

### 2022-06-13 PPforestVizupdate.Rnw
update
updated figures and commented diagrams

@@
 \end{figure}
+Two data set are used in this paper, crab ~\citep{CM74} and fishcatch data~\citep{fishcatch}.
+To illustrate the mapping of diagnostics to visualization crab data are used because its simplicity.
+The second data set is introduced to illustrate the interactive web app which combines the visualizations introduced in this paper and present a more complex problem to explore within the applicarion.
 
@@
 The web app is described in \ref{app} and further work is discussed in Section \ref{fur}.
-
-\newpage
 
@@
 
-Figure \ref{pptreeviz1} shows a visual ensemble of plots of a tree model on the crab data using LDA index(Equation \ref{lda}) and two variables in each node split.
+Figure \ref{pptreeviz1} shows a visual ensemble of plots of a tree model on the crab data using LDA index (Equation \ref{lda}) and two variables in each node split.
 There are three nodes for the four class problem.
@@
 The dashed vertical red line indicates the split value of each node.
-At node 1 the blue species is separated from orange species.
-Nodes 2 and 3 separate the sexes, which are more confused for the blue species.
+At node 1 separates the sexes.
+Nodes 2 separate males from species and node 3 separate females from species.
 Mosaic plots of the confusion table for each split (bottom).
-Node 1 shows the clear split of the species, with a small number of misclassifications.
-Node 2 where orange females are separated from orange males indicates small number of misclassifications.
-Node 3 where blue females are separated from blue males, shows a larger misclassification than for orange species.\label{pptreeviz1}}
+Node 1 shows the clear split of sexes, with a small number of misclassifications.
+Node 2 where blue males are separated from orange males with small number of misclassifications.
+Node 3 where blue females are separated from orange females, almost perfect classification.\label{pptreeviz1}}
 \end{figure}
@@
 The forest was built using random samples of two variables for each node, hence there are two coefficients for each node.
-At node 1, BD has a high value and CW contributes much less.
+At node 1 uses CW and RW with similar contribution.
 The scatterplot at right shows these two variables and the resulting boundary between groups that this would produce.
-Node 2 uses CL and RW, and RW contributes the most to the separation.
+Node 2 uses BD and CL, BD contributes the most to the separation.
 The plot in the right shows the boundary that is induced.
-Node 3 uses FL and RW, and this is a much more even contribution by the two variables.
+Node 3 uses CW and FL, and with an even contribution by the two variables.
 For each tree in the forest, different decision rules are defined; the resulting boundaries on the previous plots are based on Rule 1 $= \frac{m_1}{2} + \frac{m_2}{2} $, where $m_1$ and $m_2$ are the mean of the left and right groups at each node.
@@
 This is close but not perfect, indicating some crabs are commonly incorrectly classified.
-The \verb#tourr# package~\citep{wickham2011tourr}, tour methods for multivariate data visualization, can be used to view the animated version (e.g.
-\url{https://vimeo.com/170522736}).
+The \verb#tourr# package~\citep{wickham2011tourr}, tour methods for multivariate data visualization, can be used to view the animated version available for this example in  \url{https://vimeo.com/170522736}.
 
@@
 \centering
-<<side, echo=FALSE, warning=FALSE, message=FALSE, fig.height=4,fig.width=4>>=
+<<side, echo=FALSE, warning=FALSE, message=FALSE, fig.height=5,fig.width=5>>=
 
 #side-by siede boxplot
-side <-  function(ppf, ang = 0, lege = "bottom", siz = 3,
+side <-  function(ppf, ang = 0, lege = "bottom", siz = 2.5,
                   ttl = "Side by side dotplot") {
@@
     scale_colour_brewer(type = "qual", palette = "Dark2") +
-    theme(legend.position = lege, legend.text=element_text(angle = ang)) +
+    theme(legend.position = lege, legend.text=element_text(angle = ang, size =8),
+legend.title = element_text(size=8)) +
     labs(colour = "Class", y = "Proportion")
@@
 
- Interaction is added to the plots described in Section \ref{visen} and other plots, and they are organized into an interactive web app using \verb#shiny#~\citep{chang11shiny} for exploring the ensemble model.
+ Interaction is added to the plots described in Section \ref{visen} and other plots, and they are organized into an interactive web app using \verb#shiny#~\citep{chang11shiny, wickham2021mastering} for exploring the ensemble model.
 The app is organized into three tabs: individual cases; models; and performance comparison, to provide a model diagnostic tool.
@@
 
-The fishcatch data~\citep{fishcatch} is used to illustrate the shiny app characteristics.
-It contains 159 observations, with 6 physical measurement variables, and  7 types of fish, all caught from the same lake (Laengelmavesi) near Tampere in Finland.
+To illustrate the shiny app characteristics a different data (fishcatch) are used which present a more challenging classification problem and then more interesting results to explore the app.
+The fishcatch data~\citep{fishcatch} contains 159 observations, with 6 physical measurement variables, and  7 types of fish, all caught from the same lake (Laengelmavesi) near Tampere in Finland.
 There are 35 bream, 11 parkki, 56 perch, 17 pike, 20 roach, 14 smelt and 6 whitewish.
@@
 The data feeding the display is an $n\times p$ data frame; containing the original data; the model statistics generated from the full $n\times G$ vote matrix along with its generalized ternary coordinates; and the first two MDS projections of the proximity matrix.
-Figure \ref{tab1} shows the arrangement of plots.
+
 The plots in the tab are (1) a parallel coordinate plot (PCP) of the data, (2) the MDS display of the proximity matrix, (3) side-by-side jittered dotplot, and (4) generalized ternary plot of the vote matrix.
 Each of these plots are interactive in the sense that each one presents individual interactions (mouse-over) and they are linked so that selections in one display are propagated to other plots (clicking and selecting).
+
+The diagram in Figure \ref{tab1diag} illustrates the data pipeline \citep{BAHM88,wickham2009plumbing} for the interactive graphics in the case level tab. 
+\begin{figure}[hbpt]
+\centering
+\includegraphics[width=1\linewidth]{figure/fishT1.pdf}
+\caption{Schematic diagram illustrating the interactivity in and between plots for individual level exploration panel of the web app.
+From the data, the four plots are generated.
+Each plot has a listener attached which collects user actions.
+When a user selects a point or line in any of the displays, it makes a change in the data which propagates updates to each of the plots.}
+\label{tab1diag}
+\end{figure}
+Solid lines indicate notifications from the source data to the plots, and dashed lines indicate notification of user action on the plot, which notifies the data source of actions to take.
+The data table is a reactive object that has a listener associated with it.
+Each of the plots is reactive and has numerous listeners.
+When users make selections on a plot, either by clicking or group selection, a change to the data is made in terms of an update on the selected cases.
+This invokes a note to other plots to re-draw themselves.
+The linking between plots is effectively one-to-one, based on the row ID of the data.
+The side-by-side jittered dotplot has $n\times G$ points, but selection can only be done within a dotplot.
+Selecting in one of the dotplots notifies the data table of the selection which triggers a re-draw of the other dotplots.
+Mouseovers on the plot pull additional information about the point or line under the cursor but do not link between plots.
+
+Figure \ref{tab1} shows the arrangement of plots in this first tab (the case level tab).
 
@@
 The features can be ordered so that there are not too many lines intersecting resulting in an unreadable chart \citet{inselberg2009parallel}.
+Two alternatives can be selected in shiny to draw the parallel coordinate plot: parallel or enhanced.
+Parallel draws the classic PCP and enhanced draws a modified version where variables are repeated \citep{hurley2011eulerian}.
+Because reading a PCP is really only possible for neighboring variables, the variables are repeated so that all variable pairs are shown next to each other once.
 
@@
 
+Figure \ref{tab1comp} shows the arrangement of plots in this first tab for selected points to illustrate how the tab works.
+In the MDS plot, the yellow points (smelt) which are grouped were selected, in the parallel plot it is clear all these observations are similar with small values in all the variables and small variability.
+These are easy points to classify for the model, in the side-by-side jittered dot plot all these points have a   probability of being classified as the correct class (smelt) close to one, and in the ternary plot, all these points are concentrated in the vertices.
+
 \begin{figure}[hbpt]
 \centering
-\includegraphics[width=1\linewidth]{figure/fish11comp.png}
+\includegraphics[width=1\linewidth]{figure/tab1com2.png}
 \caption{Entry page for the web app, focusing on examining cases, for selected points.  \label{tab1comp}}
@@
 
-The diagram in Figure \ref{tab1diag} illustrates the data pipeline \citep{BAHM88,wickham2009plumbing} for the interactive graphics in the case level tab.
-Solid lines indicate notifications from the source data to the plots, and dashed lines indicate notification of user action on the plot, which notifies the data source of actions to take.
-The data table is a reactive object that has a listener associated with it.
-Each of the plots is reactive and has numerous listeners.
-When users make selections on a plot, either by clicking or group selection, a change to the data is made in terms of an update on the selected cases.
-This invokes a note to other plots to re-draw themselves.
-The linking between plots is effectively one-to-one, based on the row ID of the data.
-The side-by-side jittered dotplot has $n\times G$ points, but selection can only be done within a dotplot.
-Selecting in one of the dotplots notifies the data table of the selection which triggers a re-draw of the other dotplots.
-Mouseovers on the plot pull additional information about the point or line under the cursor but do not link between plots.
-
-Two alternatives can be selected in shiny to draw the parallel coordinate plot: parallel or enhanced.
-Parallel draws the classic PCP and enhanced draws a modified version where variables are repeated \citep{hurley2011eulerian}.
-Because reading a PCP is really only possible for neighboring variables, the variables are repeated so that all variable pairs are shown next to each other once.
-
-\begin{figure}[hbpt]
-\centering
-\includegraphics[width=1\linewidth]{figure/fishT1.pdf}
-\caption{Schematic diagram illustrating the interactivity in and between plots for individual level exploration panel of the web app.
-From the data, the four plots are generated.
-Each plot has a listener attached which collects user actions.
-When a user selects a point or line in any of the displays, it makes a change in the data which propagates updates to each of the plots.}
-\label{tab1diag}
-\end{figure}
+\newpage
 
@@
 There are five plots, with varying levels of interaction: (1) a boxplot of oob error for all trees including the data as jittered points, (2) a jittered side-by-side dotplot showing variable importance for the top three nodes of all trees in the forest, (3) a static display of one PPtree, (4) a faceted density plot of projected data at each node of the tree, with split point indicated by a vertical line, and (5) a mosaic plot showing the confusion matrix for each node of the tree.
-The interaction is driven from the variable importance plot -- when the user selects a point in that display, the corresponding tree density displays and mosaic plots are drawn.
+The interaction is driven in two ways, from the boxplot plot -- when the user selects a point in that display, the corresponding importance variable plot, tree plot from the \verb#PPtreeViz#, tree density displays and mosaic plots are drawn.
 The tree plot from the \verb#PPtreeViz# is used to visualize the selected tree structure.
-Also highlighted are the variable importance values for each variable for each of the top three nodes and the oob error value for the tree on the boxplot.
+Also highlighted are the variable importance values for each variable for each of the top three nodes by default and the oob error value for the tree on the boxplot.
+Also it is possible other type of interaction, after selecting a tree, the user can select the nodes to show in the importance variable plot, density plot and mosaic plot (right corner).
 
@@
 \centering
-\includegraphics[width=1\linewidth]{figure/fish21.png}
+\includegraphics[width=1\linewidth]{figure/tab2com.png}
 \caption{The individual model tab in the web app.

### 2022-06-16 PPforestVizupdate.Rnw
update plot descriptions

@@
 There are five plots, with varying levels of interaction: (1) a boxplot of oob error for all trees including the data as jittered points, (2) a jittered side-by-side dotplot showing variable importance for the top three nodes of all trees in the forest, (3) a static display of one PPtree, (4) a faceted density plot of projected data at each node of the tree, with split point indicated by a vertical line, and (5) a mosaic plot showing the confusion matrix for each node of the tree.
-The interaction is driven in two ways, from the boxplot plot -- when the user selects a point in that display, the corresponding importance variable plot, tree plot from the \verb#PPtreeViz#, tree density displays and mosaic plots are drawn.
+The interaction is driven in three ways, from the boxplot -- when the user selects a point in that display, the corresponding importance variable plot, tree plot from the \verb#PPtreeViz#, tree density displays and mosaic plots are drawn.
 The tree plot from the \verb#PPtreeViz# is used to visualize the selected tree structure.
-Also highlighted are the variable importance values for each variable for each of the top three nodes by default and the oob error value for the tree on the boxplot.
-Also it is possible other type of interaction, after selecting a tree, the user can select the nodes to show in the importance variable plot, density plot and mosaic plot (right corner).
+Also the variable importance values are highlighted for each variable for each of the top three nodes, by default, and the oob error value for the selected tree on the boxplot (both in red).
+Additionally it is possible to select specific nodes and update the importance variable plot, density plot and mosaic plot (nodes can be selected on the right corner on this second tab) and finally the interaction can be driven from the variable importance plot.
+
+The selected tree in Figure \ref{tab2} has very small oob error shown as red point in the boxplot.
+Two variables were use in each node partition, the first four nodes were selected and display in the importance variable plot, density plot and mosaic plot.
+In the importance variable plot `height` and `lenght1` were used in the first node partition being `height` the important variable to separate Bream and Parkki (Green and orange in density and mosaic plots) from the rest.
+In the second node `length1` and `length2` were both evenly important  to separate Bream from Parkki.
+A similar analysis can be done for other nodes.
 
@@
 \centering
-\includegraphics[width=1\linewidth]{figure/tab2com.png}
+\includegraphics[width=1\linewidth]{figure/tab2comp.png}
 \caption{The individual model tab in the web app.
@@
 Clicking a point in the oob error boxplot triggers various updates: each of the importance values for the same tree are highlighted (red), the tree that this corresponds to is drawn, the error for the tree is shown on the boxplot (in red), and the density plot and the mosaic plot are also updated.
+Also specific nodes can be selected to update the plots.
 \label{tab2}}
@@
 The data source is a \verb#PPforest# object.
-Interaction is driven by the variable importance plot.
+Interaction is driven in different ways, by the boxplot including jittered points.
 Selecting a point triggers a change in the data which cascades to re-draws of the other displays.
 Each plot has some information available on mouse over.
+Also selecting points in the variable importance triggers a change in the data and re-draws the displays corresponding to a specific tree.
+Finally selecting nodes update the data and re-draws the importance variable plot, density plot and mosaic plot.
 
@@
 \caption{Schematic diagram illustrating the interactivity in and between plots for model level exploration panel of the web app.
-Only the dotplot of variable importance has click selection, which invokes changes of the tree display, boxplot, density plots, and mosaic plots.
-Selecting a point makes a change in the data, which propagates the importance values for other variables in this tree to be highlighted (red), draws the tree, highlights the error value of the tree, and shows the projections and confusion matrix for the three top nodes.}
+The boxplot and the dotplot of variable importance has click selection, which invokes changes of the corresponding displays.
+Selecting a point in the boxplot update the node tree information (Shiny Gui) and update all the plots for the three first nodes, also nodes can be selected and update the corresponding plots.
+Finally selecting a point in the importance variable plot makes a change in the data, which propagates the importance values for other variables in this tree to be highlighted (red), draws the tree, highlights the error value of the tree, and shows the projections and confusion matrix for the three top nodes.}
 \label{tab2diag}
@@
 
+The application have some limitations that are need to pointed out, if the number of observation is big the overploting will be an issues in some of the selectedd visualizations and if the number of variables increases, plots like parallel plot and importance variable plot should include the option to select some variables each time or a different approach.
+
 There are many possible extensions to the app, that could help it to be a tool for model refinement: (1) Using the diagnostics to weed out under-performing models in the ensemble; (2) Identifying and boosting models that perform well, particularly if they do well for problematic subsets of the data; (3) Problematic cases could be removed, and ensembles re-fit; (4) Classes as a whole could be aggregated or re-organised as suggested by the model diagnostics, to produce a more effective hierarchical approach to the multiple class problem.

### 2022-06-28 PPforestVizupdate.Rnw
Fix some typos

@@
 The process of bagging and combining results from multiple trees produces numerous diagnostics which, with interactive graphics, can provide a lot of insight into class structure in high dimensions.
-Various aspects are explored in this article, to assess model complexity, individual model contributions, variable importance and dimension reduction, and uncertainty in prediction associated with individual observations.
+Various aspects of models are explored in this article, to assess model complexity, individual model contributions, variable importance and dimension reduction, and uncertainty in prediction associated with individual observations.
 The ideas are applied to the random forest algorithm and projection pursuit forest but could be more broadly applied to other bagged ensembles helping in the interpretability deficit of these methods.
@@
 
-Ensemble classifiers have grown in popularity~\citep{dietterish00}~\citep{talbot09}, and the basic ideas behind the random forest can be applied to virtually any type of model.
+Ensemble classifiers have grown in popularity~\citep{dietterish00, talbot09}, and the basic ideas behind the random forest can be applied to virtually any type of model.
 The benefits for classification are reduced variability in predictive error, and the suite of diagnostics provides the potential for better understanding the class structure in the high-dimensional data space.
@@
 
-There has been some, but not a lot of, research on visualizing classification models.
+There has been some, but not a abundant, research on visualizing classification models.
 \cite{urbanek2002exploring} presents interactive tree visualization implemented in the java software called \verb# KLIMT# that includes zooming, selection, multiple views, interactive pruning, and tree construction as well as the interactive analysis of forests of trees using treemaps.
@@
 The purpose is to provide readily available tools for users to explore and improve ensemble fits and obtain an intuition for the underlying class structure in data.
-Interactive plots are a key component for model visualization that helps the user see multivariate relationships and be more efficient in the model diagnosis.
+Interactive plots are a key component for model visualization that helps the user see multivariate relationships and be more efficient in model diagnosis.
 Multiple levels of data are constructed for exploration: observation, model, and ensemble summaries.
@@
 Two data set are used in this paper, crab ~\citep{CM74} and fishcatch data~\citep{fishcatch}.
-To illustrate the mapping of diagnostics to visualization crab data are used because its simplicity.
-The second data set is introduced to illustrate the interactive web app which combines the visualizations introduced in this paper and present a more complex problem to explore within the applicarion.
+To illustrate the mapping of diagnostics to visualization crab data are used because of its simplicity.
+The second data set is introduced to illustrate the interactive web app which combines the visualizations introduced in this paper and present a more complex problem to explore within the application.
 
 The article is structured as follows.
-Section \ref{key} describes the ensemble components to be accessed.
+Section \ref{key} describes the ensemble components to be assessed.
 Section \ref{visen} maps the ensemble components to the visual elements.
-The web app is described in \ref{app} and further work is discussed in Section \ref{fur}.
+The web app is described in Section \ref{app} and further work is discussed in Section \ref{fur}.
 
@@
 Cases that are predicted to be multiple classes, with similar class probabilities, indicate difficulty to classify observations.
-They may be important by indicating neighborhoods of the data space that would benefit from a more complex model, or more simply, they may be errors in measurements in the data.\\
+They may be important by indicating neighborhoods of the data space that would benefit from a more complex model, or simply, measurement errors in the data.\\
 
@@
 
-\item {\em \textbf{Similarity measure for pairs of observations:}} In each model, each pair of observations will be either in the same terminal node or not.
+\item {\em \textbf{Similarity measure for pairs of observations:}} In each model, a pair of observations will be either in the same terminal node or not.
 This is used to compute a proximity matrix.
@@
 \end{itemize}
-In addition to these overall ensemble statistics, each component model has its own diagnostics, measuring error, variables utilized, and class predictions.
+In addition to these overall ensemble statistics, each individual model has its own diagnostics, measuring error, variables utilized, and class predictions.
 Visualization will enable the individual models to be examined, relate these to the data and their contribution to the ensemble.
@@
 
-The PPF is composed of individual projection pursuit trees, PPtree algorithm uses a multi-step approach to fit a multi-class model by finding linear combinations to split on based on projection pursuit algorithm.
+The PPF is composed of individual projection pursuit trees.
+PPtree algorithm uses a multi-step approach to fit a multi-class model by finding linear combinations to split on based on projection pursuit algorithm.
 Two projection pursuit indexes are used to find projections that separate classes,  LDA  \citep{lee2005projection} and PDA \citep{lee2010projection}.
@@
 
-\noindent where notation is the same as for the LDA index, with the addition of $\lambda \in [0,1)$ is a shrinkage parameter, and a different within group sum of squares, $\blW_{PDA}(\lambda)=\mbox{diag}(\blW)+(1-\lambda)\mbox{offdiag}(\blW)$.
+\noindent where the notation is analogous to the LDA index, with the addition of $\lambda \in [0,1)$ which is a shrinkage parameter, and a different within group sum of squares, $\blW_{PDA}(\lambda)=\mbox{diag}(\blW)+(1-\lambda)\mbox{offdiag}(\blW)$.
 
@@
 There are three nodes for the four class problem.
-The nodes of this tree are based on projections of the data; the coefficients of which form the building block to calculate the variable importance.
+The nodes of this tree are based on projections of the data and their coefficients form the building block to calculate the variable importance.
 The density plot displays the data projection at each node, and the mosaic plot shows the confusion matrix for the nodes.
@@
 The dashed vertical red line indicates the split value of each node.
-At node 1 separates the sexes.
-Nodes 2 separate males from species and node 3 separate females from species.
+Node 1 separates the sexes.
+Node 2 separate males from species and node 3 separate females from species.
 Mosaic plots of the confusion table for each split (bottom).
@@
 The global permuted importance measure is the average importance over all the trees in the forest.
-This measure is based on comparing the accuracy of classifying oob observations using the true class with permuted (nonsense) class.
+This measure is based on comparing the accuracy of classifying oob observations using the true class with permuted (nonsense) classes.
 
@@
 
-\noindent where $e_k$ is the oob error rate of tree $k$, $I_{sjk}$ is the projection pursuit index value of node $s$ ~\cite{da2021projection}.
+\noindent where $e_k$ is the oob error rate of tree $k$, $I_{sjk}$ is the projection pursuit index value of node $s$ ~\citep{da2021projection}.
 
@@
 The scatterplot at right shows these two variables and the resulting boundary between groups that this would produce.
-Node 2 uses BD and CL, BD contributes the most to the separation.
+Node 2 uses BD and CL where BD contributes the most to the separation.
 The plot in the right shows the boundary that is induced.
-Node 3 uses CW and FL, and with an even contribution by the two variables.
+Node 3 uses CW and FL with an even contribution by the two variables.
 For each tree in the forest, different decision rules are defined; the resulting boundaries on the previous plots are based on Rule 1 $= \frac{m_1}{2} + \frac{m_2}{2} $, where $m_1$ and $m_2$ are the mean of the left and right groups at each node.
@@
 \caption{Visualizing the variable importance of all the trees in the forest, for three nodes.
-Each node of each tree has an importance value for each variable.
+Every particular node of each tree has an importance value for the corresponding variable.
 The values for the whole forest are displayed using a side-by-side jittered dot plot.
@@
 Tallying this up across all trees in a forest gives the proximity matrix, an $n\times n$ matrix of the proportion of trees that the pair share a terminal node.
-It can be considered a similarity matrix.
+It can be considered as a similarity matrix.
 
@@
 Color indicates the true species and sex.
-For this data two dimensions are enough to see the four groups separated quite well.
+For this data, two dimensions are enough to see the four groups separated quite well.
 Some crabs are clearly more similar to a different group, especially in examining the sex differences.
@@
 @
-\caption{Generalized ternary plot ((G-1) dimensional simplex, here it is a tetrahedron) representation of the vote matrix for four classes.
+\caption{Generalized ternary plot ((G-1) dimensional simplex, in this case a tetrahedron) representation of the vote matrix for four classes.
 The tetrahedron is shown pairwise.
@@
 Because visualizing the vote matrix with a $(G-1)$ dimensional tetrahedron requires dynamic graphics, a low-dimensional option is also provided.
-For each class, each case has a value between 0-1.
+For each class, every individual case has a value between 0-1.
 A side-by-side jittered dotplot is used for the display, where the class is displayed on one axis and proportion is displayed on the other.
@@
 
- Interaction is added to the plots described in Section \ref{visen} and other plots, and they are organized into an interactive web app using \verb#shiny#~\citep{chang11shiny, wickham2021mastering} for exploring the ensemble model.
+ Interaction is added to the plots described in Section \ref{visen} as well as other plots, which are organized into an interactive web app using \verb#shiny#~\citep{chang11shiny, wickham2021mastering} for exploring the ensemble model.
 The app is organized into three tabs: individual cases; models; and performance comparison, to provide a model diagnostic tool.
-Interaction is provided as mouse-over labeling, mouse-click selection, and brushing, with results linked across multiple plots.
+Interaction is provided as mouse-over labeling, mouse-click selection and brushing, with results linked across multiple plots.
 The app takes advantage of new tools provided in the \verb#plotly#~\citep{plotly} package, developed as a part of Sievert's PhD thesis research~\citep{sievertthesis}.
@@
 
-To illustrate the shiny app characteristics a different data (fishcatch) are used which present a more challenging classification problem and then more interesting results to explore the app.
-The fishcatch data~\citep{fishcatch} contains 159 observations, with 6 physical measurement variables, and  7 types of fish, all caught from the same lake (Laengelmavesi) near Tampere in Finland.
+To illustrate the shiny app characteristics, a different dataset (fishcatch) are used which presents a more challenging classification problem and then more interesting results to explore the app.
+The fishcatch dataset~\citep{fishcatch} contains 159 observations, with 6 physical measurement variables, and  7 types of fish, all caught from the same lake (Laengelmavesi) near Tampere in Finland.
 There are 35 bream, 11 parkki, 56 perch, 17 pike, 20 roach, 14 smelt and 6 whitewish.
@@
 In a PCP many variables can be compared together and seeing the relationships between them, identify and characterize groups of observations.
-The features can be ordered so that there are not too many lines intersecting resulting in an unreadable chart \citet{inselberg2009parallel}.
+The features can be ordered so that there are not too many lines intersecting resulting in an unreadable chart \citep{inselberg2009parallel}.
 Two alternatives can be selected in shiny to draw the parallel coordinate plot: parallel or enhanced.
@@
 
-\newpage
-
 \subsection{Models}
@@
 The tree plot from the \verb#PPtreeViz# is used to visualize the selected tree structure.
-Also the variable importance values are highlighted for each variable for each of the top three nodes, by default, and the oob error value for the selected tree on the boxplot (both in red).
+In addition, the variable importance values are highlighted for each variable for the top three nodes, by default, and the oob error value for the selected tree on the boxplot (both in red).
 Additionally it is possible to select specific nodes and update the importance variable plot, density plot and mosaic plot (nodes can be selected on the right corner on this second tab) and finally the interaction can be driven from the variable importance plot.
@@
 The selected tree in Figure \ref{tab2} has very small oob error shown as red point in the boxplot.
-Two variables were use in each node partition, the first four nodes were selected and display in the importance variable plot, density plot and mosaic plot.
-In the importance variable plot `height` and `lenght1` were used in the first node partition being `height` the important variable to separate Bream and Parkki (Green and orange in density and mosaic plots) from the rest.
-In the second node `length1` and `length2` were both evenly important  to separate Bream from Parkki.
+Two variables were use in each node partition, the first four nodes were selected and shown in the importance variable plot, density plot and mosaic plot.
+In the importance variable plot \verb#height# and \verb#lenght1# were used in the first node partition being \verb#height# the important variable to separate Bream and Parkki (Green and orange in density and mosaic plots) from the rest.
+In the second node \verb#length1# and \verb#length2# were both evenly important  to separate Bream from Parkki.
 A similar analysis can be done for other nodes.
+This tab allow us to understand individual trees in the forest and analyzed which variables are important for the node partition related with the oob error involved in that specific tree.
 
@@
 \end{figure}
+\newpage
 
@@
 The third tab (Figure \ref{tab3}) examines the PPF fit and compares the result with a RF fit.
-There are four displays for each type of model: (1) Variable importance for all trees in the forest (same as in the models tab), (2) a receiver operating characteristic curve (ROC) curve comparing sensitivity and specificity for each class, (3) oob error by the number of trees to assess complexity, (4) overall variable importance.
+There are four displays for each type of model: (1) Variable importance for all trees in the forest (same as in the models tab), (2) a receiver operating characteristic curve (ROC) comparing sensitivity and specificity for each class, (3) oob error by the number of trees to assess complexity, (4) overall variable importance.
 There is very little interaction on this tab.
@@
 
-Having better tools to open up black box models will provide for better understanding the data, the model strengths and weaknesses, and how the model will perform on future data.
+Having better tools to open up black box models will provide a better understanding of the data, the model strengths and weaknesses, and how the model will perform on future data.
 This visualisation app provides a selection of interactive plots to diagnose PPF models.
@@
 
-The application have some limitations that are need to pointed out, if the number of observation is big the overploting will be an issues in some of the selectedd visualizations and if the number of variables increases, plots like parallel plot and importance variable plot should include the option to select some variables each time or a different approach.
-
-There are many possible extensions to the app, that could help it to be a tool for model refinement: (1) Using the diagnostics to weed out under-performing models in the ensemble; (2) Identifying and boosting models that perform well, particularly if they do well for problematic subsets of the data; (3) Problematic cases could be removed, and ensembles re-fit; (4) Classes as a whole could be aggregated or re-organised as suggested by the model diagnostics, to produce a more effective hierarchical approach to the multiple class problem.
+The application have some limitations that need to be pointed out.
+If the number of observation is big, the overploting will be an issue in some of the selected visualizations, and if the number of variables increases, plots like parallel coordinate plot and importance variable plot should include the option to select some variables each time or a different approach must be considered.
+
+There are many possible extensions to the app that could help to become a tool for model refinement: (1) Using the diagnostics to weed out under-performing models in the ensemble; (2) Identifying and boosting models that perform well, particularly if they do well for problematic subsets of the data; (3) Problematic cases could be removed, and ensembles re-fit; (4) Classes as a whole could be aggregated or re-organised as suggested by the model diagnostics, to produce a more effective hierarchical approach to the multiple class problem.
 Working within the R environment makes all of these desires available using command line outside the app, given the unique IDs of models and cases can be exported from the app.
 
-The app has helped to identify ways to improve the  PPtree algorithm and consequently, the PPF model.
-These especially apply to multiclass problems.
+The app has helped to identify ways to improve the  PPtree algorithm and consequently, the PPF model which especially apply to multiclass problems.
 Multiple splits for the same class would enable nonlinear classifications.

### 2022-06-29 PPforestVizupdate.Rnw
Fig 3 code change

@@
 \end{figure}
-Two data set are used in this paper, crab ~\citep{CM74} and fishcatch data~\citep{fishcatch}.
+Two data set are used in this paper, crab~\citep{CM74} and fishcatch data~\citep{fishcatch}.
 To illustrate the mapping of diagnostics to visualization crab data are used because of its simplicity.
@@
   if(node%in%1) dat = dat_2d(node)
-  if(node%in%2) dat = dat_2d(node) %>%  filter(Type %in% c("OrangeFemale", "OrangeMale") )
-  if(node%in%3) dat = dat_2d(node) %>%  filter(Type %in% c("BlueFemale", "BlueMale") )
+  if(node%in%2) dat = dat_2d(node) %>%  filter(Type %in% c("BlueMale", "OrangeMale") )
+  if(node%in%3) dat = dat_2d(node) %>%  filter(Type %in% c("BlueFemale", "OrangeFemale") )
 
@@
   if(node%in%2){
-    p1 <-   p1 +  scale_color_manual(values = myColors[3:4])
+    p1 <-   p1 +  scale_color_manual(values = myColors[c(2,4)])
   }
   if(node%in%3){
-    p1<-  p1 +  scale_color_manual(values = myColors[1:2])
+    p1<-  p1 +  scale_color_manual(values = myColors[c(1,3)])
   }
@@
 Interaction is provided as mouse-over labeling, mouse-click selection and brushing, with results linked across multiple plots.
-The app takes advantage of new tools provided in the \verb#plotly#~\citep{plotly} package, developed as a part of Sievert's PhD thesis research~\citep{sievertthesis}.
+The app takes advantage of new tools provided in the \verb#plotly# package ~\citep{plotly,sievert2020interactive}, developed as a part of Sievert's PhD thesis research~\citep{sievertthesis}.
 

### 2022-06-29 response_reviewers.Rmd
update response to reviewer

@@
 
-\textcolor{violet}{We appreciate all of your feedback. 
+\textcolor{violet}{We appreciate all of your feedback which help to improve our paper. 
  }
@@
 \textcolor{violet}{The app was modified to adress this issue.
-Now the user can select specific nodes to see (three or more nodes).
-If there are a lot of nodes will be better to select 4 or less each time to see better the complete display.}
+In the new application the user can select specific nodes to see (from one to all the nodes).
+If there are a lot of nodes will be better to select 4 or less each time to see better the complete display but the application is not limited to see just the first three nodes.}
 
@@
 
-\textcolor{violet}{TO DO}
+\textcolor{violet}{In the discussion section we incorporate the app limitations.}
 
@@
 \textcolor{violet}{The second tab was modifyed in different ways taking into account your valuable comments.
-We have included three main changes in the second tabset, first we have included jittered points in the OOB error boxplot where the user can select trees based in their performance instead of selecting trees based on importance variable.
+We have included three main changes in the second tabset, first we have included jittered points in the OOB error boxplot where the user can select trees based in their performance.
 Second, we have included a panel to select the nodes to see in the importance variable plot, the mosaic plot and the density plot.
 The panel allows you to select more or less than three nodes and the mentioned plots are updated acordingly.
-Finally, we rearrange some of the plots to make the selection and interaction more intuitive  presented in Figure 8.}
+Finally, we rearrange some of the plots to make the selection and interaction more intuitive  presented in Figure 10.}
 
@@
 Furthermore, I would show some actual examples of screenshots of what shows when you hover over observations.
+
+\textcolor{violet}{In Section 4 we have included more details and examples of use and interpreting results to address your comment.}
 
@@
 
-\textcolor{violet}{Formulas for all VI were defined in equations and mentionedd in the text. It  is more clear now.}
+\textcolor{violet}{Formulas for all VI were defined in equations and mentioned in the text. It  is more clear now.}
 
@@
 
-\textcolor{violet}{NEED HELP, not sure how to improve this sentence and I do not know what is more descriptive for widget.
-At their core Shiny widgets are mini-applications created using the shinyApp function.}
+\textcolor{violet}{ At their core Shiny widgets are mini-applications created using the shinyApp function.}
 

### 2022-06-29 PPforestVizupdate.Rnw
update

@@
 load("ppf2022.RData")
-load("rf_fish.RData")
+load("rf_fish.Rdata")
  ppf<- ppf2022

### 2022-06-30 response_reviewers.Rmd
modifications to response
revised Discussion

@@
 
+<!--
 Dear Mrs. da Silva:
+-->
 
@@
 Your list of responses should be uploaded as a file in addition to your revised manuscript.
+
+<!--
 
@@
 
-\textcolor{violet}{Thank you. We appreciate the chance to do a further revision, and hope that these changes satisfy the AE and reviewer.}
+-->
+
+\textcolor{violet}{Thank you. We appreciate the chance to do a further revision, and hope that these changes are satisfactory.}
 
@@
 
-\textcolor{violet}{In the discussion section we incorporate the app limitations.}
+\textcolor{violet}{The discussion has been extended to include limitations.}
 

### 2022-06-30 PPforestVizupdate.Rnw
modifications to response
revised Discussion

@@
 \begin{document}
+\SweaveOpts{concordance=TRUE}
 
@@
 
-The app is implemented with new technology for interactive graphics provided by the \verb# plotly# package. It is one of the first uses of these new tools.
-
-One challenge to use plotly is that  when layers with different data are created in a ggplot2,  it is difficult to  specify the unique keys required for linking with another plot.
-
-The application have some limitations that need to be pointed out.
-If the number of observation is big, the overploting will be an issue in some of the selected visualizations, and if the number of variables increases, plots like parallel coordinate plot and importance variable plot should include the option to select some variables each time or a different approach must be considered.
-
-There are many possible extensions to the app that could help to become a tool for model refinement: (1) Using the diagnostics to weed out under-performing models in the ensemble; (2) Identifying and boosting models that perform well, particularly if they do well for problematic subsets of the data; (3) Problematic cases could be removed, and ensembles re-fit; (4) Classes as a whole could be aggregated or re-organised as suggested by the model diagnostics, to produce a more effective hierarchical approach to the multiple class problem.
-Working within the R environment makes all of these desires available using command line outside the app, given the unique IDs of models and cases can be exported from the app.
+The app is implemented using the technology for interactive graphics provided by the \verb# plotly# package.
+It was one of the first to experiment with the application of these tools.
+One challenge with plotly is that  when layers containing different data are created in a ggplot2, it is difficult to specify the unique keys required for linking with other plots.
+
+There are some limitations with providing functionality through the app, and generally with the displays provided.
+Successful apps make it easier for the user to do common tasks, and this is the design principal here.
+Only limited functionality is provided to the user, but the code is available and a knowledgeable data analyst might use this as a basis for extending the methods as appropriate for theit data.
+Performance of the app is good for relatively small data sets.
+If the number of observation is large, overploting will be an issue in some of the visualizations.
+If the number of variables is large, the parallel coordinate plot and plot of variable importance would benefit by the addition of a user input to select variables.
+
+There are many possible extensions to the app that could help it to be a tool for model refinement:
+
+\begin{enumerate}
+\item Using the diagnostics to weed out under-performing models in the ensemble. 
+\item Identifying and boosting models that perform well, particularly if they do well for problematic subsets of the data.
+\item Problematic cases could be removed, and ensembles re-fit. 
+\item Classes as a whole could be aggregated or re-organised as suggested by the model diagnostics, to produce a more effective hierarchical approach to the multiple class problem.
+
+\noindent Working within the R environment makes all of these possible when using command line outside the app, given that the unique IDs of models and cases can be exported from the app.
 
@@
 Split criteria tend to place boundaries too close to some groups, due to heteroskedasticity being induced by aggregating classes.
-Forests are not always better than their constituent trees, and if the trees can be built better, the forest will provide stronger predictions.
+Forests are not always better than their constituent trees, and if the trees can be built better, the forest will provide better predictions.
 

### 2022-06-30 response_reviewers.Rmd
finished reviewing the changes

@@
 
-\textcolor{violet}{We have included a description of parallel coordinate plot in Section 4 were this figure appears and the corresponding reference for further details.
+\textcolor{violet}{We have included more explanaition about the parallel coordinate plot in Section 4 where this figure appears and added a corresponding reference for further details.
 In the vertical axis the values for each variable are standardized to easily compare them.}
@@
 
-\textcolor{violet}{The second tab was modifyed in different ways taking into account your valuable comments.
+\textcolor{violet}{The second tab was modified based on your helpful comments.
 We have included three main changes in the second tabset, first we have included jittered points in the OOB error boxplot where the user can select trees based in their performance.
@@
 The panel allows you to select more or less than three nodes and the mentioned plots are updated acordingly.
-Finally, we rearrange some of the plots to make the selection and interaction more intuitive  presented in Figure 10.}
+Finally, we rearranged some of the plots to make the selection and interaction more intuitively presented in Figure 10.}
 
@@
 
-\textcolor{violet}{In Section 4 we have included more details and examples of use and interpreting results to address your comment.}
+\textcolor{violet}{In Section 4 we have added more details and examples of use and interpreting results to address your comment.}
 
@@
 
-\textcolor{violet}{Random Forest presents two sources of randomness in the tree: random variable selection and random bootstrap samples.
-The random variables selection used to node partition, each split are identified from a random subset of all variables.
-There is not an initial sample of variables for each tree, the sample is for each tree in each  node partition, in each case using all the variables.
-We did not explore formally the sensitivity of the random variable sample size in the importance variable measure.
-In the algorithm you can select the proportion of variables used in each node partition.
-In the examples presented in the paper for crab and fishcatch data we use 2 variables in each node partition.
+\textcolor{violet}{A random forest has two sources of randomness: variable sampling and bootstrap samples.
+The variable sampling is used for node partition, where each split is produced from a random subset of variables.
+There is not an initial sample of variables for each tree.
+
+We did not formally explore the sensitivity of the random variable sample size for the measure of variable importance.
+In the algorithm you can select the proportion of variables to be sampled at each node.
+In the crab and fishcatch examples presented in the paper two variables were sampled at each node.
+
 We have included an image from the original paper with the boundary that would result from a classification tree fitted using the PPtree algorithm to take into account your last request.
 The plot shows a simple example from simulated data with two features and three classes.
-The partitions generated by PPtree algorithm are oblique to the axis, incorporating the association between the two variables.
+The partitions generated by the PPtree algorithm are oblique to the axis, incorporating the association between the two variables.
 }
@@
 But in the upper right corner, the plots of the variables distributions, for instance for node 2, are not in the standardized range (node 2 has an x-axis that goes up to 5 and 6).
-\textcolor{violet}{As you mentioned the data should be standardized for the coefficients to be interpretable as importances, we fixed the density plot and the x axis are in standardized range now.}
+\textcolor{violet}{As you mentioned the data should be standardized for the coefficients to be interpretable as importances. We have fixed the density plot and the x axis accordingly.}
 
 Is the distribution being plotted here the value of the linear combination for each class, rather than a single variable? 
-
-\textcolor{violet}{The density plots shows the data projection at each node. }
+\textcolor{violet}{Yes, the density plots shows the data projection at each node. This is now made clear in the caption.}
 
@@
 
-\textcolor{violet}{We have changed the example but even in the previous one in the first node partition the separation was not perfect as you mentioned.
-We explore your suggestion (changing the bound for the KDE to observe range) but the result was not better and we keep the original.}
+\textcolor{violet}{We explore your suggestion (changing the bound for the KDE to observe range) but the result was not better so we have kept the original.}
 
@@
 
-\textcolor{violet}{We have complete control over the color palette and the selected palette (Dark2) is a qualitative palette which is suited to represent categorical data and is also colorblind friendly.
-In this case we are not changing the color palette as you suggested  because maybe for the density plot your idea will work but in the other visualizations using different shades to differentiate sexes will not be easy to identify when the geometric element for example are points.
-Also in terms of perception using different colors is better than using shapes to differentiate classes.}
+\textcolor{violet}{The selected palette (Dark2) is a qualitative palette suited to mapping a categorical data because it is colorblind friendly.
+It is a little unfortunate that the categories in the crabs data are colors, which makes the explanation confusing.
+We have changed the category names to be simply O and B so that there is less confusion.}
 
 Also, can you include labeling in your plot that the sub-plots are node ID (e.g., put the label "node" above the 3 density plots).
-\textcolor{violet}{We have included in each sub-plots "Node" before the number and also update this in the app}
+\textcolor{violet}{We have included in each sub-plots "Node" before the number and also update this in the app.}
 
-What do the numbers in the grey boxes at the terminal nodes represent?
-
-\textcolor{violet}{The numbers  in the gray boxes are the predicted class, we fixed the plot to show the class names}
+What do the numbers in the grey boxes at the terminal nodes represent? \textcolor{violet}{The numbers in the gray boxes are the predicted class, we fixed the plot to show the class names}
 
@@
 \textcolor{violet}{Yes, the coefficient values are between -1 and 1 by construction.
-We find the linear combinations, \textbf{a} which maximize $\frac{a^TBa}{a^TWa}$, where B y W are the between and within group sums of squares.
+We find the linear combinations, \textbf{a} which maximize $\frac{a^TBa}{a^TWa}$, where $B$ and $W$ are the between and within group sums of squares.
 We get $\hat a_1=e_1$ where $e_1...e_s$ are eigenvectors of $W^{-1}B$.
 In this case we find the 1-D projection optimizing the LDA index.
-We have included in Section 3 LDA and PDA indexes which are optimized to get the linear projections and it is more clear now how the coefficients are computed.
+We have included in Section 3 LDA and PDA indexes which are optimized to get the linear projections to make it more clear how the coefficients are computed.
 }
@@
 
-\textcolor{violet}{Formulas for all VI were defined in equations and mentioned in the text. It  is more clear now.}
+\textcolor{violet}{Formulas for all VI were defined in equations and mentioned in the text. It is clearer now.}
 
@@
 
-\textcolor{violet}{MDS citation was included.
+\textcolor{violet}{A reference for MDS has been included.
 In Section 3 we describe the mapping of diagnostics to visualizations.
-MDS is presented in Subsection 3.3 (Similarity of cases) the key connection with the forest is that we are using the proximity matrix information which is one of the diagnostics described for PPforest or RF and we use this information to do the MDS.
-This visualization can be useful to identify in which cases the model is working well and when is having problems to classify cases.
-MDS and generalized ternary plot (tetrahedral plot in case there are  four clases) are both implemented in the app and we included a comment in the paper to clarify this point. }
+MDS is presented in Subsection 3.3 (Similarity of cases) the key connection with the forest is that we are using the proximity matrix information which is one of the diagnostics described for PPforest or RF and we use this information to compute the MDS.
+This visualization can be useful to identify in which cases the model is performing well and which it is having problems.
+
+MDS and generalized ternary plot (tetrahedral plot in case there are  four clases) are both implemented in the app and we have added a comment in the paper to clarify this. }
 
@@
 
-\textcolor{violet}{We selected a jittered side-by-side dot plot in this case instead of a box plot because we want to see the values for each tree.
-Jittering implys to add random noise to data in order to prevent overplotting.
-The x axis represents each variable but instead of using the same x position for all the points we use jittered values in an interval.
-This plot have more sence when is presented in the app conected with the others in the model level analysis.
-}
+\textcolor{violet}{We selected a jittered side-by-side dot plot in this case instead of a box plot because we want to see the values for each tree, and it makes it more sensible when linking observations between multiple plots.
+Jittering spreads the data on the categorical variable axis to better read the distribution.}
 
@@
 
-\textcolor{violet}{In Figure 4 the dimension of the simplex is G-1, since in crab data there are 4 clases in this case we work with a three dimensional simplex.
-Ternary plot can be generalized to more than three dimensional space, with G classes, the ternary plot idea is generalized to a G-1 dimensional simplex.
-Then if we have 4 classes the generalized ternary plot is a tetrahedron as shown in Figure 4 but if we have more than four classes we can also generalize the ternary plot idea.
-In the fishcatch data used in this paper to illustrate the shiny app there are 7 classes and we generalized the ternary plot to a 6-dimensional simplex as it is shown in the last panel of Figure 6 (6-didmensional simplex shown pairwise), using tourr package you can see better the 6-dimensional simplex in different rotations https://vimeo.com/707047875.
-The structure is complex because a 6-dimensional simplex has 7 vertices, 21 eddges and 35 faces.}
+\textcolor{violet}{In Figure 4 the dimension of the simplex is $G-1=3$, since in crab data there are 4 classes.
+Ternary plot can be generalized to more than three dimensional -- with $G>4$ classes, the ternary plot idea is generalized to a $G-1$ dimensional simplex.
+In the fishcatch data there are 7 classes and the ternary plot to a 6-dimensional simplex as it is shown in pairwise plots in Figure 6.
+Using the tourr package you can see the 6-dimensional simplex in a rotation through 6D, see https://vimeo.com/707047875.
+The structure is complex because a 6-dimensional simplex has 7 vertices, 21 eddges and 35 faces.
+This has been clarified in the paper.}
 
@@
 
-\textcolor{violet}{We are using points because it is important to see the individual cases and the jitter is to overcome overploting.
-It is not recommended using jittered scatterplots because you are changing the real values of your data.
-In this case we are not changing the values of a continuous variable just the position to overcome overplotting in a categorical one.
-}
+\textcolor{violet}{We are using points because it is important to see the individual cases and the jitter is to overcome overplotting on the categorical variable.
+It is not recommended using jittered scatterplots because it would change the real values of your data.}
 
@@
 
-\textcolor{violet}{ At their core Shiny widgets are mini-applications created using the shinyApp function.}
+\textcolor{violet}{At their core Shiny widgets are mini-applications created using the shinyApp function. We have changed the language of the paragraph to help with this terminology.}
 
@@
 
-\textcolor{violet}{Yes fishcatch data are more difficult to classify and then more interesting to show some of the results.
-For simplicity to describe the main plot of the app we use crab data but to describe the app we prefer to present a more complex problem, more clases, umbalanced data...}
+\textcolor{violet}{Yes, fishcatch data are more difficult to classify and then more interesting to show some of the results.
+For simplicity to describe the main plot of the app we use crab data but to describe the app we prefer to present a more complex problem, more classes, and imbalanced data.}
 
@@
 
-\textcolor{violet}{PDP can be order in different ways, we have modified the PDP and included variable order based on the mean value.
-As you suggested, we have also change the grid lines and just keep the vertical grid in gray, it is more clear now.
-In the MDS we just selected two dimensions but can be modify if it is needed by the problem.
+\textcolor{violet}{A PCP can be ordered in different ways, and we have modified the PCP to include variable order based on the mean value.
+As you suggested, we have also change the grid lines to keep only the vertical grid in gray.
+In the MDS we only used two dimensions but can be more if deemed more appropriate for a particular problem.
 We have included labels in the generalized ternary plot as we did in Figure 7.}
@@
 \textcolor{violet}{This comment is in page 12, In the original RF there is not an option to prune under-performing trees.
-This comment is just to mention a possible use for the information we get from the app, also we are presenting the app using PPforest and this can be a possible extension for our algorithm.}
+The comment is just to mention a possible use for the information we get from the app, also we are presenting the app using PPforest and this can be a possible extension for our algorithm.}
 
@@
 \textcolor{violet}{In the upper LH corner we are using the function `selectizeInput()` from shiny R package which creates a select list that can be used to choose a single or multiple items from a list of values, you do not have to type the class names you just select the names from a list.
-For the bottom two plots (variable importance plots) we have fixed the axis and  include titles in each plot, now is easy to compare them.}
+For the bottom two plots (variable importance plots) we have fixed the axis and include titles in each plot, so it is easier to compare them.}
 
@@
 
-\textcolor{violet}{Not necessarily, in Breiman Random Forest original paper always is written as `out-of-bag' without abbreviation.
-I think if we are consistent with the abbreviation we can use oob or OOB indistinctly.}
+\textcolor{violet}{We have changed to using OOB now.}
 
@@
 \textcolor{violet}{Cases that are predicted to be multiple classes, without a clear majority in any of them, indicate that those cases are difficult to classify.
-The paragraph was not clear and we modify it to:" Cases that are predicted to be multiple classes, with similar class probabilities, indicate difficulty to classify observations"  }
+The paragraph was not clear and we modify it to:" Cases that are predicted to be multiple classes, with similar class probabilities, indicate difficult to classify observations"  }
 
@@
 
-\textcolor{violet}{Yes, should be Thus. It is done}
+\textcolor{violet}{Yes, should be Thus. It is fixed.}
 

### 2022-06-30 PPforestVizupdate.Rnw
finished reviewing the changes

@@
 If a case is always predicted to be the true class, there is no uncertainty about an observation.
-Cases that are predicted to be multiple classes, with similar class probabilities, indicate difficulty to classify observations.
+Cases that are predicted to be multiple classes, with similar class probabilities, indicate  difficult to classify observations.
 They may be important by indicating neighborhoods of the data space that would benefit from a more complex model, or simply, measurement errors in the data.\\
@@
 
-As \cite{sievert2020interactive} describes, one of the biggest difficulties when it comes to the app managing linking between plots is the data structure management for each widget.
-Each widget has it own data structure and interaction.
+As \cite{sievert2020interactive} describes, one of the biggest difficulties when it comes to the app is managing linking between plots (also called a widget in the plotly documentation) requiring careful data structure management.
+Each plot has it own data structure and interaction.
 Putting them into the structure of a shiny app facilitates access to the widget data and coordinates selections across multiple plots.

### 2022-06-30 response_reviewers.Rmd
finished reviewing the changes

@@
 The variable sampling is used for node partition, where each split is produced from a random subset of variables.
-There is not an initial sample of variables for each tree.
+There is not an initial sample of variables for each tree.}
 
-We did not formally explore the sensitivity of the random variable sample size for the measure of variable importance.
+\textcolor{violet}{We did not formally explore the sensitivity of the random variable sample size for the measure of variable importance.
 In the algorithm you can select the proportion of variables to be sampled at each node.
-In the crab and fishcatch examples presented in the paper two variables were sampled at each node.
+In the crab and fishcatch examples presented in the paper two variables were sampled at each node.}
 
-We have included an image from the original paper with the boundary that would result from a classification tree fitted using the PPtree algorithm to take into account your last request.
+\textcolor{violet}{We have included an image from the original paper with the boundary that would result from a classification tree fitted using the PPtree algorithm to take into account your last request.
 The plot shows a simple example from simulated data with two features and three classes.
-The partitions generated by the PPtree algorithm are oblique to the axis, incorporating the association between the two variables.
-}
+The partitions generated by the PPtree algorithm are oblique to the axis, incorporating the association between the two variables.}
 
@@
 MDS is presented in Subsection 3.3 (Similarity of cases) the key connection with the forest is that we are using the proximity matrix information which is one of the diagnostics described for PPforest or RF and we use this information to compute the MDS.
-This visualization can be useful to identify in which cases the model is performing well and which it is having problems.
+This visualization can be useful to identify in which cases the model is performing well and which it is having problems.}
 
-MDS and generalized ternary plot (tetrahedral plot in case there are  four clases) are both implemented in the app and we have added a comment in the paper to clarify this. }
+\textcolor{violet}{MDS and generalized ternary plot (tetrahedral plot in case there are  four clases) are both implemented in the app and we have added a comment in the paper to clarify this.}
 

### 2022-06-30 PPforestVizupdate.Rnw
incorporate Di comments from github issues

@@
 \begin{document}
-\SweaveOpts{concordance=TRUE}
 
@@
 It combines the predictions from individual classification and regression trees (CART)~\citep{breiman1984cl} built by bagging observations~\citep{breiman1996bagging} and random predictor sample at each tree node.
-These produce diagnostics in the form of uncertainty in predictions for each observation, importance of variables for the prediction, predictive error for future samples based on out-of-bag (oob) case predictions, and similarity of observations based on how often they group together in the trees.
+These produce diagnostics in the form of uncertainty in predictions for each observation, importance of variables for the prediction, predictive error for future samples based on out-of-bag (OOB) case predictions, and similarity of observations based on how often they group together in the trees.
 
@@
 Predicting the response for these cases gives a better estimate for the error of the model with future data.
-The oob error rate is a measure for each model that is combined in the ensemble and is used to provide the overall error of the ensemble.\\
+The OOB error rate is a measure for each model that is combined in the ensemble and is used to provide the overall error of the ensemble.\\
 
@@
 These are illustrated using the Australian crabs data~\citep{CM74}.
-The data has 200 cases, 5 predictors, and 4 classes (combinations of species and sex, blue male, blue female, orange male, and orange female).
+The data has 200 cases, 5 predictors, and 4 classes (combinations of species and sex, BM, BF, OM, and OF).
 The predictors are: FL (the size of the frontal lobe length, in mm), RW (rear width, in mm), CL (length of mid-line of the carapace, in mm), CW (maximum width of carapace, in mm), BD (depth of the body; for females, measured after displacement of the abdomen, in mm).
@@
 Node 1 shows the clear split of sexes, with a small number of misclassifications.
-Node 2 where blue males are separated from orange males with small number of misclassifications.
-Node 3 where blue females are separated from orange females, almost perfect classification.\label{pptreeviz1}}
+Node 2 where BM are separated from OM with small number of misclassifications.
+Node 3 where BF are separated from OF, almost perfect classification.\label{pptreeviz1}}
 \end{figure}
@@
 The first importance measure is comparable with the measure defined in the classical random forest algorithm.
-Importance variable is computed based on the oob cases for the tree $k\;\;(O_k)$ for each $j$ predictor variable.
+Importance variable is computed based on the OOB cases for the tree $k\;\;(O_k)$ for each $j$ predictor variable.
 Permuted importance of the variable $j$ in the tree $k$ is defined in Equation \ref{vi1}.
@@
 The global permuted importance measure is the average importance over all the trees in the forest.
-This measure is based on comparing the accuracy of classifying oob observations using the true class with permuted (nonsense) classes.
+This measure is based on comparing the accuracy of classifying OOB observations using the true class with permuted (nonsense) classes.
 
@@
 
-\noindent where $e_k$ is the oob error rate of tree $k$, $I_{sjk}$ is the projection pursuit index value of node $s$ ~\citep{da2021projection}.
+\noindent where $e_k$ is the OOB error rate of tree $k$, $I_{sjk}$ is the projection pursuit index value of node $s$ ~\citep{da2021projection}.
 
@@
   if(node%in%3) dat = dat_2d(node) %>%  filter(Type %in% c("BlueFemale", "OrangeFemale") )
+
+   levels(dat$Type) <- c("BF","BM", "OF", "OM")
 
@@
 df <- data.frame(Class = ppf$train[, 1], rf.mds$points)
+levels(df$Class) <- c("BF", "BM", "OF", "OM")
 p12 <-  ggplot2::ggplot(data = df) +
@@
 
-The vote matrix ($n \times G$) contains the proportion of times each observation was classified to each class while oob. Two approaches to visualize the vote matrix information are used.
+The vote matrix ($n \times G$) contains the proportion of times each observation was classified to each class while OOB. Two approaches to visualize the vote matrix information are used.
 
@@
 This data is close to the ideal but not perfect, e.g.
-there are a few blue male crabs (orange) that are frequently predicted to be blue females (green), and a few blue female crabs predicted to be another class.
+there are a few BM crabs (orange) that are frequently predicted to be BF (green), and a few BF crabs predicted to be another class.
 
@@
   gather(Class, Probability, -pred, -ids, -Type)
+
+  levels(voteinf$Type) <- c("BF", "BM", "OF", "OM")
+  levels(voteinf$pred) <- c("BF", "BM", "OF", "OM")
+  voteinf$Class <- as.factor(voteinf$Class)
+  levels(voteinf$Class) <- c("BF", "BM", "OF", "OM")
 
@@
 On each dotplot, a single color dominates the top, indicating fairly clear distinctions between classes.
-Crabs from the blue species, both male and female, have more uncertainty in predictions, as seen by more crabs from other classes having higher vote proportions than is seen in the orange species.}
+Crabs from the B species, both male and female, have more uncertainty in predictions, as seen by more crabs from other classes having higher vote proportions than is seen in the O species.}
 \label{sideby}
@@
 This second tab in the app focuses on teasing apart the forest to examine the qualities of each tree.
-For each tree, information on the variable importance, the projections used at each node, and the oob error is available.
+For each tree, information on the variable importance, the projections used at each node, and the OOB error is available.
 The data feeding into this tab is a list of models along with the original data frame.
@@
 Figure \ref{tab2} is a screenshot of the models tab.
-There are five plots, with varying levels of interaction: (1) a boxplot of oob error for all trees including the data as jittered points, (2) a jittered side-by-side dotplot showing variable importance for the top three nodes of all trees in the forest, (3) a static display of one PPtree, (4) a faceted density plot of projected data at each node of the tree, with split point indicated by a vertical line, and (5) a mosaic plot showing the confusion matrix for each node of the tree.
+There are five plots, with varying levels of interaction: (1) a boxplot of OOB error for all trees including the data as jittered points, (2) a jittered side-by-side dotplot showing variable importance for the top three nodes of all trees in the forest, (3) a static display of one PPtree, (4) a faceted density plot of projected data at each node of the tree, with split point indicated by a vertical line, and (5) a mosaic plot showing the confusion matrix for each node of the tree.
 The interaction is driven in three ways, from the boxplot -- when the user selects a point in that display, the corresponding importance variable plot, tree plot from the \verb#PPtreeViz#, tree density displays and mosaic plots are drawn.
 The tree plot from the \verb#PPtreeViz# is used to visualize the selected tree structure.
-In addition, the variable importance values are highlighted for each variable for the top three nodes, by default, and the oob error value for the selected tree on the boxplot (both in red).
+In addition, the variable importance values are highlighted for each variable for the top three nodes, by default, and the OOB error value for the selected tree on the boxplot (both in red).
 Additionally it is possible to select specific nodes and update the importance variable plot, density plot and mosaic plot (nodes can be selected on the right corner on this second tab) and finally the interaction can be driven from the variable importance plot.
 
-The selected tree in Figure \ref{tab2} has very small oob error shown as red point in the boxplot.
+The selected tree in Figure \ref{tab2} has very small OOB error shown as red point in the boxplot.
 Two variables were use in each node partition, the first four nodes were selected and shown in the importance variable plot, density plot and mosaic plot.
@@
 A similar analysis can be done for other nodes.
-This tab allow us to understand individual trees in the forest and analyzed which variables are important for the node partition related with the oob error involved in that specific tree.
+This tab allow us to understand individual trees in the forest and analyzed which variables are important for the node partition related with the OOB error involved in that specific tree.
 
@@
 \caption{The individual model tab in the web app.
-The oob error is displayed in a boxplot  including jittered points for all trees in the forest.
+The OOB error is displayed in a boxplot  including jittered points for all trees in the forest.
 Variable importance is displayed as jittered dot plots for selected nodes of all trees.
 The boxplot is linked to a display of the PPTree, the variable importance plot, and a density plot of the data projections showing splits at each selected nodes and confusion tables as mosaic plots.
-Clicking a point in the oob error boxplot triggers various updates: each of the importance values for the same tree are highlighted (red), the tree that this corresponds to is drawn, the error for the tree is shown on the boxplot (in red), and the density plot and the mosaic plot are also updated.
+Clicking a point in the OOB error boxplot triggers various updates: each of the importance values for the same tree are highlighted (red), the tree that this corresponds to is drawn, the error for the tree is shown on the boxplot (in red), and the density plot and the mosaic plot are also updated.
 Also specific nodes can be selected to update the plots.
@@
 The third tab (Figure \ref{tab3}) examines the PPF fit and compares the result with a RF fit.
-There are four displays for each type of model: (1) Variable importance for all trees in the forest (same as in the models tab), (2) a receiver operating characteristic curve (ROC) comparing sensitivity and specificity for each class, (3) oob error by the number of trees to assess complexity, (4) overall variable importance.
+There are four displays for each type of model: (1) Variable importance for all trees in the forest (same as in the models tab), (2) a receiver operating characteristic curve (ROC) comparing sensitivity and specificity for each class, (3) OOB error by the number of trees to assess complexity, (4) overall variable importance.
 There is very little interaction on this tab.
@@
 Being able to focus on class can help to better understand how well the model performs across classes and the focus can be especially useful for unbalanced data.
-Examining the oob error by trees enables an assessment of how few trees might be used to provide an equally accurate prediction of future data.
+Examining the OOB error by trees enables an assessment of how few trees might be used to provide an equally accurate prediction of future data.
 
@@
 \caption{Performance comparison tab of the web app.
-ROC curves displaying sensitivity against specificity for the classes are shown, along with the oob error by the number of trees used to build the forest, and overall variable importance.
+ROC curves displaying sensitivity against specificity for the classes are shown, along with the OOB error by the number of trees used to build the forest, and overall variable importance.
 Displays are shown for the PPF and RF, for comparison purposes.
@@
 Successful apps make it easier for the user to do common tasks, and this is the design principal here.
-Only limited functionality is provided to the user, but the code is available and a knowledgeable data analyst might use this as a basis for extending the methods as appropriate for theit data.
+Only limited functionality is provided to the user, but the code is available and a knowledgeable data analyst might use this as a basis for extending the methods as appropriate for their data.
 Performance of the app is good for relatively small data sets.
@@
 \item Classes as a whole could be aggregated or re-organised as suggested by the model diagnostics, to produce a more effective hierarchical approach to the multiple class problem. 
-
+\end{enumerate}
 \noindent Working within the R environment makes all of these possible when using command line outside the app, given that the unique IDs of models and cases can be exported from the app.
@@
 
+\section{Supplementary Materials}
+This article was written with the R packages knitr \citep{xie:2015}, ggplot2 \citep{hadley:2009}, and dplyr \citep{dplyr}, and the files to reproduce the article and results is available at https://github.com/natydasilva/PPforestViz.
+The code to reproduce the shiny app is available at https://github.com/natydasilva/shinyPPforest.
+
 \bibliographystyle{bibstyle}

### 2022-06-30 response_reviewers.Rmd
update some figure numbers and typos

@@
 \textcolor{violet}{As we mentioned in page 7 the ternary plot can be generalized to more than three dimensional space, with G classes, the ternary plot idea is generalized to a G-1 dimensional simplex.
-Then if we have 4 classes the generalized ternary plot is a tetrahedron as shown in Figure 4 but if we have more than four classes we can also generalize the ternary plot idea.
-In the fishcatch data used in this paper to illustrate the shiny app there are 7 classes and we generalized the ternary plot to a 6-dimensional simplex as is shown in the last panel of Figure 6 (6-didmensional simplex shown pairwise), using tourr package you can see better the 6-dimensional simplex in different rotations https://vimeo.com/707047875.
-The structure is complex because a 6-dimensional simplex has 7 vertices, 21 eddges and 35 faces.}
+Then if we have 4 classes the generalized ternary plot is a tetrahedron as shown in Figure 5 but if we have more than four classes we can also generalize the ternary plot idea.
+In the fishcatch data used in this paper to illustrate the shiny app there are 7 classes and we generalized the ternary plot to a 6-dimensional simplex as is shown in the last panel of Figure 8 (6-didmensional simplex shown pairwise), using tourr package you can see better the 6-dimensional simplex in different rotations https://vimeo.com/707047875.
+The structure is complex because a 6-dimensional simplex has 7 vertices, 21 edges and 35 faces.}
 
@@
 
-\textcolor{violet}{In Figure 4 the dimension of the simplex is $G-1=3$, since in crab data there are 4 classes.
+\textcolor{violet}{he dimension of the simplex is $G-1=3$ (Figure 5 revised version), since in crab data there are 4 classes.
 Ternary plot can be generalized to more than three dimensional -- with $G>4$ classes, the ternary plot idea is generalized to a $G-1$ dimensional simplex.
@@
 \textcolor{violet}{Yes, fishcatch data are more difficult to classify and then more interesting to show some of the results.
-For simplicity to describe the main plot of the app we use crab data but to describe the app we prefer to present a more complex problem, more classes, and imbalanced data.}
+For simplicity to describe the main plot of the app we use crab data but to describe the app we prefer to present a more complex problem, more classes, and umbalanced data.}
 
@@
 In the MDS we only used two dimensions but can be more if deemed more appropriate for a particular problem.
-We have included labels in the generalized ternary plot as we did in Figure 7.}
+We have included labels in the generalized ternary plot.}
 

### 2022-10-04 PPforestVizupdate.Rnw
second revision

@@
 library(ggmosaic)
+library(patchwork)
 set.seed(123)
@@
 \centering
-<<ternary, echo=FALSE, warning=FALSE, message=FALSE, fig.height=4, fig.width=7.5, out.extra='trim=0cm 2cm 0cm 2cm, clip'>>=
+<<ternary, echo=FALSE, warning=FALSE, message=FALSE, fig.height=4, fig.width=7.5>>=
 f.helmert <- function(d)
@@
 ##ternary plot
-ternaryshell <- function(ppf, sp = 3, dx = 1, dy = 2, v1=1, v2=2, v3=3){
+ternaryshell <- function(ppf, sp = 3, dx = 1, dy = 2, v1=1, v2=2, v3=3, leg=FALSE){
+
   s <- simplex(sp)
@@
   gg1 <- ternarydata(ppf, v1, v2, v3)
+  levels(gg1$Class) <- c("BF", "BM", "OF", "OM")
 
@@
                     y1=pts[,dy][s$edg[,1]], y2=pts[,dy][s$edg[,2]])
-
+if(leg==FALSE){
   p1  <- gg1 %>% filter(pair %in% paste(dx,dy, sep="-") ) %>%
@@
     theme(aspect.ratio=1)
-
+}else{
+    p1  <- gg1 %>% filter(pair %in% paste(dx,dy, sep="-") ) %>%
+    select(Class, x, y) %>%
+    ggplot(aes(x, y, color = Class)) +
+      geom_segment(data = edg, aes(x = x1, xend = x2,
+                                 y = y1, yend = y2), color = "black" ) +
+      geom_point(size = I(3), alpha = .5) +
+      labs(y = "",  x = "") +
+      theme(legend.position = "bottom", aspect.ratio = 1) +
+      scale_colour_brewer(type = "qual", palette = "Dark2") +
+      labs(x = paste0("T",dx,""), y = paste0("T",dy,"")) +
+    theme(aspect.ratio=1)
+}
   p1
 }
-t1 <- ternaryshell(ppf, 3, 1, 2)
-t2 <- ternaryshell(ppf, 3, 1, 3)
-t3 <- ternaryshell(ppf, 3, 2, 3)
-
-grid.arrange(t1, t2, t3, ncol = 3)
-
+t1 <- ternaryshell(ppf, 3, 1, 2,leg=TRUE)
+t2 <- ternaryshell(ppf, 3, 1, 3, leg=TRUE)
+t3 <- ternaryshell(ppf, 3, 2, 3,leg=TRUE)
+
+t1+t2+t3+plot_layout(guides = "collect") &
+  theme(legend.position='bottom')
 #vimeoternary fish code
@@
 \centering
-\includegraphics[width=1\linewidth]{figure/fish11.png}
+\includegraphics[width=1.05\linewidth]{figure/fish11.png}
 \caption{Entry page for the web app, focusing on examining cases, in terms of similarity and uncertainty.
@@
 \centering
-\includegraphics[width=1\linewidth]{figure/tab1com2.png}
+\includegraphics[width=1.1\linewidth]{figure/tab1com2.png}
 \caption{Entry page for the web app, focusing on examining cases, for selected points.  \label{tab1comp}}
@@
 \centering
-\includegraphics[width=1\linewidth]{figure/tab2comp.png}
+\includegraphics[width=1.1\linewidth]{figure/tab2comp.png}
 \caption{The individual model tab in the web app.
@@
 \centering
-\includegraphics[width=1\linewidth]{figure/fishT2.pdf}
+\includegraphics[width=1.05\linewidth]{figure/fishT2.pdf}
 \caption{Schematic diagram illustrating the interactivity in and between plots for model level exploration panel of the web app.
@@
 Some of the proposed reduced the multi-class problem to a set of binary problems.
-The approach used for a multi-class ROC analysis in this article is called one-against-all ~\citep{allwein2000reducing}.
+The approach used for a multi-class ROC analysis in this article is called one-against-all \citep{allwein2000reducing}.
 
 \begin{figure}[hbpt]
-\includegraphics[width=.8\linewidth]{figure/fish31.png}
+\includegraphics[width=1\linewidth]{figure/fish31.png}
 \caption{Performance comparison tab of the web app.
@@
 
-\bibliographystyle{bibstyle}
+\bibliographystyle{apalike}
 \bibliography{ppforestpaperbib}

### 2022-10-11 response_reviewers._July22.Rmd
update

--- 2022-06-30 response_reviewers.Rmd
@@
 
-Our responses to the review are written in \textcolor{violet}{violet}.
+Our responses to the review are written in \textcolor{blue}{\textbf{blue}}.
 
@@
 
-\textcolor{violet}{Thank you. We appreciate the chance to do a further revision, and hope that these changes are satisfactory.}
+\textcolor{blue}{\textbf{Thank you. We appreciate the chance to do a further revision, and hope that these changes are satisfactory.}}
 
@@
 
-\textcolor{violet}{We appreciate all of your feedback which help to improve our paper. 
- }
+\textcolor{blue}{\textbf{We appreciate all of your feedback which help to improve our paper. 
+ }}
 
@@
 In particular, the app focuses on the first three nodes of the trees in question, which would be much less informative in settings requiring deep trees.
-\textcolor{violet}{The app was modified to adress this issue.
+\textcolor{blue}{\textbf{The app was modified to adress this issue.
 In the new application the user can select specific nodes to see (from one to all the nodes).
-If there are a lot of nodes will be better to select 4 or less each time to see better the complete display but the application is not limited to see just the first three nodes.}
+If there are a lot of nodes will be better to select 4 or less each time to see better the complete display but the application is not limited to see just the first three nodes.}}
 
@@
 
-\textcolor{violet}{The discussion has been extended to include limitations.}
+\textcolor{blue}{\textbf{The discussion has been extended to include limitations.}}
 
@@
 
-\textcolor{violet}{As we mentioned in page 7 the ternary plot can be generalized to more than three dimensional space, with G classes, the ternary plot idea is generalized to a G-1 dimensional simplex.
+\textcolor{blue}{\textbf{As we mentioned in page 7 the ternary plot can be generalized to more than three dimensional space, with G classes, the ternary plot idea is generalized to a G-1 dimensional simplex.
 Then if we have 4 classes the generalized ternary plot is a tetrahedron as shown in Figure 5 but if we have more than four classes we can also generalize the ternary plot idea.
 In the fishcatch data used in this paper to illustrate the shiny app there are 7 classes and we generalized the ternary plot to a 6-dimensional simplex as is shown in the last panel of Figure 8 (6-didmensional simplex shown pairwise), using tourr package you can see better the 6-dimensional simplex in different rotations https://vimeo.com/707047875.
-The structure is complex because a 6-dimensional simplex has 7 vertices, 21 edges and 35 faces.}
+The structure is complex because a 6-dimensional simplex has 7 vertices, 21 edges and 35 faces.}}
 
@@
 
-\textcolor{violet}{We have included more explanaition about the parallel coordinate plot in Section 4 where this figure appears and added a corresponding reference for further details.
-In the vertical axis the values for each variable are standardized to easily compare them.}
+\textcolor{blue}{\textbf{We have included more explanaition about the parallel coordinate plot in Section 4 where this figure appears and added a corresponding reference for further details.
+In the vertical axis the values for each variable are standardized to easily compare them.}}
 
@@
 
-\textcolor{violet}{The second tab was modified based on your helpful comments.
+\textcolor{blue}{\textbf{The second tab was modified based on your helpful comments.
 We have included three main changes in the second tabset, first we have included jittered points in the OOB error boxplot where the user can select trees based in their performance.
@@
 The panel allows you to select more or less than three nodes and the mentioned plots are updated acordingly.
-Finally, we rearranged some of the plots to make the selection and interaction more intuitively presented in Figure 10.}
+Finally, we rearranged some of the plots to make the selection and interaction more intuitively presented in Figure 10.}}
 
@@
 
-\textcolor{violet}{We explore different alternatives as you suggested but the result was not clearly better.
+\textcolor{blue}{\textbf{We explore different alternatives as you suggested but the result was not clearly better.
 We have removed the points and just keep thiner lines which improve the visualization.
-Also in the app, because the interactivity, it is easy to see the oob error values mousing over the plot which in the static version it is not possible.}
+Also in the app, because the interactivity, it is easy to see the oob error values mousing over the plot which in the static version it is not possible.}}
 
@@
 
-\textcolor{violet}{I have removed $G_s$, as you notices it is unnecessary here}
+\textcolor{blue}{\textbf{I have removed $G_s$, as you notices it is unnecessary here}}
 
@@
 
-\textcolor{violet}{Done}
+\textcolor{blue}{\textbf{Done}}
 
@@
 
-\textcolor{violet}{Class color legends were not included in the interactive web app because the interactivity allows you to get the class name when you mouse over each plot.}
+\textcolor{blue}{\textbf{Class color legends were not included in the interactive web app because the interactivity allows you to get the class name when you mouse over each plot.}}
 
@@
 
-\textcolor{violet}{In Section 4 we have added more details and examples of use and interpreting results to address your comment.}
+\textcolor{blue}{\textbf{In Section 4 we have added more details and examples of use and interpreting results to address your comment.}}
 
@@
 
-\textcolor{violet}{A random forest has two sources of randomness: variable sampling and bootstrap samples.
+\textcolor{blue}{\textbf{A random forest has two sources of randomness: variable sampling and bootstrap samples.
 The variable sampling is used for node partition, where each split is produced from a random subset of variables.
-There is not an initial sample of variables for each tree.}
+There is not an initial sample of variables for each tree.}}
 
-\textcolor{violet}{We did not formally explore the sensitivity of the random variable sample size for the measure of variable importance.
+\textcolor{blue}{\textbf{We did not formally explore the sensitivity of the random variable sample size for the measure of variable importance.
 In the algorithm you can select the proportion of variables to be sampled at each node.
-In the crab and fishcatch examples presented in the paper two variables were sampled at each node.}
+In the crab and fishcatch examples presented in the paper two variables were sampled at each node.}}
 
-\textcolor{violet}{We have included an image from the original paper with the boundary that would result from a classification tree fitted using the PPtree algorithm to take into account your last request.
+\textcolor{blue}{\textbf{We have included an image from the original paper with the boundary that would result from a classification tree fitted using the PPtree algorithm to take into account your last request.
 The plot shows a simple example from simulated data with two features and three classes.
-The partitions generated by the PPtree algorithm are oblique to the axis, incorporating the association between the two variables.}
+The partitions generated by the PPtree algorithm are oblique to the axis, incorporating the association between the two variables.}}
 
@@
 
-\textcolor{violet}{Subscript j for $O_k$ is not needed because for every variable j $O_k$ will be the same and not change depending on j, the same for $\hat{y}_{ik}$.
-In the only case subscript $j$ can be included is in  $\tilde y_{ik}$ to indicate the permuted variable but for simplicity we just keep the formula as it is in the original PPforest paper (https://www.tandfonline.com/doi/abs/10.1080/10618600.2020.1870480) }
+\textcolor{blue}{\textbf{Subscript j for $O_k$ is not needed because for every variable j $O_k$ will be the same and not change depending on j, the same for $\hat{y}_{ik}$.
+In the only case subscript $j$ can be included is in  $\tilde y_{ik}$ to indicate the permuted variable but for simplicity we just keep the formula as it is in the original PPforest paper (https://www.tandfonline.com/doi/abs/10.1080/10618600.2020.1870480) }}
 
@@
 But in the upper right corner, the plots of the variables distributions, for instance for node 2, are not in the standardized range (node 2 has an x-axis that goes up to 5 and 6).
-\textcolor{violet}{As you mentioned the data should be standardized for the coefficients to be interpretable as importances. We have fixed the density plot and the x axis accordingly.}
+\textcolor{blue}{\textbf{As you mentioned the data should be standardized for the coefficients to be interpretable as importances. We have fixed the density plot and the x axis accordingly.}}
 
 Is the distribution being plotted here the value of the linear combination for each class, rather than a single variable? 
-\textcolor{violet}{Yes, the density plots shows the data projection at each node. This is now made clear in the caption.}
+\textcolor{blue}{\textbf{Yes, the density plots shows the data projection at each node. This is now made clear in the caption.}}
 
 Is there something going on here with the default x limits for doing KDE? 
-\textcolor{violet}{The x limits were fixed.}
+\textcolor{blue}{\textbf{The x limits were fixed.} }
 
@@
 
-\textcolor{violet}{We explore your suggestion (changing the bound for the KDE to observe range) but the result was not better so we have kept the original.}
+\textcolor{blue}{\textbf{We explore your suggestion (changing the bound for the KDE to observe range) but the result was not better so we have kept the original.} }
 
@@
 
-\textcolor{violet}{The selected palette (Dark2) is a qualitative palette suited to mapping a categorical data because it is colorblind friendly.
+\textcolor{blue}{\textbf{The selected palette (Dark2) is a qualitative palette suited to mapping a categorical data because it is colorblind friendly.
 It is a little unfortunate that the categories in the crabs data are colors, which makes the explanation confusing.
-We have changed the category names to be simply O and B so that there is less confusion.}
+We have changed the category names to be simply O and B so that there is less confusion.}}
 
 Also, can you include labeling in your plot that the sub-plots are node ID (e.g., put the label "node" above the 3 density plots).
-\textcolor{violet}{We have included in each sub-plots "Node" before the number and also update this in the app.}
+\textcolor{blue}{\textbf{We have included in each sub-plots "Node" before the number and also update this in the app.}}
 
-What do the numbers in the grey boxes at the terminal nodes represent? \textcolor{violet}{The numbers in the gray boxes are the predicted class, we fixed the plot to show the class names}
+What do the numbers in the grey boxes at the terminal nodes represent? \textcolor{blue}{\textbf{The numbers in the gray boxes are the predicted class, we fixed the plot to show the class names}
+}
 
@@
 
-\textcolor{violet}{Yes, the coefficient values are between -1 and 1 by construction.
+\textcolor{blue}{\textbf{Yes, the coefficient values are between -1 and 1 because the length of the projection vector is 1.
 We find the linear combinations, \textbf{a} which maximize $\frac{a^TBa}{a^TWa}$, where $B$ and $W$ are the between and within group sums of squares.
@@
 We have included in Section 3 LDA and PDA indexes which are optimized to get the linear projections to make it more clear how the coefficients are computed.
-}
+}}
 
@@
 
-\textcolor{violet}{Formulas for all VI were defined in equations and mentioned in the text. It is clearer now.}
+\textcolor{blue}{\textbf{Formulas for all VI were defined in equations and mentioned in the text. It is clearer now.}}
 
@@
 
-\textcolor{violet}{A reference for MDS has been included.
+\textcolor{blue}{\textbf{A reference for MDS has been included.
 In Section 3 we describe the mapping of diagnostics to visualizations.
 MDS is presented in Subsection 3.3 (Similarity of cases) the key connection with the forest is that we are using the proximity matrix information which is one of the diagnostics described for PPforest or RF and we use this information to compute the MDS.
-This visualization can be useful to identify in which cases the model is performing well and which it is having problems.}
+This visualization can be useful to identify in which cases the model is performing well and which it is having problems.}}
 
-\textcolor{violet}{MDS and generalized ternary plot (tetrahedral plot in case there are  four clases) are both implemented in the app and we have added a comment in the paper to clarify this.}
+\textcolor{blue}{\textbf{MDS and generalized ternary plot (tetrahedral plot in case there are  four clases) are both implemented in the app and we have added a comment in the paper to clarify this.}}
 
@@
 
-\textcolor{violet}{We selected a jittered side-by-side dot plot in this case instead of a box plot because we want to see the values for each tree, and it makes it more sensible when linking observations between multiple plots.
-Jittering spreads the data on the categorical variable axis to better read the distribution.}
+\textcolor{blue}{\textbf{We selected a jittered side-by-side dot plot in this case instead of a box plot because we want to see the values for each tree, and it makes it more sensible when linking observations between multiple plots.
+Jittering spreads the data on the categorical variable axis to better read the distribution.}}
 
@@
 
-\textcolor{violet}{he dimension of the simplex is $G-1=3$ (Figure 5 revised version), since in crab data there are 4 classes.
+\textcolor{blue}{\textbf{he dimension of the simplex is $G-1=3$ (Figure 5 revised version), since in crab data there are 4 classes.
 Ternary plot can be generalized to more than three dimensional -- with $G>4$ classes, the ternary plot idea is generalized to a $G-1$ dimensional simplex.
@@
 The structure is complex because a 6-dimensional simplex has 7 vertices, 21 eddges and 35 faces.
-This has been clarified in the paper.}
+This has been clarified in the paper.} }
 
@@
 
-\textcolor{violet}{We are using points because it is important to see the individual cases and the jitter is to overcome overplotting on the categorical variable.
-It is not recommended using jittered scatterplots because it would change the real values of your data.}
+\textcolor{blue}{\textbf{We are using points because it is important to see the individual cases and the jitter is to overcome overplotting on the categorical variable.
+It is not recommended using jittered scatterplots because it would change the real values of your data.}}
 
@@
 
-\textcolor{violet}{At their core Shiny widgets are mini-applications created using the shinyApp function. We have changed the language of the paragraph to help with this terminology.}
+\textcolor{blue}{\textbf{At their core Shiny widgets are mini-applications created using the shinyApp function. We have changed the language of the paragraph to help with this terminology.}}
 
@@
 
-\textcolor{violet}{Yes, fishcatch data are more difficult to classify and then more interesting to show some of the results.
-For simplicity to describe the main plot of the app we use crab data but to describe the app we prefer to present a more complex problem, more classes, and umbalanced data.}
+\textcolor{blue}{\textbf{Yes, fishcatch data are more difficult to classify and then more interesting to show some of the results.
+For simplicity to describe the main plot of the app we use crab data but to describe the app we prefer to present a more complex problem, more classes, and umbalanced data.}}
 
@@
 
-\textcolor{violet}{A PCP can be ordered in different ways, and we have modified the PCP to include variable order based on the mean value.
+\textcolor{blue}{\textbf{A PCP can be ordered in different ways, and we have modified the PCP to include variable order based on the mean value.
 As you suggested, we have also change the grid lines to keep only the vertical grid in gray.
 In the MDS we only used two dimensions but can be more if deemed more appropriate for a particular problem.
-We have included labels in the generalized ternary plot.}
+We have included labels in the generalized ternary plot.}}
 
@@
 
-\textcolor{violet}{This comment is in page 12, In the original RF there is not an option to prune under-performing trees.
-The comment is just to mention a possible use for the information we get from the app, also we are presenting the app using PPforest and this can be a possible extension for our algorithm.}
+\textcolor{blue}{\textbf{This comment is in page 12, In the original RF there is not an option to prune under-performing trees.
+The comment is just to mention a possible use for the information we get from the app, also we are presenting the app using PPforest and this can be a possible extension for our algorithm.}}
 
@@
 
-\textcolor{violet}{In the upper LH corner we are using the function `selectizeInput()` from shiny R package which creates a select list that can be used to choose a single or multiple items from a list of values, you do not have to type the class names you just select the names from a list.
-For the bottom two plots (variable importance plots) we have fixed the axis and include titles in each plot, so it is easier to compare them.}
+\textcolor{blue}{\textbf{In the upper LH corner we are using the function `selectizeInput()` from shiny R package which creates a select list that can be used to choose a single or multiple items from a list of values, you do not have to type the class names you just select the names from a list.
+For the bottom two plots (variable importance plots) we have fixed the axis and include titles in each plot, so it is easier to compare them.}}
 
@@
 
-\textcolor{violet}{We have changed to using OOB now.}
+\textcolor{blue}{\textbf{We have changed to using OOB now.}}
 
@@
 
-\textcolor{violet}{Cases that are predicted to be multiple classes, without a clear majority in any of them, indicate that those cases are difficult to classify.
-The paragraph was not clear and we modify it to:" Cases that are predicted to be multiple classes, with similar class probabilities, indicate difficult to classify observations"  }
+\textcolor{blue}{\textbf{Cases that are predicted to be multiple classes, without a clear majority in any of them, indicate that those cases are difficult to classify.
+The paragraph was not clear and we modify it to:" Cases that are predicted to be multiple classes, with similar class probabilities, indicate difficult to classify observations".
+}}
 
@@
 
-\textcolor{violet}{Yes, should be Thus. It is fixed.}
+\textcolor{blue}{\textbf{Yes, should be Thus. It is fixed.}}
 

### 2022-10-11 response_reviewers_Oct22.Rmd
update

---
title: "Authors Response to Second Review"
output: pdf_document
header-includes:  \usepackage{xcolor}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Our responses to the review are written in \textcolor{blue}{\textbf{blue}}.

\textcolor{blue}{\textbf{Thank you. We appreciate the chance to do a further revision, and hope that these changes are satisfactory.}}

## Response to Guest Editors

Address the remaining comments from both reviewers as far as possible. If something cannot be done easily, suggest this for future work.

- Add a color legend to Fig. 5 (and others - as needed).

\textcolor{blue}{\textbf{The color legend for Figure 5 was included.
Also in the schematic diagrams, Figure 7 and Figure 11, we have included one color legend as reference for all the plots at bottom of each diagram.
In Section 4 (Interactive Web App) there are some Figures without color legend because are images from the web app available at https://natydasilva.shinyapps.io/shinyppforest/ and as I mentioned in previous review Class color legends were not included in the interactive web app because the interactivity allows you to get the class name when you mouse over each plot and will be redundant to include them on the shinyapp.}}

- Enlarge all figures to full page width, e.g., Fig.12 \textcolor{blue}{\textbf{Done}}

Please make a final review to detect typos, e.g., Line 18 on page 13. It seems to be that “… it …” should be “… its …”.\textcolor{blue}{\textbf{Done}}

Please review reference style. Carefully review and adjust all your references. There are numerous problems:

\textcolor{blue}{\textbf{The references were modified as you have suggested}}

- Capitalize all journal names \textcolor{blue}{\textbf{Done}}
- Add missing volume, issue, and page numbers \textcolor{blue}{\textbf{Done}}
- Simplify the publisher info for Inselberg to "Springer, New York, NY" \textcolor{blue}{\textbf{Done}}
- Extend "et-al" author lists and list all co-authors 
\textcolor{blue}{\textbf{Done}}
- Fill in missing information, e.g,. for Quach (what is this & where can it be obtained)
- Same for Sievert (2017)\textcolor{blue}{\textbf{Done}}
- Include the location (city, state) for all books & book chapters \textcolor{blue}{\textbf{Done}}
- Remove ISBN numbers and series information (e.g., useR) from books \textcolor{blue}{\textbf{Done}}
- Capitalize R and other proper names \textcolor{blue}{\textbf{Done}}
- Remove months (e.g., in Wickham 2009) \textcolor{blue}{\textbf{Done}}
- Use a simplified URL for Puranen: http://jse.amstat.org/jse_data_archive.htm
\textcolor{blue}{\textbf{Done}}

## Response to Reviewer 2

The updated version of your paper addresses all of the concerns that I brought up in my original review. Both the added functionality of the package (e.g., flexibility in selecting tree nodes for exploration) and the explicit discussion of limitations will hopefully prove useful to the users of your package and readers of Computational Statistics. Moreover, the current version of the paper is cleaner and easier to read thanks to the edits and minor fixes (e.g., typos).

\textcolor{blue}{\textbf{We appreciate all of your feedback which help to improve our paper.
}}

The only remaining minor comment that I have is that the side-by-side jittered dotplot in Figure 7 seems to be based on the crabs dataset instead of the fishcatch dataset used in other plots in that figure.

\textcolor{blue}{\textbf{Yes you are right, we have changed the side-by-side jittered dotplot with fishcatch data and include a general color legend as reference.}}

## Response to Reviewer 4

The paper looks better now. Regarding the points I raised in the comments about the dot plots with jittering across the x-axis when the x-axis is categorical, I still think you might want to consider an alternative. I think in many cases, the overplotting of the points makes the plots very cluttered, and in reality you only need to plot the individual points for a few observations. I understand that in this case there is the application of being able to select points and connect between plots. However, I think that this will mostly be of interest for the most extreme points. Here is a suggestion: perhaps in the future the default view (e.g. Fig 3, importance plot in Fig 10), you could use a violin plot with only the most extreme (lowest and/or highest quantiles, with a check to plot only the extreme values that would not coincide with a "fat" point of the violin plot) values plotted (see eg. https://r-charts.com/distribution/violin-plot-points-ggplot2/). There could be a button to toggle on/off the option to see all points as a scatterplot rather than violin plot. In the cases of say Figure 3 and 10 where you want to hilight one of the points in red corresponding to a selected tree, you could only plot that point (rather than all) in red (in addition to the outliers) when it is selected, but not all the time.
Perhaps you could also do something similar for the side by side plots (e.g. Fig 8) where you have violin plots by color rather than the points (or at least have the option to do so if there are few colors).

\textcolor{blue}{\textbf{ Thank you for your comments, based on your suggestions I have explored two alternatives: the violin plot and also a boxplot without whiskers (just the interquartile range) adding the points  using transparency. At least for the case examples we are including in the paper the result was not clearly better For the fishcatch example the importance have an structure (groups of specific values for each variable ) and when we use violin plot that structure get lost. 
The best solution  could be to include a button to toggle on/off  the violin option or other, this will introduce an additional level of reactivity in the second tab of the application. The second tab will be more complex because there will be a lot of reactivities to control (error tree in the boxplot, node selection, importance variable values and finally plot options for the importance variable)  and it is not clear how to order the reactivities sequence. This is something for sure to explore in the future.
I have included a comment about alternative visualizations for importance plot after Figure 3.
}}

### 2022-10-11 COST_submission_July22/PPforestViz_journal.tex
update

--- 2022-06-30 PPforestVizupdate.Rnw
@@
-\documentclass[smallextended,natbib]{svjour3}  
+\documentclass[smallextended,natbib]{svjour3}\usepackage[]{graphicx}\usepackage[]{xcolor}
+% maxwidth is the original width if it is less than linewidth
+% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\makeatletter
+\def\maxwidth{ %
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
+\makeatother
+
+\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
+\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
+\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
+\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
+\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
+\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
+\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
+\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
+\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
+\let\hlipl\hlkwb
+
+\usepackage{framed}
+\makeatletter
+\newenvironment{kframe}{%
+ \def\at@end@of@kframe{}%
+ \ifinner\ifhmode%
+  \def\at@end@of@kframe{\end{minipage}}%
+  \begin{minipage}{\columnwidth}%
+ \fi\fi%
+ \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
+ \colorbox{shadecolor}{##1}\hskip-\fboxsep
+     % There is no \\@totalrightmargin, so:
+     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
+ \MakeFramed {\advance\hsize-\width
+   \@totalleftmargin\z@ \linewidth\hsize
+   \@setminipage}}%
+ {\par\unskip\endMakeFramed%
+ \at@end@of@kframe}
+\makeatother
+
+\definecolor{shadecolor}{rgb}{.97, .97, .97}
+\definecolor{messagecolor}{rgb}{0, 0, 0}
+\definecolor{warningcolor}{rgb}{1, 0, 1}
+\definecolor{errorcolor}{rgb}{1, 0, 0}
+\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
+
+\usepackage{alltt}  
 \smartqed  % flush right qed marks, e.g. at end of proof
@@
 \newcommand{\blalpha}{\mbox{\boldmath {\alpha}}}
-
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}
@@
 
-<<pkg, echo=FALSE, message=FALSE, warning=FALSE>>=
-library(ggplot2)
-library(PPforest)
-library(shiny)
-library(GGally)
-library(productplots)
-library(PPtreeViz)
-library(tidyr)
-library(dplyr)
-library(pracma)
-library(RColorBrewer)
-library(randomForest)
-library(pROC)
-library(gridExtra)
-library(grid)
-library(ggmosaic)
-set.seed(123)
-#  ppf2022 <-PPforest(data = crab, class= "Type", std = TRUE, size.tr = 1 , m = 500, PPmethod = 'LDA', size.p =.33,
-#   lambda = .1, parallel = FALSE, cores = 2, rule = 1)
-# save(ppf2022, file = "ppf2022.RData")
-load("ppf2022.RData")
-load("rf_fish.Rdata")
- ppf<- ppf2022
-@
-
 \begin{figure}[!t]
-<<boundss, cache=FALSE, echo=FALSE, fig.height=5, fig.width=8, message=FALSE>>=
-grid_arrange_shared_legend <- function(..., ncol = length(list(...)), nrow = 1, position = c("bottom", "right")) {
-
-  plots <- list(...)
-  position <- match.arg(position)
-  g <- ggplotGrob(plots[[1]] + theme(legend.position = position))$grobs
-  legend <- g[[which(sapply(g, function(x) x$name) == "guide-box")]]
-  lheight <- sum(legend$height)
-  lwidth <- sum(legend$width)
-  gl <- lapply(plots, function(x) x + theme(legend.position="none"))
-  gl <- c(gl, ncol = ncol, nrow = nrow)
-
-  combined <- switch(position,
-       "bottom" = arrangeGrob(do.call(arrangeGrob, gl),
-                   legend,
-                   ncol = 1,
-                   heights = unit.c(unit(1, "npc") - lheight, lheight)),
-       "right" = arrangeGrob(do.call(arrangeGrob, gl),
-                   legend,
-                   ncol = 2,
-                   widths = unit.c(unit(1, "npc") - lwidth, lwidth)))
-
-  grid.draw(combined)
-}
-
-simu3 <- function(mux1, mux2, muy1, muy2, muz1, muz2,
-                  cor1, cor2, cor3) {
-  bivn <- MASS::mvrnorm(100, mu = c(mux1, mux2),
-                  Sigma = matrix(c(1, cor1, cor1, 1), 2))
-  bivn2 <- MASS::mvrnorm(100, mu = c(muy1, muy2),
-                   Sigma = matrix(c(1, cor2, cor2, 1), 2))
-  bivn3 <- MASS::mvrnorm(100, mu = c(muz1, muz2),
-                   Sigma = matrix(c(1, cor3, cor3, 1), 2))
-
-  d1 <- data.frame(Sim = "sim1", bivn, stringsAsFactors = TRUE)
-  d2 <- data.frame(Sim = "sim2", bivn2, stringsAsFactors = TRUE)
-  d3 <- data.frame(Sim = "sim3", bivn3, stringsAsFactors = TRUE)
-  rbind(d1, d2, d3)
-}
-
-set.seed(666)
-dat.pl2 <- simu3(-1,0.6,0,-0.6, 2,-1,0.95, 0.95, 0.95)
-dat.plnew <- simu3(0,0.6,0,-0.6, 0,-1,-0.1, -0.1, -0.1)
-
-grilla <- expand.grid(X1 = seq(-4,4.8,,100),
-                            X2 = seq(-4.3,3.3,,100))
-
-pptree <- PPtreeViz::PPTreeclass(Sim~., data = dat.pl2, "LDA")
-ppred.sim <- PPtreeViz::PPclassify(pptree, test.data = grilla,
-                                   Rule = 6)
-grilla$ppred<-ppred.sim[[2]]
-
-rpart.crab <- rpart::rpart(Sim ~ X1 + X2, data = dat.pl2)
-rpart.pred <- predict(rpart.crab, newdata = grilla, type="class")
-
-bnds <- data.frame(
-  a = c(pptree$splitCutoff.node[[1]] / pptree$projbest.node[[3]],
-      pptree$splitCutoff.node[[2]] / pptree$projbest.node[[4]]),
-  b = c(-pptree$projbest.node[[1]] / pptree$projbest.node[[3]],
-      -pptree$projbest.node[[2]] / pptree$projbest.node[[4]]))
-p <- ggplot(data = grilla ) +
-       geom_point(aes(x = X1, y = X2,
-                  color = as.factor(ppred),
-                  shape=as.factor(ppred)),
-                  alpha = .20) +
-      scale_colour_brewer(name="Class", type="qual",
-                      palette = "Dark2") +
-
-      scale_shape_discrete(name = 'Class') +
-      theme_bw() +
-      theme(aspect.ratio = 1, legend.position="none")
-
-pl.pp <- p + geom_point(data = dat.pl2,
-                        aes(x = X1, y = X2,
-                            group = Sim,
-                            shape = Sim, color=Sim), size = 3) +
-  scale_y_continuous(expand = c(0,0)) +
-  scale_x_continuous(expand = c(0,0))
-
-pl.pp
-
-@
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+\includegraphics[width=\maxwidth]{boundss-1} 
+\end{knitrout}
  \vspace*{-0.3cm}
@@
 \centering
-<<trcomb, echo=FALSE, warning=FALSE, message=FALSE, fig.height=4, fig.width=6.5, results='hide', error=FALSE>>=
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+\includegraphics[width=\maxwidth]{trcomb-1}
 
-tree <- 11
-
-#Data trans density and mosaic
-dat_dens <- function(PPclassOBJ, node.id, Rule, legend = TRUE, std = TRUE,
-                     image = FALSE, diff.prop = 0.2,c1=FALSE) {
-
-  searchGroup <- function(node.id, TS, gName) {
-    flag <- TRUE
-    sel.id <- TS[node.id, 2:3]
-    LR.id <- c(TRUE, FALSE)
-    sel.group <- NULL
-    i <- 1
-    while ((sel.id[i] != 0) && (i < length(sel.id))) {
-      if (TS[sel.id[i], 2] != 0) {
-        sel.id <- c(sel.id, TS[sel.id[i], 2:3])
-        if (LR.id[i])
-          LR.id <- c(LR.id, c(TRUE, TRUE))
-        else LR.id <- c(LR.id, c(FALSE, FALSE))
-      }
-      if (TS[sel.id[i + 1], 2] != 0) {
-        sel.id <- c(sel.id, TS[sel.id[i + 1], 2:3])
-        if (LR.id[i + 1])
-          LR.id <- c(LR.id, c(TRUE, TRUE))
-        else LR.id <- c(LR.id, c(FALSE, FALSE))
-      }
-      i <- i + 2
-    }
-    sel.Name <- TS[sel.id[which(TS[sel.id, 2] == 0)], 3]
-    selName <- sort(gName[sel.Name])
-    L.list <- sort(gName[sel.Name[LR.id[which(TS[sel.id,
-                                                 2] == 0)]]])
-    R.list <- sort(gName[sel.Name[!LR.id[which(TS[sel.id,
-                                                  2] == 0)]]])
-    return(list(selName = selName, Llist = L.list, Rlist = R.list))
-  }
-
-  TS <- PPclassOBJ$Tree.Struct
-  Alpha <- PPclassOBJ$projbest.node
-  cut.off <- PPclassOBJ$splitCutoff.node
-  origdata <- PPclassOBJ$origdata
-  origclass <- PPclassOBJ$origclass
-  p <- ncol(origdata)
-  gName <- names(table(origclass))
-
-  if (TS[node.id, 2] != 0) {
-    SG.result <- searchGroup(node.id, TS, gName)
-    selG <- SG.result$selName
-    selL <- SG.result$Llist
-    selR <- SG.result$Rlist
-    sel.id <- NULL
-    LR.class <- NULL
-    for (i in 1:length(selG)) {
-      sel.id <- c(sel.id, which(origclass == selG[i]))
-      LR.class <- c(LR.class, rep(ifelse(sum(selL == selG[i]) !=
-                                           0, "L", "R"), length(which(origclass == selG[i]))))
-    }
-    proj.data <- c(as.matrix(origdata) %*% as.matrix(Alpha[TS[node.id,
-                                                              4], ]))[sel.id]
-
-    proj.class <- origclass[sel.id]
-    plot.data <- data.frame(proj.data = proj.data, origclass = proj.class,
-                            cut = cut.off[TS[node.id, 4], Rule], node.id =node.id, LR.class,Dir=as.factor(proj.data>cut.off[TS[node.id, 4], Rule]))
-    colnames(plot.data)[2]<-"Class"
-
-    plot.data
-
-  }
-}
-
-PPtree_dens <- function(ppf, tr, nodes = NULL, pos ="none") {
-
-  if(length(ppf[[8]][[tr]]$Tree.Struct[ppf[[8]][[tr]]$Tree.Struct[,4]!=0,1])<3){
-    ns =length(ppf[[8]][[tr]]$Tree.Struct[ppf[[8]][[tr]]$Tree.Struct[,4]!=0,1])
-  }else{
-    ns=3
-  }
-  if(is.null(nodes)){
-    nodes <- ppf[[8]][[tr]]$Tree.Struct[ppf[[8]][[tr]]$Tree.Struct[,4]!=0,1][1:ns]
-  }
-  nn <- data.frame(nn = nodes)
-  densf <- function(x) {
-    dat_dens(PPclassOBJ = ppf[["output.trees"]][[tr]],
-             node.id = x,
-             Rule = 1)
-
-  }
-
-dat_pl <- lapply(as.list(nodes), densf) %>% bind_rows(.id='node')
-
-  myColors <- brewer.pal(dim(unique(ppf$train[ppf$class.var]))[1], "Dark2")
-  names(myColors) <- levels(ppf$train[ppf$class.var][, 1])
-  dat_pl$Class <- as.factor(dat_pl$Class)
-
-  if(is.factor(ppf$train[ppf$class.var][, 1])){
-    levels(dat_pl$Class) <-  levels(ppf$train[ppf$class.var][, 1])
-
-  }else{
-    levels(dat_pl$Class) <-  levels(as.factor(ppf$train[ppf$class.var][, 1]))
-  }
-
-  p1 <- dat_pl %>% filter(node.id %in%nodes) %>% mutate(node.id = paste("Node",node.id)) %>% 
-    ggplot( aes(  x = proj.data, group = Class, fill = Class ) ) +
-    geom_density(alpha = .5) + facet_grid(~ node.id, scales = 'free') +
-    scale_fill_manual(values = myColors, labels=c("BF", "BM", "OF", "OM"), name="") +geom_vline(aes(xintercept = cut),
-                                                      linetype = "dashed",
-                                                      color = 2) + labs(x="") +
-    scale_x_continuous( breaks = c(-0.5,0,0.5),
-                    limits=c(-0.5,0.5)) +theme(axis.text.x = element_text(size=8))
-
-  p1 + theme(legend.position = pos,aspect.ratio = 1) +
-    guides(col=guide_legend(nrow = 2))
-}
-
-p1b <- PPtree_dens(ppf, tree, nodes = c(1,2,3), pos="none")
-
-#Mosaic plot Tab 2
-PPtree_mosaic <- function(ppf,tr, nodes = NULL, pos="none"){
-  if(length(ppf[[8]][[tr]]$Tree.Struct[ppf[[8]][[tr]]$Tree.Struct[,4]!=0,1])<3){
-    ns =length(ppf[[8]][[tr]]$Tree.Struct[ppf[[8]][[tr]]$Tree.Struct[,4]!=0,1])
-  }else{
-    ns=3
-  }
-
-  if(is.null(nodes)){
-    nodes <- ppf[[8]][[tr]]$Tree.Struct[ppf[[8]][[tr]]$Tree.Struct[,4]!=0,1][1:ns]
-  }
-  nn <- data.frame(nn = nodes)
-
-  densf <- function(x){
-    dat_dens(PPclassOBJ=ppf[[8]][[tr]],node.id=x,Rule=1)
-  }
-
-dat_pl <- lapply(as.list(nodes), densf) %>% bind_rows(.id='node')
-  levels(dat_pl$Dir)<-c("Left", "Right")
-  myColors <- brewer.pal(length( unique( ppf[[8]][[tr]]$origclass ) ), "Dark2")
-  names(myColors) <- levels(dat_pl$Class)
-
-  dat_mosaic <- data.frame( with(dat_pl, table(Class, Dir,node.id) ) )
-
-  p1 <- dat_mosaic %>% filter(node.id %in%nodes) %>% mutate(node.id = paste("Node",node.id))%>% ggplot() +
-    geom_mosaic( aes(weight = Freq, x = product(Class,Dir) ,fill = Class),  offset=.005)+facet_grid(~node.id)+
-    scale_fill_manual(values = myColors) + theme(legend.position=pos,axis.text.x  = element_text(angle=90, vjust=0.5),aspect.ratio = 1) +xlab("")
-
- p1
-
-}
-
-p1c <- PPtree_mosaic(ppf, tree, nodes = c(1,2,3))
-
-nodes <- nrow(ppf[["output.trees"]][[tree]]$projbest.node)
-projbest.node <- ppf[["output.trees"]][[tree]]$projbest.node
-coefs <- data.frame(node = as.factor(1:nodes) , projbest.node)
-colnames(coefs)  <- c("node", colnames(ppf$train[, -1]))
-coefs <- coefs %>% gather(variable, value,-node) %>% mutate(node = paste("Node",node))
-
-p1d <-
-  coefs %>%
-  ggplot(aes(x = variable, y = value)) + geom_segment(aes(yend = 0, xend = variable), size = 3) +
-  facet_grid(node ~ .) + geom_hline(yintercept = 0) + geom_hline(
-    yintercept = c(-1, 1) * 1 / sqrt(ncol(ppf$train[, -1])), col = 2,
-    linetype = "dashed") + labs(y = "Coefficients")
-
-g_legend <- function(a.gplot){
-  tmp <- ggplot_gtable(ggplot_build(a.gplot))
-  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
-  legend <- tmp$grobs[[leg]]
-  return(legend)
-}
-
-layoutmat <- cbind(c(1, 1, 1), c(2, 3, 4))
-
-labpl <-PPtree_dens(ppf, tree, nodes =c(1,2,3), pos="bottom")
-ppf[["output.trees"]][[tree]]$origclass <-ppf[["output.trees"]][[tree]]$origclass %>% recode(`1` ="BF", `2`="BM", `3`="OF",`4`="OM" )
-
-plot(ppf[["output.trees"]][[tree]], main = "", font.size = 12)
-         grid.arrange(p1d , p1b, p1c,  g_legend(labpl),layout_matrix=layoutmat  ,
-       widths =c(2,2) ,heights=c(2,2,.5), as.table =TRUE)
-
-@
+\includegraphics[width=\maxwidth]{trcomb-2} 
+\end{knitrout}
 \caption{Visualizing the PPtree model of the crab data.
@@
 \centering
-<<treenode, echo=FALSE, warning=FALSE, message=FALSE,fig.height=7,fig.width=7.5>>=
-
-# Jittered side by side plot with the absolute importance variable for a each tree in each node partition. The red dot in each
-#facet identified a specific tree in the forest. In crab data we have two points for each node partition because in each node
-# we have selected to variables at random.
-
-impotree <- function(data,ppf, nodes = c(1, 2, 3) ){
-
-bnf <- function(x) {
-  bn <- abs(x$projbest.node)
-  bn[bn == 0] <- NA
-  data.frame(node = 1:dim(x$projbest.node)[1], bn)
-}
-
-bestnode <- ppf[["output.trees"]] %>%  lapply(bnf) %>% bind_rows()
-
-colnames(bestnode)[-1] <- colnames(data[, -1])
-bestnode$node <- as.factor(bestnode$node)
-
-aux <- bestnode %>% filter(node %in% nodes) %>% mutate(ids = rep(1:ppf$n.tree, each = length(nodes))) %>% mutate(node = paste("Node",node))%>% gather(var, value, -ids, -node)
-
-aux$Variables <- as.numeric(as.factor(aux$var) )
-aux$Abs.importance <- round( aux$value, 2)
-
-p <- ggplot(filter(aux, !ids %in% tree),
-            aes(x = Variables , y = Abs.importance , group = ids)) + geom_jitter(height = 0, size = I(3),
-                                                                                 alpha = 0.3) +
-  facet_grid(node ~ .) + scale_x_discrete(limits = levels(as.factor(aux$var))) +
-  theme(legend.position = "none")
-
-   p + geom_jitter(
-  data = filter(aux, ids %in% tree),
-   aes(
-    x = Variables ,
-    y = Abs.importance ,
-    group = ids,
-    colour = "red"
-  ),
-  height = 0,
-  size = I(3)
-) + facet_grid(node ~ .)  +
-  theme(legend.position = "none") + labs(y = "Absolute coefficient")
-}
-
-   pimp <- impotree(crab, ppf )
-
-# This plot show a scatter plot for each of the two variables used in the node partition. The line represents the boundary defined
-# in each node partition.
-
-# function to select the 2 variables used in each node partition in each tree
-dat_2d <- function(node) {
-  indtree <- ppf[["output.trees"]][[tree]]
-  sel <- (1:(ncol(ppf$train) - 1))[indtree$projbest.node[node, ] !=0] #variables used in the specific node partition
-  aux <- ppf$train[ppf$oob.obs[node, ] == 0, c("Type", colnames(ppf$train[, -1])[sel[1]], colnames(ppf$train[, -1])[sel[2]])]
-  aux
-}
-
-myColors <- brewer.pal(length(unique(ppf$train[, ppf$class.var])), "Dark2")
-
-#This function  defines the intercept and the slope of the decision boundary for the scatterplot with the two selected variables, for each node partition.
-# Since in each node there is a two step projection one to transform the problem in a two clkass problm
-# and the second to separate the two new classes.
-#The decision are made if A'X >c asign to right node else to left.
-#  a1x1 + a2x2<c (c based on rule 1) we are using Rule 1 then we define the splope as - cutoff is is based on rule 1 -
-# a1x1 < c -a2x2
-#x1=(c -a2x2)/a1
-# c/a1 is the intercept
-# and a2/a1 is the slope
-
-interslope <- function(node){
-  indtree <- ppf[["output.trees"]][[tree]]
-  ppf$train[,-1]
-  sel <- (1:(ncol(ppf$train) - 1))[indtree$projbest.node[node, ] !=0] # select the oob observations for selected tree
-  intercept <- indtree$splitCutoff.node[node, 1] / indtree$projbest.node[node, sel[1]]
-  slope <- -indtree$projbest.node[node, sel[2]] / indtree$projbest.node[node, sel[1]]
-
-  return(list(intercept,slope))
-  }
-
-d_boundary <- function(node){
-  indtree <- ppf[["output.trees"]][[tree]]
-  sel <- (1:(ncol(ppf$train) - 1))[indtree$projbest.node[node, ] !=0]
-  if(node%in%1) dat = dat_2d(node)
-  if(node%in%2) dat = dat_2d(node) %>%  filter(Type %in% c("BlueMale", "OrangeMale") )
-  if(node%in%3) dat = dat_2d(node) %>%  filter(Type %in% c("BlueFemale", "OrangeFemale") )
-
-   levels(dat$Type) <- c("BF","BM", "OF", "OM")
-
-  p1 <- ggplot(data = dat) + geom_point(ggplot2::aes_string(y = (colnames(dat_2d(node ))[[2]]),
-                                                            x = colnames(dat_2d(node ))[[3]],
-                                                            color =  colnames(dat_2d(node ))[[1]]),
-                                        alpha = .5, size = I(3)) + xlim(c(min(dat[,2],dat[,3]),max(dat[,2],dat[,3]))) + ylim(c(min(dat[,2],dat[,3]),max(dat[,2],dat[,3]) ))+
-    geom_abline(intercept = interslope(node)[[1]], slope = interslope(node)[[2]] ,size = 1 ) +
-    labs(color = "Class") + theme( aspect.ratio = 1) + labs(color = "Class")
-
-  if(node%in%1){
-    p1 <- p1 +  scale_colour_brewer(name = "Class", type = "qual", palette = "Dark2")
-  }
-  if(node%in%2){
-    p1 <-   p1 +  scale_color_manual(values = myColors[c(2,4)])
-  }
-  if(node%in%3){
-    p1<-  p1 +  scale_color_manual(values = myColors[c(1,3)])
-  }
-  p1
-}
-
-p1 <-  d_boundary(1)
-p2 <-  d_boundary(2)
-p3 <-  d_boundary(3)
-
-# combined plot with importance variable for each tree and node partition and SP plot with boundary in each node
-
-  grid.arrange(pimp, arrangeGrob(p1, p2, p3, ncol=1), ncol=2)
-
-@
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+\includegraphics[width=\maxwidth]{treenode-1} 
+\end{knitrout}
 \caption{Visualizing the variable importance of all the trees in the forest, for three nodes.
@@
 \centering
-<<mds, echo=FALSE, warning=FALSE, message=FALSE, fig.height=4, fig.width=7.5>>=
-  #Multidimensional scaling plot using proximity matrix information from PPforest for CRAB
-mdspl2d <- function(ppf, lege = "bottom", siz = 3, k = 2) {
-  #n.class <- length(levels(ppf[[1]])) #ppf$train %>% select_(ppf$class.var) %>% unique() %>% nrow()
-  #lev <- levels(ppf[[1]]) #ppf$train %>% select_(ppf$class.var) %>% sapply(levels) %>% as.factor()
-
-  d <- diag(nrow(ppf$train))
-  d <- as.dist(d + 1 - ppf$proximity)
-  rf.mds <- stats::cmdscale(d, eig = TRUE,  k = k)
-  colnames(rf.mds$points) <- paste("MDS", 1:k, sep = "")
-  #nlevs <- nlevels(ppf$train[, 1])
-
-  df <- data.frame(Class = ppf$train[, 1], rf.mds$points)
-  #if (lege) {
-  mds <- ggplot2::ggplot(data = df) +
-   geom_point(ggplot2::aes(x = MDS1, y = MDS2, color = Class),
-              size = I(siz), alpha = .5) +
-   scale_colour_brewer(type = "qual", palette = "Dark2", name = "Class") +
-   theme(legend.position = lege, aspect.ratio = 1)
-  #}
-  #else  {
-  #  mds <- ggplot2::ggplot(data = df) +
-  #    geom_point(ggplot2::aes(x = MDS1, y = MDS2, color = Class),
-  #               size = I(siz), alpha = .5) +
-  #    scale_colour_brewer(type = "qual", palette = "Dark2", name = "Class") +
-  #    theme(legend.position = "none", aspect.ratio = 1)
-  #}
-  mds
-}
-
-grid_arrange_shared_legend <- function(..., ncol = length(list(...)), nrow = 1, position = c("bottom", "right")) {
-
-  plots <- list(...)
-  position <- match.arg(position)
-  g <- ggplotGrob(plots[[1]] + theme(legend.position = position))$grobs
-  legend <- g[[which(sapply(g, function(x) x$name) == "guide-box")]]
-  lheight <- sum(legend$height)
-  lwidth <- sum(legend$width)
-  gl <- lapply(plots, function(x) x + theme(legend.position="none"))
-  gl <- c(gl, ncol = ncol, nrow = nrow)
-
-  combined <- switch(position,
-       "bottom" = arrangeGrob(do.call(arrangeGrob, gl),
-                   legend,
-                   ncol = 1,
-                   heights = unit.c(unit(1, "npc") - lheight, lheight)),
-       "right" = arrangeGrob(do.call(arrangeGrob, gl),
-                   legend,
-                   ncol = 2,
-                   widths = unit.c(unit(1, "npc") - lwidth, lwidth)))
-  #grid.newpage()
-  grid.draw(combined)
-}
-
-k=3
-lege = "none"
-siz = 3
-d <- diag(nrow(ppf$train))
-d <- as.dist(d + 1 - ppf$proximity)
-rf.mds <- stats::cmdscale(d, eig = TRUE,  k = k)
-colnames(rf.mds$points) <- paste("MDS", 1:k, sep = "")
-df <- data.frame(Class = ppf$train[, 1], rf.mds$points)
-levels(df$Class) <- c("BF", "BM", "OF", "OM")
-p12 <-  ggplot2::ggplot(data = df) +
-  geom_point(ggplot2::aes(x = MDS1, y = MDS2, color = Class),
-             size = I(siz), alpha = .5) +
-  scale_colour_brewer(type = "qual", palette = "Dark2", name = "Class") +
-  theme(legend.position = lege, aspect.ratio = 1)
-p13 <-  ggplot2::ggplot(data = df) +
-  geom_point(ggplot2::aes(x = MDS1, y = MDS3, color = Class),
-             size = I(siz), alpha = .5) +
-  scale_colour_brewer(type = "qual", palette = "Dark2", name = "Class") +
-  theme(legend.position = lege, aspect.ratio = 1)
-p23 <-  ggplot2::ggplot(data = df) +
-  geom_point(ggplot2::aes(x = MDS2, y = MDS3, color = Class),
-             size = I(siz), alpha = .5) +
-  scale_colour_brewer(type = "qual", palette = "Dark2", name = "Class") +
-  theme(legend.position = lege, aspect.ratio = 1)
-
-grid_arrange_shared_legend(p12, p13, p23, ncol=3, nrow=1)
-
-  @
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+\includegraphics[width=\maxwidth]{mds-1} 
+\end{knitrout}
 \caption{Examining similarity between cases, using pairwise plots of multidimensional scaling into a three-dimensional space.
@@
 \centering
-<<ternary, echo=FALSE, warning=FALSE, message=FALSE, fig.height=4, fig.width=7.5, out.extra='trim=0cm 2cm 0cm 2cm, clip'>>=
-f.helmert <- function(d)
-{
-  helmert <- rep(1 / sqrt(d), d)
-  for (i in 1:(d - 1))
-  {
-    x <- rep(1 / sqrt(i * (i + 1)), i)
-    x <- c(x, -i / sqrt(i * (i + 1)))
-    x <- c(x, rep(0, d - i - 1))
-    helmert <- rbind(helmert, x)
-  }
-
-  return(helmert)
-}
-
-makePairs <- function(dat, id = c(a, b, c)) {
-  aux <- dat[,-c(1, 2)]
-
-  d <- aux[, id]
-
-  grid <- expand.grid(x = id, y = id)
-  grid <- subset(grid, x != y)
-  all <- do.call("rbind", lapply(1:nrow(grid), function(i) {
-    xcol <- grid[i, "x"]
-    ycol <- grid[i, "y"]
-    data.frame(
-      Class = dat[, 1],
-      ids = dat[, 2],
-      x = dat[, xcol+2],
-      y = dat[, ycol+2],
-      pair = paste(grid[i, ], collapse = '-')
-    )
-  }))
-
-  all
-}
-
-#ppf PPforest object
-#V1,v2,v3 select the 3 proj directions
-ternarydata <- function(ppf, v1, v2, v3){
-  n.class <- ppf$train %>% select_(ppf$class.var) %>% unique() %>% nrow()
-  projct <- t(f.helmert(length(unique(ppf$train[, ppf$class.var])))[-1,])
-
-  dat3 <-
-    data.frame(
-      Class = ppf$train[, ppf$class.var],
-      ids = 1:nrow(ppf$train),
-      proj.vote = as.matrix(ppf$votes) %*% projct
-    )
-
-  ##with 3 or less classes
-  empt <- rep(1:nrow(dat3), 3)
-  #dat3.empt <- dat3[empt, ] %>% mutate(rep = rep(1:3, each = nrow(dat3)))
-  if (n.class > 3) {
-    gg1 <-  makePairs(dat3, c(v1,  v2, v3))
-  }
-
-  gg1 <-  makePairs(dat3, id = c(v1, v2, v3))
-
-  return(gg1)
-}
-
-#need to run f_hermite
-f_composition <- function(data) {
-  d <- dim(data)[2]
-  hm <- f.helmert(d)
-  x <- data - matrix(1 / d, dim(data)[1], d)
-  return((x %*% t(hm))[,-1])
-}
-
-simplex <- function(p = 3) {
-  vert <- f_composition(diag(p + 1))
-  colnames(vert) <- paste0("d", 1:ncol(vert))
-
-  wires <-
-    do.call(expand.grid, list(c(1:nrow(vert)), c(1:nrow(vert))))
-
-  structure(list(points = vert,
-                 edges = wires[!(wires[, 1] == wires[, 2]),]))
-}
-
-##ternary plot
-ternaryshell <- function(ppf, sp = 3, dx = 1, dy = 2, v1=1, v2=2, v3=3){
-  s <- simplex(sp)
-  pts <- data.frame(s$points)
-  #pts <- data.frame(s$points[,c(dx,dy)])
-  gg1 <- ternarydata(ppf, v1, v2, v3)
-
-  edg <- data.frame(x1=pts[,dx][s$edges[,1]], x2=pts[,dx][s$edg[,2]],
-                    y1=pts[,dy][s$edg[,1]], y2=pts[,dy][s$edg[,2]])
-
-  p1  <- gg1 %>% filter(pair %in% paste(dx,dy, sep="-") ) %>%
-    select(Class, x, y) %>%
-    ggplot(aes(x, y, color = Class)) +
-      geom_segment(data = edg, aes(x = x1, xend = x2,
-                                 y = y1, yend = y2), color = "black" ) +
-      geom_point(size = I(3), alpha = .5) +
-      labs(y = "",  x = "") +
-      theme(legend.position = "none", aspect.ratio = 1) +
-      scale_colour_brewer(type = "qual", palette = "Dark2") +
-      labs(x = paste0("T",dx,""), y = paste0("T",dy,"")) +
-    theme(aspect.ratio=1)
-
-  p1
-}
-t1 <- ternaryshell(ppf, 3, 1, 2)
-t2 <- ternaryshell(ppf, 3, 1, 3)
-t3 <- ternaryshell(ppf, 3, 2, 3)
-
-grid.arrange(t1, t2, t3, ncol = 3)
-
-#vimeoternary fish code
-# library(colorspace)
-# library(RColorBrewer)
-# library(tourr)
-# projct <- t(f.helmert(7)[-1,] )
-# 
-# dat7 <- data.frame(class=ppf[[1]], ppf$votes%*%projct)
-# colnames(dat7)[-1] <- paste("x", 1:7,sep="")
-# pal <- brewer.pal(7,"Dark2") 
-# col <- pal[as.numeric(dat7[, 1])]
-# 
-# edges <- rbind(c(1,0,0,0,0,0,0), c(0,1,0,0,0,0,0), c(0,0,1,0,0,0,0), c(0,0,0,1,0,0,0), 
-#                c(0,0,0,0,1,0,0), c(0,0,0,0,0,1,0),c(0,0,0,0,0,0,1) )%*%projct
-# colnames(edges) <- colnames(dat7)[-1]
-# dd7 <- rbind(dat7, data.frame(class=levels(dat7$class),edges ))
-# 
-# x11()
-# animate_xy(dd7[,2:7], col = col, pch=19,axes ="off", edges=as.matrix(expand.grid(160:166,160:166) )[c(2:7,10:14,18:21, 26:28,34:35,42),])
-#
-
-@
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+\includegraphics[width=\maxwidth,trim=0cm 2cm 0cm 2cm, clip]{ternary-1} 
+\end{knitrout}
 \caption{Generalized ternary plot ((G-1) dimensional simplex, in this case a tetrahedron) representation of the vote matrix for four classes.
@@
 \centering
-<<side, echo=FALSE, warning=FALSE, message=FALSE, fig.height=5,fig.width=5>>=
-
-#side-by siede boxplot
-side <-  function(ppf, ang = 0, lege = "bottom", siz = 2.5,
-                  ttl = "Side by side dotplot") {
-  voteinf <- data.frame(ids = 1:length(ppf$train[, 1]), Type = ppf$train[, 1],
-                      ppf$votes, pred = ppf$prediction.oob ) %>%
-  gather(Class, Probability, -pred, -ids, -Type)
-
-  levels(voteinf$Type) <- c("BF", "BM", "OF", "OM")
-  levels(voteinf$pred) <- c("BF", "BM", "OF", "OM")
-  voteinf$Class <- as.factor(voteinf$Class)
-  levels(voteinf$Class) <- c("BF", "BM", "OF", "OM")
-
-  ggplot(data = voteinf, aes(Class, Probability, color = Type)) +
-    geom_jitter(height = 0, size = I(siz), alpha = .5) +
-    ggtitle(ttl) +
-    scale_colour_brewer(type = "qual", palette = "Dark2") +
-    theme(legend.position = lege, legend.text=element_text(angle = ang, size =8),
-legend.title = element_text(size=8)) +
-    labs(colour = "Class", y = "Proportion")
-}
-
-side(ppf, ttl="")
-
-@
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+\includegraphics[width=\maxwidth]{side-1} 
+\end{knitrout}
 \caption{Another representation of the vote matrix, as a jittered side-by-side dotplot.
@@
 \centering
-\includegraphics[width=1\linewidth]{figure/fishT1.pdf}
+\includegraphics[width=1\linewidth]{fishT1.pdf}
 \caption{Schematic diagram illustrating the interactivity in and between plots for individual level exploration panel of the web app.
@@
 \centering
-\includegraphics[width=1\linewidth]{figure/fish11.png}
+\includegraphics[width=1\linewidth]{fish11.png}
 \caption{Entry page for the web app, focusing on examining cases, in terms of similarity and uncertainty.
@@
 \centering
-\includegraphics[width=1\linewidth]{figure/tab1com2.png}
+\includegraphics[width=1\linewidth]{tab1com2.png}
 \caption{Entry page for the web app, focusing on examining cases, for selected points.  \label{tab1comp}}
@@
 \centering
-\includegraphics[width=1\linewidth]{figure/tab2comp.png}
+\includegraphics[width=1\linewidth]{tab2comp.png}
 \caption{The individual model tab in the web app.
@@
 \centering
-\includegraphics[width=1\linewidth]{figure/fishT2.pdf}
+\includegraphics[width=1\linewidth]{fishT2.pdf}
 \caption{Schematic diagram illustrating the interactivity in and between plots for model level exploration panel of the web app.
@@
 \begin{figure}[hbpt]
-\includegraphics[width=.8\linewidth]{figure/fish31.png}
+\includegraphics[width=.8\linewidth]{fish31.png}
 \caption{Performance comparison tab of the web app.

### 2022-10-11 COST_submission_Oct22/PPforestViz_journal.Rnw
update

--- 2022-10-04 PPforestVizupdate.Rnw
@@
 
-There has been some, but not a abundant, research on visualizing classification models.
+There has been some, but not abundant, research on visualizing classification models.
 \cite{urbanek2002exploring} presents interactive tree visualization implemented in the java software called \verb# KLIMT# that includes zooming, selection, multiple views, interactive pruning, and tree construction as well as the interactive analysis of forests of trees using treemaps.
@@
 \end{figure}
-Two data set are used in this paper, crab~\citep{CM74} and fishcatch data~\citep{fishcatch}.
+Two data sets are used in this paper, crab~\citep{CM74} and fishcatch data~\citep{fishcatch}.
 To illustrate the mapping of diagnostics to visualization crab data are used because of its simplicity.
@@
 
-\item {\em \textbf{Uncertainty measure for each observation:}} Across individual (classification) models we can compute the proportion of times that a case is predicted to be each class.
+\item {\em \textbf{Uncertainty measure for each observation:}} Across individual (classification) models we can compute the proportion of times that a case is predicted to be in each class.
 If a case is always predicted to be the true class, there is no uncertainty about an observation.
@@
 \centering
-<<trcomb, echo=FALSE, warning=FALSE, message=FALSE, fig.height=4, fig.width=6.5, results='hide', error=FALSE>>=
+<<trcomb, echo=FALSE, warning=FALSE, message=FALSE, fig.height=4, fig.width=7, results='hide', error=FALSE>>=
 
@@
 Node 1 separates the sexes.
-Node 2 separate males from species and node 3 separate females from species.
+Node 2 separates males from species and node 3 separates females from species.
 Mosaic plots of the confusion table for each split (bottom).
 Node 1 shows the clear split of sexes, with a small number of misclassifications.
-Node 2 where BM are separated from OM with small number of misclassifications.
+Node 2 where BM are separated from OM with a small number of misclassifications.
 Node 3 where BF are separated from OF, almost perfect classification.\label{pptreeviz1}}
@@
 Node 3 uses CW and FL with an even contribution by the two variables.
+Alternative visualizations for the importance variable plot can be considered, for example violin plots or boxplots however, the side-by-side jittered dot plot was preferred since presents individual details on importance measure distribution.
+
 For each tree in the forest, different decision rules are defined; the resulting boundaries on the previous plots are based on Rule 1 $= \frac{m_1}{2} + \frac{m_2}{2} $, where $m_1$ and $m_2$ are the mean of the left and right groups at each node.
@@
 \centering
-<<treenode, echo=FALSE, warning=FALSE, message=FALSE,fig.height=7,fig.width=7.5>>=
+<<treenode, echo=FALSE, warning=FALSE, message=FALSE,fig.height=12,fig.width=12>>=
 
@@
 p <- ggplot(filter(aux, !ids %in% tree),
-            aes(x = Variables , y = Abs.importance , group = ids)) + geom_jitter(height = 0, size = I(3),
-                                                                                 alpha = 0.3) +
+            aes(x = Variables , y = Abs.importance , group = ids)) + geom_jitter(height = 0, size = I(4),
+                                                                                 alpha = 0.2) +
   facet_grid(node ~ .) + scale_x_discrete(limits = levels(as.factor(aux$var))) +
@@
   height = 0,
-  size = I(3)
+  size = I(4)
 ) + facet_grid(node ~ .)  +
@@
                                                             color =  colnames(dat_2d(node ))[[1]]),
-                                        alpha = .5, size = I(3)) + xlim(c(min(dat[,2],dat[,3]),max(dat[,2],dat[,3]))) + ylim(c(min(dat[,2],dat[,3]),max(dat[,2],dat[,3]) ))+
+                                        alpha = .5, size = I(4)) + xlim(c(min(dat[,2],dat[,3]),max(dat[,2],dat[,3]))) + ylim(c(min(dat[,2],dat[,3]),max(dat[,2],dat[,3]) ))+
     geom_abline(intercept = interslope(node)[[1]], slope = interslope(node)[[2]] ,size = 1 ) +
@@
 As \cite{sievert2020interactive} describes, one of the biggest difficulties when it comes to the app is managing linking between plots (also called a widget in the plotly documentation) requiring careful data structure management.
-Each plot has it own data structure and interaction.
+Each plot has its own data structure and interaction.
 Putting them into the structure of a shiny app facilitates access to the widget data and coordinates selections across multiple plots.
@@
 Figure \ref{tab1} shows the arrangement of plots in this first tab (the case level tab).
-
 This selection of plots enables aspects of the model, relating to performance for individual cases, to be examined in the data space.
@@
 
-The selected tree in Figure \ref{tab2} has very small OOB error shown as red point in the boxplot.
+The selected tree in Figure \ref{tab2} has a very small OOB error shown as a red point in the boxplot.
 Two variables were use in each node partition, the first four nodes were selected and shown in the importance variable plot, density plot and mosaic plot.
@@
 A similar analysis can be done for other nodes.
-This tab allow us to understand individual trees in the forest and analyzed which variables are important for the node partition related with the OOB error involved in that specific tree.
+This tab allows us to understand individual trees in the forest and analyze which variables are important for the node partition related to the OOB error involved in that specific tree.
 
@@
 Variable importance is displayed as jittered dot plots for selected nodes of all trees.
-The boxplot is linked to a display of the PPTree, the variable importance plot, and a density plot of the data projections showing splits at each selected nodes and confusion tables as mosaic plots.
+The boxplot is linked to a display of the PPTree, the variable importance plot, and a density plot of the data projections showing splits at each selected node and confusion tables as mosaic plots.
 Clicking a point in the OOB error boxplot triggers various updates: each of the importance values for the same tree are highlighted (red), the tree that this corresponds to is drawn, the error for the tree is shown on the boxplot (in red), and the density plot and the mosaic plot are also updated.
@@
 Having better tools to open up black box models will provide a better understanding of the data, the model strengths and weaknesses, and how the model will perform on future data.
-This visualisation app provides a selection of interactive plots to diagnose PPF models.
+This visualization app provides a selection of interactive plots to diagnose PPF models.
 This shell could be used to make an app for other ensemble classifiers.
@@
 Performance of the app is good for relatively small data sets.
-If the number of observation is large, overploting will be an issue in some of the visualizations.
+If the number of observations is large, overploting will be an issue in some of the visualizations.
 If the number of variables is large, the parallel coordinate plot and plot of variable importance would benefit by the addition of a user input to select variables.

### 2022-10-18 COST_submission_18_Oct22/PPforestVizupdate.Rnw
update repo

--- 2022-10-11 COST_submission_Oct22/PPforestViz_journal.Rnw
@@
 
-\institute{Natalia da Silva \at
-              Instituto de Estadística (IESTA), Universidad de la República\\
-              \email{natalia.dasilva@fcea.edu.uy}
+\institute{Corresponding author: Natalia da Silva \at
+              Instituto de Estadística (IESTA), Universidad de la República, Montevideo, Uruguay.\\
+              \email{natalia.dasilva@fcea.edu.uy}\\
+              ORCID:0000-0002-6031-7451\\
            \and
           Dianne Cook \at 
-          Department of Econometrics and Business Statistics, Monash University\\
-          \email{dicook@monash.edu}\and
+          Department of Econometrics and Business Statistics, Monash University, Melbourne, Australia.\\
+          \email{dicook@monash.edu}\\
+          ORCID:0000-0002-3813-7155\\
+          \and
           Eun-Kyung Lee \at
-          Department of Statistics, Ewha Womans University\\
+          Department of Statistics, Ewha Womans University, Seoul, South Korea.\\
           \email{lee.eunk@gmail.com}

### 2022-10-18 COST_submission_18_Oct22/PPforestVizupdate.tex
update repo

--- 2022-10-11 COST_submission_July22/PPforestViz_journal.tex
@@
 \definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth,trim=0cm 2cm 0cm 2cm, clip]{ternary-1} 
+\includegraphics[width=\maxwidth]{ternary-1} 
 \end{knitrout}
@@
 \centering
-\includegraphics[width=1\linewidth]{fish11.png}
+\includegraphics[width=1.05\linewidth]{fish11.png}
 \caption{Entry page for the web app, focusing on examining cases, in terms of similarity and uncertainty.
@@
 \centering
-\includegraphics[width=1\linewidth]{tab1com2.png}
+\includegraphics[width=1.1\linewidth]{tab1com2.png}
 \caption{Entry page for the web app, focusing on examining cases, for selected points.  \label{tab1comp}}
@@
 \centering
-\includegraphics[width=1\linewidth]{tab2comp.png}
+\includegraphics[width=1.1\linewidth]{tab2comp.png}
 \caption{The individual model tab in the web app.
@@
 \centering
-\includegraphics[width=1\linewidth]{fishT2.pdf}
+\includegraphics[width=1.05\linewidth]{fishT2.pdf}
 \caption{Schematic diagram illustrating the interactivity in and between plots for model level exploration panel of the web app.
@@
 Some of the proposed reduced the multi-class problem to a set of binary problems.
-The approach used for a multi-class ROC analysis in this article is called one-against-all ~\citep{allwein2000reducing}.
+The approach used for a multi-class ROC analysis in this article is called one-against-all \citep{allwein2000reducing}.
 
 \begin{figure}[hbpt]
-\includegraphics[width=.8\linewidth]{fish31.png}
+\includegraphics[width=1\linewidth]{fish31.png}
 \caption{Performance comparison tab of the web app.

### 2022-10-18 COST_submission_9_Oct22/PPforestViz_journal.tex
update

--- 2022-10-18 COST_submission_18_Oct22/PPforestVizupdate.tex
@@
 
-\institute{Corresponding author: Natalia da Silva \at
-              Instituto de Estadística (IESTA), Universidad de la República, Montevideo, Uruguay.\\
-              \email{natalia.dasilva@fcea.edu.uy}\\
-              ORCID:0000-0002-6031-7451\\
+\institute{Natalia da Silva \at
+              Instituto de Estadística (IESTA), Universidad de la República\\
+              \email{natalia.dasilva@fcea.edu.uy}
            \and
           Dianne Cook \at 
-          Department of Econometrics and Business Statistics, Monash University, Melbourne, Australia.\\
-          \email{dicook@monash.edu}\\
-          ORCID:0000-0002-3813-7155\\
-          \and
+          Department of Econometrics and Business Statistics, Monash University\\
+          \email{dicook@monash.edu}\and
           Eun-Kyung Lee \at
-          Department of Statistics, Ewha Womans University, Seoul, South Korea.\\
+          Department of Statistics, Ewha Womans University\\
           \email{lee.eunk@gmail.com}
